### 2013-02-19 tidy-data.tex
Resubmit to JSS

\documentclass[article]{jss}

% jss.cls ----------------------------------------------------------------------

%% almost as usual
\author{Hadley Wickham\\RStudio}
\title{Tidy data}

\Plainauthor{Hadley Wickham}
\Plaintitle{Tidy data}

%% an abstract and keywords
\Abstract{
A huge amount of effort is spent cleaning data to get it ready for data analysis, but there has been little research on how to make data cleaning as easy and effective as possible. This paper tackles a small, but important, subset of data cleaning: data ``tidying''. Tidy datasets are easy to manipulate, model and visualise, and have a specific structure: variables are stored in columns, observations in rows, and a single type of experimental unit per dataset. This framework makes it easy to tidy messy datasets because only a small set of tools are needed to deal with a wide range of un-tidy datasets. This structure also makes it easier to develop tidy tools for data analysis, tools that both take in and put out tidy datasets. The advantages of a consistent data structure and matching tools are demonstrated with a case study free from mundane data manipulation chores.
}
\Keywords{data cleaning, data tidying, relational databases, \proglang{R}}
\Plainkeywords{data cleaning, data tidying, relational databases, R}

%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

\Address{
  Hadley Wickham\\
  Chief Scientist, RStudio\\
  Adjunct Assitant Professor, Rice University\\
  E-mail: \email{h.wickham@gmail.com}\\
  URL: \url{http://had.co.nz}
}

% jss.cls ----------------------------------------------------------------------

\DeclareGraphicsExtensions{.png,.pdf,.jpg}
\usepackage[format=plain,font=small]{caption}
\usepackage{booktabs}
\usepackage{verbatim}
\usepackage{epigraph}
\usepackage[subrefformat=parens]{subfig}

\begin{document}
\maketitle

\section{Introduction}

It's often said that 80\% of the effort in a data analysis is spent on data cleaning \citep{dasu:2003}, the process of getting the data ready to analyse. Data cleaning is not only a vital first step, but it is often repeated multiple times over the course of an analysis as new problems come to light. Despite the amount of time it takes, there has been little research on how to clean data well. Part of the challenge is the breadth of activities that cleaning encompasses, from outlier checking, to date parsing, to missing value imputation. To get a handle on the problem, this paper focusses on a small, but important, subset of data cleaning that I call data ``tidying'': structuring data values to make the process of data analysis as easy as possible.

The principles of tidy data provide a standard way of organising the data values within a dataset. This makes initial data cleaning easier because you have a \emph{standard} that you can start with: you don't need to reinvent the wheel for each new dataset. It also makes it easier to develop data analysis tools that work together seamlessly. Current tools are often inconsistent, so you need to spend time munging the output from one tool into the input for another. Tidy datasets and tidy tools work hand in hand to make data analysis easier, allowing you to focus on the interesting domain problem, not the uninteresting details of data structure.

The principles of tidy data are closely tied to relational databases and Codd's relational algebra \citep{codd:1990}, but are framed in a language more familiar to statisticians. Computer scientists have also contributed much to the study of data cleaning. For example, \citet{lakshmanan:1996} defines a extension to SQL allowing it to operate on messy datasets, \citet{raman:2001} provides a framework for cleaning datasets, and \citet{kandel:2011} develops an interactive tool that makes data cleaning easy, automatically creating the code to clean datasets with a friendly user interface. These tools are useful, but are presented in language foreign to most statisticians, fail to give much advice on how datasets should be structured, and lack connections to the tools of data analysis.

The development of tidy data has been driven by my struggles working with real-world datasets, which are often organised in bizarre ways. I have spent countless hours struggling to get these datasets organised in a way that makes data analysis possible, let alone easy. I have also struggled to impart these skills to my students, so they can tackle problems in the real-world, where there are few constraints on how datasets are organised. In the course of these struggles I developed the reshape and reshape2 \citep{wickham:2007b} R packages, but while I could intuitively use the tools, I lacked the framework to make my understanding explicit. This paper provides that framework, and provides a consistent ``philosophy of data'' that underlies my work in the plyr \citep{me:plyr} and ggplot2 \citep{me:ggplot2} packages.

The paper proceeds as follows. Section~\ref{sec:defining} begins by defining the three characteristics that make a dataset tidy. Most real world datasets are not tidy, so Section~\ref{sec:tidying} describes the operations needed to make messy datasets tidy, and illustrates the techniques with a range of real-life examples. Section~\ref{sec:tidy-tools} defines tidy tools, tools that input and output tidy datasets, and discusses how together tidy data and tidy tools make it easier to do a data analysis. These principles are illustrated with a small case study in Section~\ref{sec:case-study}. Section~\ref{sec:discussion} concludes with a discussion of what this framework misses and other approaches that might be fruitful to pursue.

\section{Defining tidy data}
\label{sec:defining}

\epigraph{Happy families are all alike; every unhappy family is unhappy in its own way}{Leo Tolstoy}

Like families, tidy datasets are all alike, but every messy dataset is messy in its own way. Here we will define the essence of tidy data, and provide important vocabulary for describing datasets.

Statistical datasets usually come in the form of a rectangular \textbf{table}, made up of \textbf{rows} and \textbf{columns}. Tables are usually labelled with column names, and sometimes with row names. The table is a collection of data \textbf{value}s, typically either numeric (if quantitative) or character (if qualitative). Values are grouped into \textbf{variables}, measurements of a single quantifiable property. Multiple measurements made on the same \textbf{observational unit} form an \textbf{observation}.

There is a little circularity in the definition, which is resolved by defining an observational unit as the smallest constituent atom of the experimental unit. For example, in a longitudinal medical study, the experimental unit might be the patient, but the observational unit is the patient at each time point. In an agricultural experiment, the experimental units might be the cages given different diets, but the observational units are the pigs whose individual weights are measured. This definition of observational unit leads to variables which are physical properties, like blood pressure and weight, not conditional properties like blood pressure at time 1, or weight of pig 2. Structure in the variable names (`trtA', `trtB', `trtC'; `jan1', `feb1', `mar1') is difficult to exploit, and suggests that the names are values of other variables.

% Experimental units are usually independent, and the statistical challenge in an analysis is capturing the dependence of observational units within an experimental unit. Failing to do so correctly leads to problems like pseudoreplication \citep{hurlbert:1984}.

Table~\ref{tbl:preg-raw} illustrates a data format commonly seen in the wild. Here it is used to store information about an imaginary experiment with two treatments. There are two columns and three rows (plus column and row names), defining three variables (name, treatment, result) and either five or six observations, depending on how you think about the missing value.

\begin{table}[htbp]
  \centering
  \begin{tabular}{lrr}
  \toprule
               & treatmentA & treatmentB \\
  \midrule
  John Smith   &      ---  &            5 \\
  Jane Doe     &        1  &            4 \\
  Mary Johnson &        2  &            3 \\
  \bottomrule
  \end{tabular}
  \caption{Typical presentation dataset.}
  \label{tbl:preg-raw}
\end{table}

It's not always obvious what qualifies as a variable, and a variable in one analysis may be a value in another. For example, if the columns in the example were were height and weight we would have been happy to call them variables. If the columns were height and width, it would be less clear cut, as we might think of height and weight as values of a dimension variable. If the columns were home phone and work phone, we could treat these as two variables, but in a fraud detection environment we might want variables phone number and number type because the use of one phone number for multiple people might suggest fraud. A general rule of thumb is that it is easier to describe functional relationships between variables/columns (e.g.\ z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of rows/observations (e.g.\ average of group a vs.\ average of group b) than between columns. If you find yourself fighting these constraints, you may not have correctly identified the appropriate variables for your problem.

In a given analysis, there may also be multiple \textbf{types} of observational unit. For example, in a trial of new allergy medication we might have three observational types: demographic data collected on \emph{each person} (age, sex, race), on \emph{each person on each day} of hay fever season (number of sneezes, redness of eyes), and on \emph{each day} (temperature, pollen count).

A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types. In {\bf tidy data}:

\begin{enumerate}
  \item Each variable forms a column,
  \item Each observation forms a row,
  \item Each table (or file) stores information about one observational type.
\end{enumerate}

\noindent This is Codd's 3rd normal form \citep{codd:1990}, but with the constraints framed in statistical language, and the focus on a single dataset, not the many connected datasets common in relational databases. I will call other arrangements of data {\bf messy}.

Table~\ref{tbl:preg-tidy} shows the tidy version of Table~\ref{tbl:preg-raw}. Now each row represents an observation, the result of one treatment on one person, and each column is a variable. We could also choose to drop the row representing the result of John Smith's treatment a, but we don't, because missing values should only be omitted if they are structural: impossible because of the design of the experiment.

\begin{table}[htbp]
  \centering
  \begin{tabular}{lrr}
    \toprule
    name &    treatment    & result  \\
    \midrule
    Jane Doe      & a & 1 \\
    Jane Doe      & b & 4 \\
    John Smith    & a & --- \\
    John Smith    & b & 5 \\
    Mary Johnson  & a & 2 \\
    Mary Johnson  & b & 3 \\
    \bottomrule
  \end{tabular}
  \caption{Tidied data.}
  \label{tbl:preg-tidy}
\end{table}

While order of variables and observations does not affect analysis, a good ordering makes it easier to scan the raw values. One way of organising variables is by their role in the analysis: are values fixed by the design of the data collection, or are they measured during the course of the experiment? \textbf{Fixed} variables describe the experimental design and are known in advance. These are often called dimensions by computer scientists, and typically denoted by subscripts on random variables. \textbf{Measured} variables are what we actually measure in the study. Fixed variables should come first, followed by measured variables, each ordered so that related variables are contiguous. Rows can then be ordered by the first variable, breaking ties with the second and subsequent variables. This is the convention adopted by all tabular displays in this paper. When using tables for communication, this so-called ``Alabama first'' \citep{wainer:2000} ordering should be abandoned, and replaced with an ordering based on a meaningful variable.

\section{Tidying messy datasets}
\label{sec:tidying}

Real datasets can violate the three precepts of tidy data in almost every way imaginable, and often do. While occasionally you do get a dataset that you can start analysing immediately, this is the exception, not the norm.  This section describes the the five most common problems that make datasets messy, along with their remedies:

\begin{itemize}
  \item column headers are values, not variable names
  \item multiple variables are stored in one column
  \item variables are stored in both rows and columns
  \item multiple types of observational unit are stored in the same table
  \item a single observational unit is stored in multiple tables
\end{itemize}

Surprisingly, most messy datasets, including types of messiness not explicitly described above, can be tidied with a small set of tools: ``melting'', string splitting, and ``casting''. The following sections illustrate each problem with a real dataset that I have encountered in practice, and show how to tidy it. The complete datasets and the R code used to tidy them are available online at \url{https://github.com/hadley/tidy-data}, and in the online supplementary materials for this paper.

\subsection{Column headers are values, not variable names}

A common type of messy dataset is tabular data designed for presentation, where variables form both the rows and columns, and column headers are values, not variable names. While in this paper I call this arrangement messy, in some cases it can be extremely useful. It provides efficient storage for completely crossed designs, and it can provide extremely efficient computation if desired operations can be expressed as matrix operations. This issue is discussed in more depth in Section~\ref{sec:discussion}.

Table~\ref{tbl:pew-raw} shows a subset of a typical dataset of this form. This dataset explores the relationship between income and religion in the US, and comes from a report\footnote{\url{http://religions.pewforum.org/pdf/comparison-Income\%20Distribution\%20of\%20Religious\%20Traditions.pdf}} produced by the Pew Research Center. The Pew Center is an American think-tank that collects data on attitudes to topics from religion to the internet, and produces many reports that contain datasets in the format shown.

\begin{table}[htbp]
  \centering
  \input{data/pew-raw.tex}
  \caption{The first ten rows of data on income and religion from the Pew Forum. Two columns have been omitted to save space: `\$100-150k' and `\$150k'} 
  \label{tbl:pew-raw}
\end{table}

This dataset has three variables, religion, income and frequency, and to tidy it we need to ``melt'', or stack it, turning columns into rows. This makes ``wide'' datasets ``long'' or ``tall'', but I will avoid those terms because they are imprecise. Melting is parameterised by a list of columns that are already variables, or \emph{colvars} for short. The other columns are converted into two variables: a new ``column'' variable that contains repeated column headings and a ``value'' variable that contains a vector of data values concatenated from the previously separate columns. This is illustrated in Table~\ref{tbl:melt} with a toy dataset. The result of melting is a ``molten'' dataset.

\begin{table}
  \centering
  \subfloat[Raw data]{\label{tbl:melt-raw} \input{data/melt-raw.tex}}%
  \hspace{2em}%
  \subfloat[Molten data]{\label{tbl:melt-molten}\input{data/melt-output.tex}}

  \caption{A simple example of melting. (a) is melted with one colvar, row, yielding the molten dataset (b). The information in each table is exactly the same, just stored in a different way.}
  \label{tbl:melt} 
\end{table}

The Pew dataset has one colvar, religion, and melting yields Table~\ref{tbl:pew-clean}. The column variable has been renamed to {\tt income}, and the value column to {\tt freq}, to better reflect their roles in this dataset. This form is tidy: each column represents a variable and each row represents an observation, in this case a demographic unit corresponding to a combination of religion and income.

\begin{table}[htbp]
  \centering
  \input{data/pew-clean.tex}
  \caption{The first ten rows of the tidied Pew survey dataset on income and religion. The variable ``column'' has been renamed to income, and ``value'' to freq.}
  \label{tbl:pew-clean}
\end{table}

Another common use of this data format is to record regularly spaced observations over time. For example, the billboard dataset shown in Table~\ref{tbl:billboard-raw} records the date a song first entered the billboard top 100. It has variables artist, track, date.entered, rank and week. The rank in each week after it enters the top 100 is recorded in 75 columns, {\tt wk1} to {\tt wk75}. If a song is in the top 100 for less than 75 weeks the remaining columns are filled with missing values.  This form of storage is not tidy, but is useful for data entry because it reduces duplication: otherwise each song in each week would need its own row, and song metadata like title and artist would need to be repeated. This issue will be discussed in more depth in Section~\ref{sub:multiple-types}.

\begin{table}[htbp]
  \centering
  \input{data/billboard-raw.tex}
  \caption{The first eight billboard top hits for 2000. Other columns not shown are {\tt wk4}, {\tt wk5}, ..., {\tt wk75}.}
  \label{tbl:billboard-raw}
\end{table}

This dataset has colvars year, artist, track, time, and date.entered. Melting yields Table~\ref{tbl:billboard-clean}. I have also done a little cleaning as well as tidying: the column variable has been converted to a week variable by extracting the week number, and I have added a date variable that combines the date the song entered the billboard with the week number.

\begin{table}[htbp]
  \centering
  \input{data/billboard-clean.tex}
  \caption{First fifteen rows of the tidied billboard dataset. The {\tt date} column does not appear in the original table, but can be computed from {\tt date.entered} and {\tt week}.}
  \label{tbl:billboard-clean}
\end{table}

\subsection{Multiple variables stored in one column}

After melting, it often happens that the ``column'' variable is a combination of multiple underlying variables. This is illustrated by the tuberculosis (TB) dataset, a sample of which is shown in Table~\ref{tbl:tb-raw}. This dataset comes from the World Health Organisation, and records the counts of confirmed tuberculosis cases by country, year, and demographic group. The demographic groups are broken down by sex (m, f) and age (0--14, 15--25, 25--34, 35--44, 45--54, 55--64, unknown).

\begin{table}[htbp]
  \centering
  \input{data/tb-raw.tex}
  \caption{Original TB dataset. Corresponding to each `m' column for males, there is also an `f' column for females: {\tt f1524}, {\tt f2534} and so on. These are not shown to conserve space. Note the mixture of 0s and missing values (---): this is due to the data collection process and the distinction is important for this dataset.}
  \label{tbl:tb-raw}
\end{table}

Column headers in this format are often separated by some character ({\tt .}, {\tt -}, {\tt \_}, {\tt :}) and the string can be broken into pieces using that character as a divider. In other cases, such as for this dataset, more careful string processing is required, or the variable names need to be matched to a lookup table that converts single compound values to values of multiple variables.

Table~\subref*{tbl:tb-molten} shows the results of melting the TB dataset, and Table~\subref*{tbl:tb-tidy} the results of splitting the column ``column'' into two real variables: age and sex.

\begin{table}[htbp]
  \centering
  \subfloat[Molten data]{\label{tbl:tb-molten}\input{data/tb-clean-1.tex}}%
  \hspace{2em}%
  \subfloat[Tidy data]{\label{tbl:tb-tidy}\input{data/tb-clean-2.tex}}

  \caption{Tidying the TB dataset requires first melting, and then splitting the ``column'' column into two variables: sex and age.}
  \label{tbl:tb-clean}
\end{table}

Storing the values in this form resolves another problem in the original data: we want to compare rates, not counts, but to compute rates we need to know the population. In the original format, there is no easy way to add a population variable, and it has to be stored in a separate table, which makes it hard to correctly match populations to counts. In tidy form, adding variables to store population and rate is easy: they just become additional columns.

\subsection{Variables are stored in both rows and columns}

The most complicated form of messy data is when variables have been stored in both rows and columns. Table~\ref{tbl:weather-raw} shows daily weather data from the Global Historical Climatology Network for one weather station (MX17004) in Mexico for five months in 2010. It has variables in individual columns (id, year, month), spread across columns (day, d1--d31) and across rows (tmin, tmax) (minimum and maximum temperature). Months with less than 31 days have structural missing values for the last day(s) of the month. The element column is not a variable: it stores the names of variables.

To tidy this dataset we first melt it with colvars id, year, month and the column that contains variable names, element, yielding Table~\subref*{tbl:weather-molten}. For presentation, we have dropped the missing values, making them implicit rather than explicit. This is permissible because we know how many days are in each month and can easily reconstruct the explicit missing values.

This dataset is mostly tidy, but we have two variables stored in rows: {\tt tmin} and {\tt tmax}, the type of observation. Not shown in this example are other meteorological variables prcp (precipitation) and snow (snowfall). Fixing this requires the cast, or unstack, operation, which performs the inverse of melting, rotating the {\tt element} variable back out into the columns to give Table~\subref*{tbl:weather-tidy}. This form is tidy: there is one variable in each column, and each row represents a day's observations. The cast operation is described in depth in \citet{wickham:2007b}.

\begin{table}[htbp]
  \centering
  \input{data/weather-raw.tex}
  \caption{Original weather dataset.  There is a column for each possible day in the month.  Columns {\tt d11} to {\tt d31} have been omitted to conserve space.}
  \label{tbl:weather-raw}
\end{table}

\begin{table}[htbp]
  \centering
  \subfloat[Molten data]%
    {\label{tbl:weather-molten}\input{data/weather-clean-1.tex}}%
  \hspace{2em}%
  \subfloat[Tidy data]%
    {\label{tbl:weather-tidy}\input{data/weather-clean-2.tex}}%

  \caption{(a) Molten weather dataset.  This is almost tidy, but the {\tt element} column contains names of variables, not values. Missing values are dropped to conserve space. (b) Tidy weather dataset. Each row representation the meteorological measurements for a single day.  There are two measured variables, minimum ({\tt tmin}) and maximum ({\tt tmax}) temperature; all other variables are fixed.}
  \label{tbl:weather-clean}
\end{table}

\subsection{Multiple types in one table}
\label{sub:multiple-types}

Datasets often involve values collected at multiple levels, on different types of observational unit. During tidying, each type of observational unit should be stored in its own table. This is closely related to the idea of database normalisation, where each fact is expressed in only one place; if not, it's possible for inconsistencies to occur.

%If you're not familiar with normalisation, it can be worthwhile to learn a little about it. There are many good tutorials available online - I found \url{http://phlonx.com/resources/nf3/} after a few minutes of searching. You certainly don't need to become an expert, as most statistical databases only need a small amount of normalisation, but it is extremely helpful for identifying inconsistencies in your data.

The billboard dataset described in Table~\ref{tbl:billboard-clean} actually contains observations on two types of observational unit: the song, and its rank in each week. This is revealed through the duplication of facts about the song: the artist and time are repeated for every song in each week. The billboard dataset needs to be broken down into two datasets: a song dataset which stores artist, song name and time; and a ranking dataset which gives the rank of the song in each week. Table~\ref{tbl:billboard-normal} shows these two datasets. You could also imagine a week dataset which would record background information about the week, maybe the total number of songs sold or similar ``demographic'' information.

\begin{table}
  \centering
  \input{data/billboard-song.tex}\hspace{2em}%
  \input{data/billboard-rank.tex}

  \caption{Normalised billboard dataset split up into song dataset (left) and rank dataset (right). First 15 rows of each dataset shown; {\tt genre} omitted from song dataset, {\tt week} omitted from rank dataset.}
  \label{tbl:billboard-normal}
\end{table}

Normalisation is useful for tidying and eliminating inconsistencies, but there are few data analysis tools that work directly with relational data, so analysis usually requires denormalisation, merging the datasets back into one table.

% Multiple-choice/check all that apply data

\subsection{One type in multiple tables}

It's also common to find data values about a single type of observational unit spread out over multiple tables, or multiple files. These tables are often split up by another variable, so each file represents a single year, person, or location. As long as the format for individual records is consistent, this is an easy problem to fix: read in all files, add a new column that records the original file name (because the filename is often the value of an important variable), then combine all tables into a single table. Once you have a single table, you can perform additional tidying as needed. An example of this type of cleaning can be found at \url{https://github.com/hadley/data-baby-names} which takes 129 yearly baby name tables provided by the US Social Security Administration and combines them into a single file.

A more complicated situation occurs when the dataset structure has changed over time: different variables, variable names, data storage formats, or different conventions for missing values. This may require each file to tidied individually (or hopefully in small groups) and then combined together. An example of this type of tidying is illustrated in \url{https://github.com/hadley/data-fuel-economy}, which shows the tidying of {\sc epa} fuel economy data for over 50,000 cars from 1978 to 2008. The raw data is available online, but each year is stored in a separate file and there are four major formats with many minor variations, making tidying this dataset a considerable challenge.

\section{Tidy tools}
\label{sec:tidy-tools}

Once you have a tidy dataset, what can you do with it? Tidy data is only worthwhile if it makes analysis easier. This section discusses tidy tools, tools that take tidy datasets as input and return tidy datasets as output. Tidy tools are useful because the output of one tool can be used as the input to another, making it straightforward to compose multiple tools to solve a problem. Tidy data also ensures that variables are stored in a consistent, explicit manner. This makes each tool simpler, because it doesn't need a Swiss Army knife of parameters for dealing with different dataset structures.

Tools can be messy for two reasons: either they take messy datasets as input (messy-input tools) or they produce messy datasets as output (messy-output tools). Messy-input tools are typically more complicated than tidy-input tools because they need to include some parts of the tidying process. This can be useful for common types of messy datasets, but it typically makes the function more complex, harder to use and harder to maintain. Messy-output tools are frustrating and slow down analysis because they can not be easily composed and you must constantly think about how to convert from one format to another. We'll see examples of both in the following sections.

Next, I give examples of tidy and messy tools for three important components of analysis: data manipulation, visualisation and modelling. I will focus particularly on tools provided by R \citep{R}, because it has many existing tidy tools, but I will also touch on other statistical programming environments.

\subsection{Manipulation}

Data manipulation includes variable-by-variable transformation (e.g. {\tt log} or {\tt sqrt}), as well as aggregation, filtering and reordering. In my experience, there are four extremely common operations that are performed over and over again in the course of an analysis. These are the four fundamental verbs of data manipulation:

\begin{itemize}

  \item Filtering, or subsetting, where observations are removed based on some
  condition.

  \item Transforming, where new variables are added or existing variables
  modified. These modifications can involve either a single variable (e.g.\ 
  log-transforming a variable), or involve multiple variables (e.g.\ computing
  density from weight and volume).

  \item Aggregating, where multiple values are collapsed into a single value,
  e.g.\ by summing or taking means.

  \item Sorting, where the order of observations is changed.

\end{itemize}

All these operations are made easier when there is a consistent way to refer to variables. Tidy data provides this because each variable lives in its own column.

In R, filtering and transforming are performed by the base R functions {\tt subset} and {\tt transform}. These are input and output-tidy. The {\tt aggregate} function performs group-wise aggregation, is input-tidy, and is output-tidy providing a single aggregation method is used. The plyr package provides a tidy {\tt summarise} and {\tt arrange} function for aggregation and sorting.

The four verbs can be, and often are, modified by the \textbf{by} preposition. We often need group-wise aggregates, transformations or subsets: pick the biggest in each group, average over replicates and so on. Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time. Many {\sc sas} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy. Base R possesses a {\tt by} function, which is input-tidy, but not output-tidy, because it produces a list. The {\tt ddply} function from the {\tt plyr} package is a tidy alternative.

% Some aggregations occur so frequently they deserve their own optimised implementations. One such operation is (weighted) counting. Base R provides the {\tt table} function for this, but it is not output-tidy: it returns a multidimensional array. An tidy alternative is the {\tt count} function from {\tt plyr}, which returns a tidy dataset with a column for each of the input variables plus a new variable {\tt freq}, which records the number of records in each category.

Other tools are needed when we have multiple datasets. An advantage of tidy data is the ease with which it can be combined with other tidy datasets: we just need a {\bf join} operator which works by matching common variables and adding new columns. This is implemented in the {\tt merge} function in base R, or the {\tt join} function in {\tt plyr}. Compare this to the difficulty of combining datasets stored in arrays; these typically require painstaking alignment before matrix operations can be used, and errors can be very hard to detect.

\subsection{Visualisation}

Tidy visualisation tools need only to be input-tidy as their output is visual. Domain specific languages work particularly well for the visualisation of tidy datasets because they can describe a visualisation as a mapping between variables in the dataset and aesthetic properties of the graphic like position, size, shape and colour. This is the idea behind the grammar of graphics \citep{wilkinson:2006}, and the layered grammar of graphics \citep{wickham:2007d}, an extension tailored specifically for R.

Most graphical tools in R are input tidy, including the base {\tt plot} function, the lattice family of plots \citep{sarkar:2008} and ggplot2 \citep{me:ggplot2}. Some specialised tools exist for visualising messy datasets. Some base R functions like {\tt barplot}, {\tt matplot}, {\tt dotchart}, and {\tt mosaicplot}, work with messy datasets where variables are spread out over multiple columns. Similarly, the parallel coordinates plot \citep{wegman:1990,inselberg:1985} can be used to create time series plots for messy datasets where each time point is a column.

\subsection{Modelling}
\label{sub:modelling}

Modelling is the inspiration that has driven much of this work, because most modelling tools work best with tidy datasets. Every statistical language has a way of describing a model as a connection between different variables, a domain specific language that connects responses to predictors:

\begin{itemize}

  \item R (lm): \verb|y ~ a + b + c * d|

  \item SAS (PROC GLM): \verb|y = a + b + c * d|

  \item SPSS (glm): \verb|y BY a b c d / DESIGN a b c d c*d |

\end{itemize}

This is not to say that tidy data is the format used internally to compute the regression. Significant transformations take place to produce a numeric matrix that can easily be fed to standard linear algebra routines. Common transformations include adding an intercept column (column of all ones), turning categorical variables into multiple binary dummy variables, and for smooth terms like splines, projecting the data values on to the appropriate basis functions.

% http://support.sas.com/documentation/cdl/en/statug/63033/HTML/default/viewer.htm#statug_glm_sect022.htm

There have been some attempts to adapt modelling functions for specific types of messy datasets. For example, in {\sc sas}'s {\sc proc glm}, multiple variables on the response side of the equation will be interpreted as repeated measures if the {\sc repeated} keyword is present. For the raw billboard data, we could construct a model of the form \verb|wk1-wk76 = track| instead of \verb|rank = week * track| on the tidy data (provided week is labelled as a categorical variable).

Another interesting example is the classic paired t-test, which can be computed in two ways depending on how the data is stored. If the data is stored as in Table~\subref*{tbl:paired}, then a paired t-test is just a t-test applied to the mean difference between x and y. If it is stored in the form of Table~\subref*{tbl:mixed}, then an equivalent model is to fit a mixed effects model, with a fixed dummy variable representing the ``variable'', and a random intercept for each id. (In R's lmer4 notation, this is \verb#value ~ variable + (1 | id)#). Either way of modelling the data yields the same result. Without more information we can't say which form of the data is tidy: if x and y represent length of left and right arms, then Table~\subref*{tbl:paired} is tidy, if x and y represent measurements on day 1 and day 10, then Table~\subref*{tbl:mixed} is tidy.

\begin{table}
  \centering
  \subfloat[Data for paired t-test]{
    \label{tbl:paired}
    \input{model-1.tex}
  }%
  \hspace{2em}%
  \subfloat[Data for mixed effects model]{
    \label{tbl:mixed}
    \input{model-2.tex}
  }

  \caption{Two data sets for performing the same test.}
  \label{label}
\end{table}

While model inputs usually require tidy inputs, model outputs, such as predictions and estimated coefficients, aren't always tidy. This makes it more difficult to combine results from multiple models. For example, in R, the default representation of model coefficients is not tidy because it does not have an explicit variable that records the variable name for each estimate: these are instead recorded as row names. In R, row names must be unique, so combining coefficients from many models (e.g.\ from bootstrap resamples, or subgroups) requires workarounds to avoid losing important information.  This knocks you out of the flow of analysis and makes it harder to combine the results from multiple models.

\section{Case study} 
\label{sec:case-study}

The following case study illustrates how tidy data and tidy tools make data analysis easier by easing the transitions between manipulation, visualisation and modelling: you will not see any code that exists solely to get the output of one function into the right format for input to another. I'll show the R code that performs the analysis, but even if you're not familiar with R or the exact idioms I use, I've tried to make it easy to understand by tightly interleaving code, results and explanation.

The case study uses individual-level mortality data from Mexico, with the goal of finding causes of death that have notably different time patterns within a day. The full dataset has information on 539,530 deaths in Mexico in 2008 and 55 variables, including the the location and time of death, the cause of death, and demographics of the deceased. Table~\ref{fig:raw} shows a small sample of the dataset, focussing on variables related to time of death (year, month, day and hour), and cause (cod).

\begin{table}
  \centering
  \input{case-study/raw.tex}
  \caption{A sample of 16 rows and 5 columns from the original dataset of 539,530 rows and 55 columns.}
  \label{fig:raw}
\end{table}

To achieve our goal of finding unusual time courses, we'll start by counting the number of deaths in each hour ({\tt hod}) for each cause ({\tt cod}) with the tidy {\tt count} function. Then we remove missing (and hence uninformative for our purpose) values with {\tt subset}. This gives Table~\subref*{tbl:counts:1}.

\begin{verbatim}
# Count the the number of deaths in each hour of the day for each 
# cause of death.  
hod2 <- count(deaths, c("hod", "cod"))

# Remove missing observations
hod2 <- subset(hod2, !is.na(hod))
\end{verbatim}

Next we join the dataset to {\tt codes} dataset that gives informative labels for the disease. These two datasets are connected by the {\tt cod} variable. This adds a new variable, {\tt disease}, shown in Table~\subref*{tbl:counts:2}).

\begin{verbatim}
hod2 <- join(hod2, codes, by = "cod")
\end{verbatim}

The total deaths for each cause vary over several orders of magnitude: there were 46,794 deaths from heart attack, and 10 deaths from avalanche. This means that it makes more sense to compare the proportion of deaths in each hour, rather than the total number. We compute this by breaking the dataset down by {\tt cod}, and then {\tt transform}ing to add a new {\tt prop} column, the hourly frequency divided by the total number of deaths from that cause. This new columns is Table~\subref*{tbl:counts:3}.

\begin{verbatim}
# ddply breaks down its first argument (hod2) by its second argument (the
# cod variable), and apply its third argument (transform) to each piece. The
# fourth argument (prop = freq / sum(freq)) is passed on to transform.
hod2 <- ddply(hod2, "cod", transform, prop = freq / sum(freq))
\end{verbatim}

We then compute the overall average death rate for each hour, and merge that back into the original dataset. This yields Table~\subref*{tbl:counts:4} and allows us to easily compare each disease with the overall pattern by comparing {\tt prop} to {\tt prop\_all}.

\begin{verbatim}
# Work out number of people dying each hour by breaking down hod2
# by hod, and summarising by computing total over each cause of death
overall <- ddply(hod2, "hod", summarise, freq_all = sum(freq))

# Work out overall proportion of people dying in each hour
overall <- transform(overall, prop_all = freq_all / sum(freq_all))

# Join overall dataset with individual dataset to make it easier to
# compare the two.
hod2 <- join(hod2, overall, by = "hod")
\end{verbatim}

\begin{table}[htbp]
  \centering
  \subfloat[]{
    \label{tbl:counts:1}\input{case-study/counts.tex}
  }%
  \subfloat[]{
    \label{tbl:counts:2}\input{case-study/counts-disease.tex}
  }%
  \subfloat[]{
    \label{tbl:counts:3}\input{case-study/counts-prop.tex}
  }%
  \subfloat[]{
    \label{tbl:counts:4}\input{case-study/counts-all.tex}
  }

  \caption{A sample of four diseases and four hours from {\tt hod2} data frame.}
  \label{tbl:counts}
\end{table}

Next we compute a distance between the time course of each cause of death and the overall time course. Here we use a simple mean squared deviation.  We also record the sample size, the total number of deaths from that cause.

\begin{verbatim}
# Compute deviation for each disease by summarising with the number
# of deaths, and mean squared deviation from the overall trend.
devi <- ddply(hod2, "cod", summarise, 
  n = sum(freq), 
  dist = mean((prop - prop_all)^2))

# Only work with diseases with more than 50 deaths (~2/hour)
devi <- subset(devi, n > 50)
\end{verbatim}

We don't know the variance characteristics of this estimator, so we explore it visually by plotting n vs.\ deviation, Figure~\ref{fig:deviation-raw}. The linear scale shows little, except the variability decreases with sample size, but on the log-log scale, Figure~\ref{fig:deviation-log} there is a clear pattern.

\begin{verbatim}
# Visualise the devi dataset,
ggplot(data = devi, aes(x = n, y = dist) + geom_point()

# Modify the previous plot, converting to log scales, and adding a
# smoother from a robust linear model.
last_plot() + 
  scale_x_log10() + 
  scale_y_log10() +
  geom_smooth(method = "rlm", se = F)
\end{verbatim}

\begin{figure}[htbp]
  \centering
  \subfloat[Linear scales]{
    \label{fig:deviation-raw}
    \includegraphics[width=0.5\linewidth]{case-study/n-dist-raw.pdf}
  }%
  \subfloat[Log scales]{
    \label{fig:deviation-log}
    \includegraphics[width=0.5\linewidth]{case-study/n-dist-log.pdf}
  }

  \caption{(Left) Plot of n vs deviation. Variability of deviation dominated by the sample size: small samples have large variability. (Right) Log-log plot makes it easy to see pattern of variation as well as unusually high values.  Blue line is a robust line of best fit.}
  \label{fig:deviation}
\end{figure}

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\linewidth]{case-study/n-dist-resid}
  \caption{Residuals from a robust linear model predicting $\log(dist)$ by $\log(n)$. Horizontal line at 1.5 shows threshold for further exploration.}
  \label{fig:devi-resid}
\end{figure}

We are interested in points that are high relative to other causes with the same number of deaths: these are diseases most different from the overall pattern. To find these unusual points, we fit a robust linear model and plot the residuals, as shown in Figure~\ref{fig:devi-resid}. The plot shows an empty region around a residual of 1.5, so somewhat arbitrarily we select those diseases with a residual greater than 1.5. We do that in two steps: first selecting the appropriate rows from {\tt devi} (one row per disease), and then finding the matching time course information in the original summary dataset (24 rows per disease).

\begin{verbatim}
# Compute residuals from robust linear model on log transformed values
devi$resid <- resid(rlm(log(dist) ~ log(n), data = devi))

# Select diseases with residuals bigger than 1.5
unusual <- subset(devi, resid > 1.5)

# Extract matching time course dataset for unusual diseases
hod_unusual <- match_df(hod2, unusual)
\end{verbatim}

Finally, we plot the time courses of the unusual causes in Figure~\ref{fig:disease}. We break the diseases into two plots because of the differences in variability: the top plot shows diseases with over 350 deaths and the bottom with less than 350. The causes of death fall in to three main groups: murders, drowning, and transportation related. Murders are more common at night, drowning in the afternoon, and transportation related during commute times.

\begin{verbatim}
# Visualise unusual deaths dataset hod on x-axis, prop on y-axis, display by line, with pale gray
# line showing overall death rate.
ggplot(data = subset(hod_unusual, n > 350), aes(x = hod, y = prop)) + 
 geom_line(aes(y = prop_all), data = overall, colour = "grey50") +
 geom_line() + 
 facet_wrap(~ disease, ncol = 3)
\end{verbatim}

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.9\textwidth]{case-study/unusual-big}
    \includegraphics[width=0.9\textwidth]{case-study/unusual-sml}
  \caption{Causes of death with unusual time courses. Overall hourly death rate shown in grey. (Top) Causes of death with more than 350 deaths over a year. (Bottom) Causes of death with less than 350 deaths.  Note that different scales are used on the y axis.}
  \label{fig:disease}
\end{figure}

\section{Discussion}
\label{sec:discussion}

Data cleaning is an important problem, but it is an uncommon subject of study in statistics. This paper carves out a small but important subset of data cleaning that I've called data tidying: getting the dataset in the right structure to make further manipulation, visualisation and modelling easy. There is still much work to be done: as well as incremental improvements as my understanding of tidy data and tidy tools improves, there are big improvements possible from exploring alternative formulations of tidy data and tackling other data cleaning problems.

There is a chicken-and-egg problem with tidy data: if tidy data is only as useful as the tools that work with it, then tidy tools are inextricably linked to tidy data. This makes it is easy to get stuck in a local maxima where independently changing data structures or data tools does not result in an improved workflow. Breaking out of this local maxima is hard, and requires long-term concerted effort with many false starts. While I hope that this framework for tidy data is not a false start, equally, I don't see it as the final solution and I hope others will build on this framework to develop even better ways of storing data and better tools for working with it.

I have found few principles to guide the design of tidy data, which must acknowledge both statistical and cognitive factors. To date, my work has been driven by my experience doing data analysis, connections to relational database design, and introspection on the tools of data analysis. The human factors, user-centred design, and human-computer interaction communities seem like they should be able to add to this conversation, but the design of data and tools to work with it has not been an active research topic. In the future, I hope to apply methodologies from these fields (user-testing, ethnography, talk-aloud protocols) to my work to build a better understanding of the cognitive side of data analysis, and how we can best design tools to support the process.

% My definition of tidy data is not exhaustive, and it focusses on the type of data that statisticians most commonly encounter: rectangular data defined by rows and columns. It is not a good fit for data on networks, or entire texts. Additionally, because the focus of this definition is on explicit definition of the data, it may not be suitable for very large datasets, such as microarray or fMRI, which may require special infrastructure for performance.

Other formulations of tidy data are possible. For example, it would be possible to construct a set of tools for dealing with values stored in multidimensional arrays. This is a common data storage format for large biomedical datasets such as microarray or fMRI data. It is also necessary for many multivariate methods whose underlying theory is based on matrix manipulation. This array-tidy format would be compact and efficient, because there are many efficient tools for working with high-dimensional arrays, even when sparse, and the format would connect more closely with the mathematical basis of statistics. This approach is taken by the Pandas python data analysis library \citep{mckinney:2010}. The biggest challenge is to connect different datasets, requiring the development of an efficient equivalent to join, and to connect to existing modelling tools without an expensive re-expression in another format. Even more interestingly, we could consider tidy tools that can ignore the underlying data representation, automatically choosing between array-tidy and dataframe-tidy formats to optimise memory usage and performance.

Apart from tidying, there are many other tasks involved in cleaning data: parsing dates and numbers, identifying missing values, correcting character encodings (for international data), matching similar but not identical values (created by typos), verifying experimental design, filling in structural missing values, not to mention model-based data cleaning that identifies suspicious values. Can we develop other frameworks to make these tasks easier?

% While the tools that power this work grew out of my personal struggle to work with data, the framework that hooks them all together did not develop until I had to teach data cleaning. I could look at a dataset and intuit what needed to be done to it, but I couldn't explain what I was doing, and I found it very difficult to teach. This description of tidy data in this paper easier to teach because students are pretty good at identifying variables and values, and then there is a straightforward path to follow to get data in the right format.

\section{Acknowledgements} 
\label{sec:acknowledgements}

This work wouldn't be possible without the many conversations I've had about data and how to deal with it statistically. I'd particularly like to thank Phil Dixon, Di Cook, and Heike Hofmann, who have put up with numerous questions over the years. I'd also like to thank the users of the reshape package who have provided many challenging problems, and my students who continue to challenge me to explain what I know in a way that they can understand. I'd also like to thank Bob Muenchen, Burt Gunter, Nick Horton and Garrett Grolemund who gave detailed comments on earlier drafts, and to particularly thank Ross Gayler who provided the nice example of the challenges of defining what a variable is and Ben Bolker who showed me the natural equivalent between a paired t-test and a mixed effects model.

% bibtool -x tidy-data.aux -c > references.bib 
\bibliography{references}

\end{document}

### 2013-02-19 case-study/case-study.r

@@ -1,0 +1,93 @@
+options(stringsAsFactors = F)
+library(reshape2)
+library(ggplot2)
+library(plyr)
+library(stringr)
+library(MASS)
+source("../data/xtable.r")
+
+# Load data and do basic cleaning for this task
+if (!exists("deaths", inherit = FALSE)) {
+  load("deaths.rdata")
+  deaths$hod[deaths$hod == 99] <- NA
+  deaths$hod[deaths$hod == 24] <- 0
+  deaths$hod[deaths$hod == 0] <- NA
+  deaths$hod <- as.integer(deaths$hod)  
+  deaths <- arrange(deaths, yod, mod, dod, hod, cod)
+}
+
+ok <- subset(deaths, yod == 2008 & mod != 0 & dod != 0)
+xtable(ok[c(1, 1:14 * 2000), c("yod", "mod", "dod", "hod", "cod")], 
+  "raw.tex")
+
+codes <- read.csv("icd-main.csv")
+codes$disease <- sapply(codes$disease, function(x)
+  str_c(strwrap(x, width = 30), collapse = "\n"))
+names(codes)[1] <- "cod"
+codes <- codes[!duplicated(codes$cod), ]
+
+# Count deaths per hour, per disease
+hod2 <- count(deaths, c("cod", "hod"))
+hod2 <- subset(hod2, !is.na(hod))
+hod2 <- join(hod2, codes)
+hod2 <- ddply(hod2, "cod", transform, prop = freq / sum(freq))
+
+# Compare to overall abundance
+overall <- ddply(hod2, "hod", summarise, freq_all = sum(freq))
+overall <- mutate(overall, prop_all = freq_all / sum(freq_all))
+
+hod2 <- join(overall, hod2, by = "hod")
+
+# Pick better subset of rows to show
+cods <- join(arrange(count(deaths, "cod"), desc(freq)), codes)
+mutate(tail(subset(cods, freq > 100), 30), disease = str_sub(disease, 1, 30))
+
+hod3 <- subset(hod2, cod %in% c("I21", "N18", "E84", "B16") & hod >= 8 & hod <= 12)[1:15, c("hod", "cod", "disease", "freq", "prop", "freq_all", "prop_all")]
+
+xtable(hod3[c("hod", "cod", "freq")], "counts.tex")
+xtable(hod3[c("disease")], "counts-disease.tex")
[…]

### 2013-02-19 data/billboard.r

@@ -1,0 +1,43 @@
+options(stringsAsFactors = FALSE)
+library(lubridate)
+library(reshape2)
+library(stringr)
+library(plyr)
+source("xtable.r")
+
+raw <- read.csv("billboard.csv")
+raw <- raw[, c("year", "artist.inverted", "track", "time", "date.entered", "x1st.week", "x2nd.week", "x3rd.week", "x4th.week", "x5th.week", "x6th.week", "x7th.week", "x8th.week", "x9th.week", "x10th.week", "x11th.week", "x12th.week", "x13th.week", "x14th.week", "x15th.week", "x16th.week", "x17th.week", "x18th.week", "x19th.week", "x20th.week", "x21st.week", "x22nd.week", "x23rd.week", "x24th.week", "x25th.week", "x26th.week", "x27th.week", "x28th.week", "x29th.week", "x30th.week", "x31st.week", "x32nd.week", "x33rd.week", "x34th.week", "x35th.week", "x36th.week", "x37th.week", "x38th.week", "x39th.week", "x40th.week", "x41st.week", "x42nd.week", "x43rd.week", "x44th.week", "x45th.week", "x46th.week", "x47th.week", "x48th.week", "x49th.week", "x50th.week", "x51st.week", "x52nd.week", "x53rd.week", "x54th.week", "x55th.week", "x56th.week", "x57th.week", "x58th.week", "x59th.week", "x60th.week", "x61st.week", "x62nd.week", "x63rd.week", "x64th.week", "x65th.week", "x66th.week", "x67th.week", "x68th.week", "x69th.week", "x70th.week", "x71st.week", "x72nd.week", "x73rd.week", "x74th.week", "x75th.week", "x76th.week")]
+names(raw)[2] <- "artist"
+
+raw$artist <- iconv(raw$artist, "MAC", "UTF-8")
+raw$track <- str_replace(raw$track, " \\(.*?\\)", "")
+names(raw)[-(1:5)] <- str_c("wk", 1:76)
+raw <- arrange(raw, year, artist, track)
+
+xtable(raw[c(1:3, 6:10), 1:8], "billboard-raw.tex")
+
+clean <- melt(raw, id = 1:5, na.rm = T)
+clean$week <- as.integer(str_replace_all(clean$variable, "[^0-9]+", ""))
+clean$variable <- NULL
+
+clean$date.entered <- ymd(clean$date.entered)
+clean$date <- clean$date.entered + weeks(clean$week - 1)
+clean$date.entered <- NULL
[…]

### 2013-02-19 data/melt.r

@@ -1,0 +1,9 @@
+library(reshape2)
+source("xtable.r")
+
+df <- data.frame(row = letters[1:3], a = 1:3, b = 4:6, c = 7:9)
+xtable(df, "melt-raw.tex")
+
+dfm <- melt(df, id = "row")
+names(dfm)[2] <- "column"
+xtable(dfm, "melt-output.tex")

### 2013-02-19 data/pew.r

@@ -1,0 +1,46 @@
+library(foreign)
+library(stringr)
+library(plyr)
+library(reshape2)
+source("xtable.r")
+
+# Data from http://pewforum.org/Datasets/Dataset-Download.aspx
+
+# Load data -----------------------------------------------------------------
+
+pew <- read.spss("pew.sav")
+pew <- as.data.frame(pew)
+
+
+religion <- pew[c("q16", "reltrad", "income")]
+religion$reltrad <- as.character(religion$reltrad)
+religion$reltrad <- str_replace(religion$reltrad, " Churches", "")
+religion$reltrad <- str_replace(religion$reltrad, " Protestant", " Prot")
+religion$reltrad[religion$q16 == " Atheist (do not believe in God) "] <- "Atheist"
+religion$reltrad[religion$q16 == " Agnostic (not sure if there is a God) "] <- "Agnostic"
+religion$reltrad <- str_trim(religion$reltrad)
+religion$reltrad <- str_replace_all(religion$reltrad, " \\(.*?\\)", "")
+
+religion$income <- c("Less than $10,000" = "<$10k", 
+  "10 to under $20,000" = "$10-20k", 
+  "20 to under $30,000" = "$20-30k", 
+  "30 to under $40,000" = "$30-40k", 
+  "40 to under $50,000" = "$40-50k", 
+  "50 to under $75,000" = "$50-75k",
+  "75 to under $100,000" = "$75-100k", 
+  "100 to under $150,000" = "$100-150k", 
+  "$150,000 or more" = ">150k", 
+  "Don't know/Refused (VOL)" = "Don't know/refused")[religion$income]
+
+religion$income <- factor(religion$income, levels = c("<$10k", "$10-20k", "$20-30k", "$30-40k", "$40-50k", "$50-75k", 
+  "$75-100k", "$100-150k", ">150k", "Don't know/refused"))
+
+counts <- count(religion, c("reltrad", "income"))
+names(counts)[1] <- "religion"
+
+xtable(counts[1:10, ], file = "pew-clean.tex")
+
+# Convert into the form in which I originally saw it -------------------------
+
+raw <- dcast(counts, religion ~ income)
+xtable(raw[1:10, 1:8], file = "pew-raw.tex")

### 2013-02-19 data/read-fwf.r

@@ -1,0 +1,13 @@
+read.fwf2 <- function(path, cols) {
+  raw_stations <- readLines(path)
+  stations <- data.frame(matrix(ncol = 0, nrow = length(raw_stations)))
+
+  for(i in 1:nrow(cols)) {
+    field <- cols[i, ]
+    stations[[field$name]] <- str_trim(str_sub(raw_stations, field$start, field$end))
+  }
+  stations[stations == ""] <- NA
+  stations[] <- lapply(stations, type.convert, as.is = TRUE)
+  
+  stations
+}

### 2013-02-19 data/tb.r

@@ -1,0 +1,40 @@
+library(reshape2)
+library(plyr)
+library(stringr)
+source("xtable.r")
+options(stringsAsFactors = FALSE)
+
+# Load -----------------------------------------------------------------------
+raw <- read.csv("tb.csv")
+raw$new_sp <- NULL
+raw <- subset(raw, year == 2000)
+names(raw)[1] <- "country"
+
+names(raw) <- str_replace(names(raw), "new_sp_", "")
+raw$m04 <- NULL
+raw$m514 <- NULL
+raw$f04 <- NULL
+raw$f514 <- NULL
+
+xtable(raw[1:10, 1:11], file = "tb-raw.tex")
+
+# Melt -----------------------------------------------------------------------
+
+clean <- melt(raw, id = c("country", "year"), na.rm = TRUE)
+names(clean)[3] <- "column"
+names(clean)[4] <- "cases"
+
+clean <- arrange(clean, country, column, year)
+xtable(clean[1:15, ], file = "tb-clean-1.tex")
+
+# Break up variable in to sex and age ----------------------------------------
+
+clean$sex <- str_sub(clean$column, 1, 1)
+
+ages <- c("04" = "0-4", "514" = "5-14", "014" = "0-14", "1524" = "15-24", "2534" = "25-34", "3544" = "35-44", "4554" = "45-54", "5564" = "55-64", "65"= "65+", "u" = NA)
+
+clean$age <- factor(ages[str_sub(clean$column, 2)], levels = ages)
+
+clean <- clean[c("country", "year", "sex", "age", "cases")]
+
+xtable(clean[1:15, ], file = "tb-clean-2.tex")

### 2013-02-19 data/weather.r

@@ -1,0 +1,50 @@
+library(stringr)
+library(reshape2)
+library(plyr)
+source("xtable.r")
+source("read-fwf.r")
+options(stringsAsFactors = FALSE)
+
+# Define format for fixed width file
+cols <- data.frame(
+  name =  c("id", "year", "month", "element"),
+  start = c(1,     12,    16,      18),
+  end =   c(11,    15,    17,      21))
+
+names <- str_c(c("value", "mflag", "qflag", "sflag"), rep(1:31, each = 4), sep = "_")
+starts <- cumsum(c(22, rep(c(5, 1, 1, 1), 31)))
+starts <- starts[-length(starts)]
+ends <- c(starts[-1], starts[length(starts)] + 1) - 1
+
+values <- data.frame(name = names, start = starts, end = ends)
+cols <- rbind(cols, values)
+
+# Load data and subset to small example
+raw <- read.fwf2("weather.txt",  cols)
+raw <- subset(raw, year == 2010 & element %in% c("TMIN", "TMAX")) 
+raw <- raw[, c(1:4, which(str_detect(names(raw), "value")))]
+raw$id <- str_c(str_sub(raw$id, 1, 2), str_sub(raw$id, -5, -1))
+
+names(raw)[-(1:4)] <- str_c("d", 1:31)
+raw[raw == -9999] <- NA
+raw[-(1:4)] <- raw[-(1:4)] / 10 
+rownames(raw) <- NULL
+raw$element <- tolower(raw$element)
+
+xtable(raw[1:10, 1:14], file = "weather-raw.tex", digits = 1)
+
+# Melt and tidy
+
+clean1 <- melt(raw, id = 1:4, na.rm = T)
+clean1$day <- as.integer(str_replace(clean1$variable, "d", ""))
+clean1$date <- as.Date(ISOdate(clean1$year, clean1$month, clean1$day))
+
+clean1 <- clean1[c("id", "date", "element", "value")]
+clean1 <- arrange(clean1, date, element)
+clean1$date <- as.character(clean1$date) # work around xtable bug
+xtable(clean1[1:10, ], file = "weather-clean-1.tex", digits = 1)
+
+# Cast
+
+clean2 <- dcast(clean1, ... ~ element)
+xtable(clean2[1:10, ], file = "weather-clean-2.tex", digits = 1)

### 2013-02-19 data/xtable.r

@@ -1,0 +1,13 @@
+# Change defaults for xtable to be more attractive
+# Inspired by: http://cameron.bracken.bz/sweave-xtable-booktabs
+library(xtable)
+
+xtable <- function(x, file = "", ...){
+  table <- xtable::xtable(x, ...)
+  print(table, floating = F, hline.after = NULL, 
+    add.to.row = list(pos = list(-1,0, nrow(x)), 
+    command = c('\\toprule\n ','\\midrule\n  ','\\bottomrule\n')),
+    include.rownames = FALSE, NA.string = "---",
+    file = file
+  )
+}

### 2013-02-19 t-test.r

@@ -1,0 +1,23 @@
+# From Ben Bolker
+library(reshape2)
+library(lme4)
+set.seed(1001)
+source("data/xtable.r")
+
+x <- rnorm(5, 20, 1)
+y <- x + rnorm(5, 2, 1)
+df1 <- data.frame(id = seq_along(y), x = x, y = y)
+df2 <- melt(df1, id = "id")
+
+xtable(df1, file = "model-1.tex")
+xtable(df2, file = "model-2.tex")
+
+t1 <- t.test(df1$x, df1$y, paired = TRUE)
+m1 <- lmer(value ~ variable + (1 | id) , data = df2, REML = TRUE)
+
+all.equal(abs(t1$statistic), coef(summary(m1))["variabley","t value"])
+
+# The t statistic is (almost) the same. (all.equal() reports a relative
+# difference of 4.618215e-07). REML=TRUE isn't necessary (it's the
+# default) but it emphasizes the fact that the paired t test is exactly
+# equivalent to REML.

### 2013-06-05 tidy-data.tex
Changes for JSS style

@@
 \author{Hadley Wickham\\RStudio}
-\title{Tidy data}
+\title{Tidy Data}
 
 \Plainauthor{Hadley Wickham}
-\Plaintitle{Tidy data}
+\Plaintitle{Tidy Data}
 
@@
 \usepackage{booktabs}
-\usepackage{verbatim}
+\usepackage{fancyvrb}
+\renewcommand{\FancyVerbFormatLine}[1]{R> #1}
 \usepackage{epigraph}
@@
 I have also struggled to impart these skills to my students, so they can tackle problems in the real-world, where there are few constraints on how datasets are organised.
-In the course of these struggles I developed the reshape and reshape2 \citep{wickham:2007b} R packages, but while I could intuitively use the tools, I lacked the framework to make my understanding explicit.
-This paper provides that framework, and provides a consistent ``philosophy of data'' that underlies my work in the plyr \citep{me:plyr} and ggplot2 \citep{me:ggplot2} packages.
+In the course of these struggles I developed the \pkg{reshape} and \pkg{reshape2} \citep{wickham:2007b} \proglang{R} packages, but while I could intuitively use the tools, I lacked the framework to make my understanding explicit.
+This paper provides that framework, and provides a consistent ``philosophy of data'' that underlies my work in the \pkg{plyr} \citep{me:plyr} and \pkg{ggplot2} \citep{me:ggplot2} packages.
 
@@
 
-Statistical datasets usually come in the form of a rectangular \textbf{table}, made up of \textbf{rows} and \textbf{columns}.
+Statistical datasets usually come in the form of a rectangular table, made up of rows and columns.
 Tables are usually labelled with column names, and sometimes with row names.
-The table is a collection of data \textbf{value}s, typically either numeric (if quantitative) or character (if qualitative).
-Values are grouped into \textbf{variables}, measurements of a single quantifiable property.
-Multiple measurements made on the same \textbf{observational unit} form an \textbf{observation}.
+The table is a collection of data values, typically either numeric (if quantitative) or character (if qualitative).
+Values are grouped into variables, measurements of a single quantifiable property.
+Multiple measurements made on the same observational unit form an observation.
 
@@
 If the columns were home phone and work phone, we could treat these as two variables, but in a fraud detection environment we might want variables phone number and number type because the use of one phone number for multiple people might suggest fraud.
-A general rule of thumb is that it is easier to describe functional relationships between variables/columns (e.g.\ z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of rows/observations (e.g.\ average of group a vs.\ average of group b) than between columns.
+A general rule of thumb is that it is easier to describe functional relationships between variables/columns (e.g., z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of rows/observations (e.g., average of group a vs.\ average of group b) than between columns.
 If you find yourself fighting these constraints, you may not have correctly identified the appropriate variables for your problem.
 
-In a given analysis, there may also be multiple \textbf{types} of observational unit.
+In a given analysis, there may also be multiple types of observational unit.
 For example, in a trial of new allergy medication we might have three observational types: demographic data collected on \emph{each person} (age, sex, race), on \emph{each person on each day} of hay fever season (number of sneezes, redness of eyes), and on \emph{each day} (temperature, pollen count).
 
-A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types. In {\bf tidy data}:
+A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types. In tidy data:
 
@@
 \noindent This is Codd's 3rd normal form \citep{codd:1990}, but with the constraints framed in statistical language, and the focus on a single dataset, not the many connected datasets common in relational databases.
-I will call other arrangements of data {\bf messy}.
+I will call other arrangements of data messy.
 
@@
 One way of organising variables is by their role in the analysis: are values fixed by the design of the data collection, or are they measured during the course of the experiment?
-\textbf{Fixed} variables describe the experimental design and are known in advance.
+Fixed variables describe the experimental design and are known in advance.
 These are often called dimensions by computer scientists, and typically denoted by subscripts on random variables.
-\textbf{Measured} variables are what we actually measure in the study.
+Measured variables are what we actually measure in the study.
 Fixed variables should come first, followed by measured variables, each ordered so that related variables are contiguous.
@@
 The following sections illustrate each problem with a real dataset that I have encountered in practice, and show how to tidy it.
-The complete datasets and the R code used to tidy them are available online at \url{https://github.com/hadley/tidy-data}, and in the online supplementary materials for this paper.
+The complete datasets and the \proglang{R} code used to tidy them are available online at \url{https://github.com/hadley/tidy-data}, and in the online supplementary materials for this paper.
 
@@
 The Pew dataset has one colvar, religion, and melting yields Table~\ref{tbl:pew-clean}.
-The column variable has been renamed to {\tt income}, and the value column to {\tt freq}, to better reflect their roles in this dataset.
+The column variable has been renamed to \code{income}, and the value column to \code{freq}, to better reflect their roles in this dataset.
 This form is tidy: each column represents a variable and each row represents an observation, in this case a demographic unit corresponding to a combination of religion and income.
@@
 It has variables artist, track, date.entered, rank and week.
-The rank in each week after it enters the top 100 is recorded in 75 columns, {\tt wk1} to {\tt wk75}.
+The rank in each week after it enters the top 100 is recorded in 75 columns, \code{wk1} to \code{wk75}.
 If a song is in the top 100 for less than 75 weeks the remaining columns are filled with missing values.
@@
   \input{data/billboard-raw.tex}
-  \caption{The first eight billboard top hits for 2000. Other columns not shown are {\tt wk4}, {\tt wk5}, ..., {\tt wk75}.}
+  \caption{The first eight billboard top hits for 2000. Other columns not shown are \code{wk4}, \code{wk5}, ..., \code{wk75}.}
   \label{tbl:billboard-raw}
@@
   \input{data/billboard-clean.tex}
-  \caption{First fifteen rows of the tidied billboard dataset. The {\tt date} column does not appear in the original table, but can be computed from {\tt date.entered} and {\tt week}.}
+  \caption{First fifteen rows of the tidied billboard dataset. The \code{date} column does not appear in the original table, but can be computed from \code{date.entered} and \code{week}.}
   \label{tbl:billboard-clean}
@@
   \caption{Original TB dataset.
-Corresponding to each `m' column for males, there is also an `f' column for females: {\tt f1524}, {\tt f2534} and so on.
+Corresponding to each `m' column for males, there is also an `f' column for females: \code{f1524}, \code{f2534} and so on.
 These are not shown to conserve space.
@@
 
-Column headers in this format are often separated by some character ({\tt .}, {\tt -}, {\tt \_}, {\tt :}) and the string can be broken into pieces using that character as a divider.
+Column headers in this format are often separated by some character (\code{.}, \code{-}, \code{\_}, \code{:}) and the string can be broken into pieces using that character as a divider.
 In other cases, such as for this dataset, more careful string processing is required, or the variable names need to be matched to a lookup table that converts single compound values to values of multiple variables.
@@
 
-This dataset is mostly tidy, but we have two variables stored in rows: {\tt tmin} and {\tt tmax}, the type of observation.
+This dataset is mostly tidy, but we have two variables stored in rows: \code{tmin} and \code{tmax}, the type of observation.
 Not shown in this example are other meteorological variables prcp (precipitation) and snow (snowfall).
-Fixing this requires the cast, or unstack, operation, which performs the inverse of melting, rotating the {\tt element} variable back out into the columns to give Table~\subref*{tbl:weather-tidy}.
+Fixing this requires the cast, or unstack, operation, which performs the inverse of melting, rotating the \code{element} variable back out into the columns to give Table~\subref*{tbl:weather-tidy}.
 This form is tidy: there is one variable in each column, and each row represents a day's observations.
@@
   \input{data/weather-raw.tex}
-  \caption{Original weather dataset.  There is a column for each possible day in the month.  Columns {\tt d11} to {\tt d31} have been omitted to conserve space.}
+  \caption{Original weather dataset.  There is a column for each possible day in the month.  Columns \code{d11} to \code{d31} have been omitted to conserve space.}
   \label{tbl:weather-raw}
@@
   \caption{(a) Molten weather dataset.
-This is almost tidy, but the {\tt element} column contains names of variables, not values.
+This is almost tidy, but the \code{element} column contains names of variables, not values.
 Missing values are dropped to conserve space.
@@
 Each row representation the meteorological measurements for a single day.
-There are two measured variables, minimum ({\tt tmin}) and maximum ({\tt tmax}) temperature; all other variables are fixed.}
+There are two measured variables, minimum (\code{tmin}) and maximum (\code{tmax}) temperature; all other variables are fixed.}
   \label{tbl:weather-clean}
@@
   \caption{Normalised billboard dataset split up into song dataset (left) and rank dataset (right).
-First 15 rows of each dataset shown; {\tt genre} omitted from song dataset, {\tt week} omitted from rank dataset.}
+First 15 rows of each dataset shown; \code{genre} omitted from song dataset, \code{week} omitted from rank dataset.}
   \label{tbl:billboard-normal}
@@
 Next, I give examples of tidy and messy tools for three important components of analysis: data manipulation, visualisation and modelling.
-I will focus particularly on tools provided by R \citep{R}, because it has many existing tidy tools, but I will also touch on other statistical programming environments.
+I will focus particularly on tools provided by \proglang{R} \citep{R}, because it has many existing tidy tools, but I will also touch on other statistical programming environments.
 
@@
 
-Data manipulation includes variable-by-variable transformation (e.g.
-{\tt log} or {\tt sqrt}), as well as aggregation, filtering and reordering.
+Data manipulation includes variable-by-variable transformation (e.g., \code{log} or \code{sqrt}), as well as aggregation, filtering and reordering.
 In my experience, there are four extremely common operations that are performed over and over again in the course of an analysis.
@@
   \item Transforming, where new variables are added or existing variables
-  modified. These modifications can involve either a single variable (e.g.\ 
-  log-transforming a variable), or involve multiple variables (e.g.\ computing
+  modified. These modifications can involve either a single variable (e.g., 
+  log-transforming a variable), or involve multiple variables (e.g., computing
   density from weight and volume).
@@
   \item Aggregating, where multiple values are collapsed into a single value,
-  e.g.\ by summing or taking means.
+  e.g., by summing or taking means.
 
@@
 
-In R, filtering and transforming are performed by the base R functions {\tt subset} and {\tt transform}.
+In \proglang{R}, filtering and transforming are performed by the base \proglang{R} functions \code{subset} and \code{transform}.
 These are input and output-tidy.
-The {\tt aggregate} function performs group-wise aggregation, is input-tidy, and is output-tidy providing a single aggregation method is used.
-The plyr package provides a tidy {\tt summarise} and {\tt arrange} function for aggregation and sorting.
+The \code{aggregate} function performs group-wise aggregation, is input-tidy, and is output-tidy providing a single aggregation method is used.
+The \pkg{plyr} package provides a tidy \code{summarise} and \code{arrange} function for aggregation and sorting.
 
-The four verbs can be, and often are, modified by the \textbf{by} preposition.
+The four verbs can be, and often are, modified by the by preposition.
 We often need group-wise aggregates, transformations or subsets: pick the biggest in each group, average over replicates and so on.
 Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time.
-Many {\sc sas} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy.
-Base R possesses a {\tt by} function, which is input-tidy, but not output-tidy, because it produces a list.
-The {\tt ddply} function from the {\tt plyr} package is a tidy alternative.
+Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy.
+Base \proglang{R} possesses a \code{by} function, which is input-tidy, but not output-tidy, because it produces a list.
+The \code{ddply} function from the \pkg{plyr} package is a tidy alternative.
 
@@
 Other tools are needed when we have multiple datasets.
-An advantage of tidy data is the ease with which it can be combined with other tidy datasets: we just need a {\bf join} operator which works by matching common variables and adding new columns.
-This is implemented in the {\tt merge} function in base R, or the {\tt join} function in {\tt plyr}.
+An advantage of tidy data is the ease with which it can be combined with other tidy datasets: we just need a join operator which works by matching common variables and adding new columns.
+This is implemented in the \code{merge} function in base \proglang{R}, or the \code{join} function in \pkg{plyr}.
 Compare this to the difficulty of combining datasets stored in arrays; these typically require painstaking alignment before matrix operations can be used, and errors can be very hard to detect.
@@
 Domain specific languages work particularly well for the visualisation of tidy datasets because they can describe a visualisation as a mapping between variables in the dataset and aesthetic properties of the graphic like position, size, shape and colour.
-This is the idea behind the grammar of graphics \citep{wilkinson:2006}, and the layered grammar of graphics \citep{wickham:2007d}, an extension tailored specifically for R.
+This is the idea behind the grammar of graphics \citep{wilkinson:2006}, and the layered grammar of graphics \citep{wickham:2007d}, an extension tailored specifically for \proglang{R}.
 
-Most graphical tools in R are input tidy, including the base {\tt plot} function, the lattice family of plots \citep{sarkar:2008} and ggplot2 \citep{me:ggplot2}.
+Most graphical tools in \proglang{R} are input tidy, including the \pkg{base} \code{plot} function, the \pkg{lattice} family of plots \citep{sarkar:2008} and \pkg{ggplot2} \citep{me:ggplot2}.
 Some specialised tools exist for visualising messy datasets.
-Some base R functions like {\tt barplot}, {\tt matplot}, {\tt dotchart}, and {\tt mosaicplot}, work with messy datasets where variables are spread out over multiple columns.
+Some base \proglang{R} functions like \code{barplot}, \code{matplot}, \code{dotchart}, and \code{mosaicplot}, work with messy datasets where variables are spread out over multiple columns.
 Similarly, the parallel coordinates plot \citep{wegman:1990,inselberg:1985} can be used to create time series plots for messy datasets where each time point is a column.
@@
 
-  \item R (lm): \verb|y ~ a + b + c * d|
+  \item \proglang{R} (\code{lm}): \code{y ~ a + b + c * d}
 
-  \item SAS (PROC GLM): \verb|y = a + b + c * d|
+  \item \proglang{SAS} (\code{PROC GLM}): \code{y = a + b + c * d}
 
-  \item SPSS (glm): \verb|y BY a b c d / DESIGN a b c d c*d |
+  \item \proglang{SPSS} (\code{glm}): \code{y BY a b c d / DESIGN a b c d c*d}
 
@@
 There have been some attempts to adapt modelling functions for specific types of messy datasets.
-For example, in {\sc sas}'s {\sc proc glm}, multiple variables on the response side of the equation will be interpreted as repeated measures if the {\sc repeated} keyword is present.
-For the raw billboard data, we could construct a model of the form \verb|wk1-wk76 = track| instead of \verb|rank = week * track| on the tidy data (provided week is labelled as a categorical variable).
+For example, in \proglang{SAS}'s \code{proc glm}, multiple variables on the response side of the equation will be interpreted as repeated measures if the {\sc repeated} keyword is present.
+For the raw billboard data, we could construct a model of the form \code{wk1-wk76 = track} instead of \code{rank = week * track} on the tidy data (provided week is labelled as a categorical variable).
 
@@
 If it is stored in the form of Table~\subref*{tbl:mixed}, then an equivalent model is to fit a mixed effects model, with a fixed dummy variable representing the ``variable'', and a random intercept for each id.
-(In R's lmer4 notation, this is \verb#value ~ variable + (1 | id)#).
+(In \proglang{R}'s lmer4 notation, this is \code{value ~ variable + (1 | id)}).
 Either way of modelling the data yields the same result.
@@
 This makes it more difficult to combine results from multiple models.
-For example, in R, the default representation of model coefficients is not tidy because it does not have an explicit variable that records the variable name for each estimate: these are instead recorded as row names.
-In R, row names must be unique, so combining coefficients from many models (e.g.\ from bootstrap resamples, or subgroups) requires workarounds to avoid losing important information.
+For example, in \proglang{R}, the default representation of model coefficients is not tidy because it does not have an explicit variable that records the variable name for each estimate: these are instead recorded as row names.
+In \proglang{R}, row names must be unique, so combining coefficients from many models (e.g., from bootstrap resamples, or subgroups) requires workarounds to avoid losing important information.
 This knocks you out of the flow of analysis and makes it harder to combine the results from multiple models.
@@
 The following case study illustrates how tidy data and tidy tools make data analysis easier by easing the transitions between manipulation, visualisation and modelling: you will not see any code that exists solely to get the output of one function into the right format for input to another.
-I'll show the R code that performs the analysis, but even if you're not familiar with R or the exact idioms I use, I've tried to make it easy to understand by tightly interleaving code, results and explanation.
+I'll show the \proglang{R} code that performs the analysis, but even if you're not familiar with \proglang{R} or the exact idioms I use, I've tried to make it easy to understand by tightly interleaving code, results and explanation.
 
@@
 
-To achieve our goal of finding unusual time courses, we'll start by counting the number of deaths in each hour ({\tt hod}) for each cause ({\tt cod}) with the tidy {\tt count} function.
-Then we remove missing (and hence uninformative for our purpose) values with {\tt subset}.
+To achieve our goal of finding unusual time courses, we'll start by counting the number of deaths in each hour (\code{hod}) for each cause (\code{cod}) with the tidy \code{count} function.
+Then we remove missing (and hence uninformative for our purpose) values with \code{subset}.
 This gives Table~\subref*{tbl:counts:1}.
 
-\begin{verbatim}
-# Count the the number of deaths in each hour of the day for each 
-# cause of death.  
+First, we count the the number of deaths in each hour of the day for each cause of death, and remove missing observations:
+
+\begin{Verbatim}
 hod2 <- count(deaths, c("hod", "cod"))
+hod2 <- subset(hod2, !is.na(hod))
+\end{Verbatim}
 
-# Remove missing observations
-hod2 <- subset(hod2, !is.na(hod))
-\end{verbatim}
+Next we join the dataset to \code{codes} dataset that gives informative labels for the disease.
+These two datasets are connected by the \code{cod} variable.
+This adds a new variable, \code{disease}, shown in Table~\subref*{tbl:counts:2}).
 
-Next we join the dataset to {\tt codes} dataset that gives informative labels for the disease.
-These two datasets are connected by the {\tt cod} variable.
-This adds a new variable, {\tt disease}, shown in Table~\subref*{tbl:counts:2}).
-
-\begin{verbatim}
+\begin{Verbatim}
 hod2 <- join(hod2, codes, by = "cod")
-\end{verbatim}
+\end{Verbatim}
 
@@
 This means that it makes more sense to compare the proportion of deaths in each hour, rather than the total number.
-We compute this by breaking the dataset down by {\tt cod}, and then {\tt transform}ing to add a new {\tt prop} column, the hourly frequency divided by the total number of deaths from that cause.
+We compute this by breaking the dataset down by \code{cod}, and then \code{transform}ing to add a new \code{prop} column, the hourly frequency divided by the total number of deaths from that cause.
 This new columns is Table~\subref*{tbl:counts:3}.
 
-\begin{verbatim}
-# ddply breaks down its first argument (hod2) by its second argument (the
-# cod variable), and apply its third argument (transform) to each piece. The
-# fourth argument (prop = freq / sum(freq)) is passed on to transform.
+\code{ddply} breaks down its first argument (\code{hod2}) by its second argument (the \code{cod} variable), and applies its third argument (\code{transform}) to each piece.
+The fourth argument (\code{prop = freq / sum(freq)}) is passed on to transform.
+
+\begin{Verbatim}
 hod2 <- ddply(hod2, "cod", transform, prop = freq / sum(freq))
-\end{verbatim}
+\end{Verbatim}
 
 We then compute the overall average death rate for each hour, and merge that back into the original dataset.
-This yields Table~\subref*{tbl:counts:4} and allows us to easily compare each disease with the overall pattern by comparing {\tt prop} to {\tt prop\_all}.
+This yields Table~\subref*{tbl:counts:4} and allows us to easily compare each disease with the overall pattern by comparing \code{prop} to \code{prop\_all}.
 
-\begin{verbatim}
-# Work out number of people dying each hour by breaking down hod2
-# by hod, and summarising by computing total over each cause of death
+First, we work out the number of people dying each hour by breaking down \code{hod2} by \code{hod}, summarising with a total for over each cause of death.
+
+\begin{Verbatim}
 overall <- ddply(hod2, "hod", summarise, freq_all = sum(freq))
+\end{Verbatim}
 
-# Work out overall proportion of people dying in each hour
+Next we work out overall proportion of people dying in each hour:
+
+\begin{Verbatim}
 overall <- transform(overall, prop_all = freq_all / sum(freq_all))
+\end{Verbatim}
 
-# Join overall dataset with individual dataset to make it easier to
-# compare the two.
+Then finally join the overall dataset with the individual dataset to make it easier to compare the two:
+
+\begin{Verbatim}
 hod2 <- join(hod2, overall, by = "hod")
-\end{verbatim}
+\end{Verbatim}
 
@@
 
-  \caption{A sample of four diseases and four hours from {\tt hod2} data frame.}
+  \caption{A sample of four diseases and four hours from \code{hod2} data frame.}
   \label{tbl:counts}
@@
 We also record the sample size, the total number of deaths from that cause.
+To ensure that the disease we work with have a decent amount of underlying data we'll only
+work with diseases with more than 50 deaths in total ($\sim$2/hour).
 
-\begin{verbatim}
-# Compute deviation for each disease by summarising with the number
-# of deaths, and mean squared deviation from the overall trend.
+\begin{Verbatim}
 devi <- ddply(hod2, "cod", summarise, 
@@
 
-# Only work with diseases with more than 50 deaths (~2/hour)
 devi <- subset(devi, n > 50)
-\end{verbatim}
+\end{Verbatim}
 
@@
 The linear scale shows little, except the variability decreases with sample size, but on the log-log scale, Figure~\ref{fig:deviation-log} there is a clear pattern.
+It's particularly easy to see the pattern when we add the line of best fit from a robust linear model.
 
-\begin{verbatim}
-# Visualise the devi dataset,
+\begin{Verbatim}
 ggplot(data = devi, aes(x = n, y = dist) + geom_point()
 
-# Modify the previous plot, converting to log scales, and adding a
-# smoother from a robust linear model.
 last_plot() + 
@@
   geom_smooth(method = "rlm", se = F)
-\end{verbatim}
+\end{Verbatim}
 
@@
 The plot shows an empty region around a residual of 1.5, so somewhat arbitrarily we select those diseases with a residual greater than 1.5.
-We do that in two steps: first selecting the appropriate rows from {\tt devi} (one row per disease), and then finding the matching time course information in the original summary dataset (24 rows per disease).
+We do that in two steps: first selecting the appropriate rows from \code{devi} (one row per disease), and then finding the matching time course information in the original summary dataset (24 rows per disease).
 
-\begin{verbatim}
-# Compute residuals from robust linear model on log transformed values
+\begin{Verbatim}
 devi$resid <- resid(rlm(log(dist) ~ log(n), data = devi))
-
-# Select diseases with residuals bigger than 1.5
 unusual <- subset(devi, resid > 1.5)
-
-# Extract matching time course dataset for unusual diseases
 hod_unusual <- match_df(hod2, unusual)
-\end{verbatim}
+\end{Verbatim}
 
@@
 Murders are more common at night, drowning in the afternoon, and transportation related during commute times.
+The pale gray line in the background shows the time course across all diseases.
 
-\begin{verbatim}
-# Visualise unusual deaths dataset hod on x-axis, prop on y-axis, display by line, with pale gray
-# line showing overall death rate.
+\begin{Verbatim}
 ggplot(data = subset(hod_unusual, n > 350), aes(x = hod, y = prop)) + 
@@
   facet_wrap(~ disease, ncol = 3)
-\end{verbatim}
+\end{Verbatim}
 
@@
 I'd particularly like to thank Phil Dixon, Di Cook, and Heike Hofmann, who have put up with numerous questions over the years.
-I'd also like to thank the users of the reshape package who have provided many challenging problems, and my students who continue to challenge me to explain what I know in a way that they can understand.
+I'd also like to thank the users of the \pkg{reshape} package who have provided many challenging problems, and my students who continue to challenge me to explain what I know in a way that they can understand.
 I'd also like to thank Bob Muenchen, Burt Gunter, Nick Horton and Garrett Grolemund who gave detailed comments on earlier drafts, and to particularly thank Ross Gayler who provided the nice example of the challenges of defining what a variable is and Ben Bolker who showed me the natural equivalent between a paired t-test and a mixed effects model.

### 2014-01-06 data/tb.r
Don't convert Namibia to NA

@@ -5,7 +5,7 @@ source("xtable.r")
 options(stringsAsFactors = FALSE)
 
 # Load -----------------------------------------------------------------------
-raw <- read.csv("tb.csv")
+raw <- read.csv("tb.csv", na.strings = "")
 raw$new_sp <- NULL
 raw <- subset(raw, year == 2000)
 names(raw)[1] <- "country"

### 2014-01-06 reviews/response.md
Collect all reviewers into one (markdown) response

# Reviewer 1

> Details of writing: There is opportunity for improvement and for 
> strengthening the exposition at a local level – in sentences and phrases. 
> This reviewer has just read Wickham's 2011 paper in the same journal on
> "Split-Apply-Combine", and the writing in that paper is good and can serve 
> as a model for polishing the present paper. The differences between the two 
> are in the details, and a thorough editing and tightening of the present 
> manuscript would almost surely pay dividends. Without trying to be 
> comprehensive, I will illustrate the kinds of small changes (including 
> just fixing typos) which taken together would have a positive effect. (R1)

> Abstract: "take in and take out" "input and output"
> p. 1 up 6: "subset" -> "component"

> p. 1 last full sentence: "The reorganization makes...because it conforms to 
> a standard that facilitates well an initial exploration and analysis of the data; you don't..."

> p. 2 line 2: Is "reformulating" an approximation for "munging"?
￼
> p. 2 line 9: -> "an extension..."

> p. 2, 5 up from subtitle: "...techniques with real examples." -- could do 
> nicely.

> p. 2, 1 up from subtitle "...misses and what other approaches might be 
> fruitful to pursue."

> p. 3 2/3 down "...were were..."

> p. 3 bottom. Consider having a new bold subtitle as a guidepost for the 
> reader: Defining tidy data:

> p. 10, up 2. Sentence is garbled.

> p. 11 bottom: (left) and (right) may not be the apt descriptors...

> p. 12, line 9: comma after file name?

> p. 12: This reviewer would prefer to avoid "hopefully" about 13 lines down. 
> More substantively, the sentence is garbled.

> p. 13 middle "...by the by preposition." Should second "by' be bolded? Or in 
> quotes" Same question arises again.

> p. 14 last paragraph. A difficulty is identified; does the author suggest 
> a solution? Reader isn't clear where we are left as we enter the Case Study.

> p. 17 middle -> " diseases we work with have..."

> p. 18 nice plots in an interesting context!

> p. 19 last line "...seem like they should ..." -> "may"

> p. 20 very effective use of exploratory data analysis

> p. 21 down 13: might be clearer if "...an efficient equivalent to join." 
> – bolding or perhaps quotes.

> This reviewer notes (in a spirit of collegiality) that the best writers about 
> data analysis (from a previous generation) – Fred Mosteller, John Tukey,
> David Hoaglin, Paul Velleman, Frank Anscombe (as illustrations) – correctly 
> used data as plural form (and occasionally datum as singular). From that 
> perspective, "the data is ..." grates more than a little ... but 
> regrettably this is a battle that is now lost. Sigh... :) :)

# Reviewer 2

> While the topic is important and the message is clear, I found the tone of
> the paper to be too conversational. The author overuses contractions,
> ambiguous pronouns ("It's often said..."), and casually addresses the reader 
> in second-person form. (R2)

> The author also switches between the use of italics, boldface, quotes, and 
> the LaTeX verbatim environment constantly, which tends to be distracting 
> at times. (R2)

> - The author uses "colvars" in italics when first naming the terminology, and 
> then subsequently mentions the Pew dataset contains one colvar, without 
> italics. (R2)

> - The author does not use formatting on the column names in the caption for 
> Table 9 but does use \verbatim to reference column names in Table 10's 
> caption. (R2)

> On page 13, middle of the page, it would be helpful to use quotes around 
> "by" where applicable, as "modified by the by preposition" does not read well.
> (R2)

> The author overuses colons. In some cases, I found hyphens to be more 
> appropriate, and in others, I would think semicolons might be more suitable.
> Some tables also have unreadable characters in them, such as the degree sign 
> in Table 12's top panel (row 6's artist). Same issue with Table 3's 5th 
> religion. Table 12's caption also describes there being a left and right 
> dataset, whereas the subtables are clearly placed above and below.
> (R2)

> The last sentence in the first paragraph of the discussion needs to be 
> rephrased for clarity.
> (R2)

> - Abstract: "variables are stored in columns, observations in rows, and a 
> single type of..." -- this list isn't logically homogeneous.

> - p.5, 3.1 -- "The Pew Center is an American ... that collects data on 
> attitudes to topics RANGING from religion to the internet"

> - "Billboard" should always be capitalized as it is a proper noun.

> - p.13, section 4.2 - "Tidy visualization tools ONLY NEED to be..."

> - p.16, 4-th line from bottom: "Next, we work out THE overall..."

> - p.16, 2nd line from the bottom: "Then finally, WE join.."

> - p. 17, "To ensure that the diseaseS we work with..."

> - p.19, "The causes of death fall INTO three main groups: ... " There should 
> be a hyphen between "transportation" and "related".

> - p. 19, 2nd paragraph of discussion: "This makes it easy"; remove "is".

> - p. 21, last sentence of first paragraph: "and A BETTER KNOWLEDGE OF how 
> we can best design tools..."

> - p. 21, last paragraph before Acknowledgements "verifying experimental 
> design, AND filling in..."

> - Author information, last page "Adjunct ASSISTANT Professor"

# Reviewer 3

> The author is obviously intimately familiar with the tools he's using in 
> this paper, and he should remember that the readers of this paper are not 
> nearly as familiar with them as he is. (R3)

> Section 2: "When using tables for communication, this so-called 
> "Alabama first" (Wainer 2000) ordering should be abandoned, and replaced 
> with an ordering based on a meaningful variable."
> 
> There doesn't seem to be any justification or explanation for this statement.
> (R3)

> Section 3.1: Since most readers of this paper will want to apply these ideas 
> using R, I think it would be a good idea to show how an R datasets containing 
> the Pew and Billboard data would look, as well as the R code that would be 
> used to melt the datasets. This is especially true for the Billboard data, 
> because additional cleaning was performed. Similar comments apply to the 
> other datasets in the paper. (R3)

> Section 3.5: I feel the examples in this section need to be explained in 
> more detail. The referenced external links contain a large number of files, 
> and it's not really clear which files are relevant to the discussion in this 
> section. (R3)

> Section 4.1: "Compare this to the difficulty of combining datasets stored in 
> arrays; these typically require painstaking alignment before matrix 
> operations can be used, and errors can be very hard to detect."
> Would anyone seriously suggest using matrices to store data which would be 
> combined with other data sets? The use of an index to combine multiple 
> datasets doesn't really seem to be related to tidy data. (R3)

> Section 4.3: I think readers of this paper might find examples from stata 
> more relevant than those from spss. (R3)

> Section 5: I was not able to download the data for the case study from 
> https://raw.github.com/hadley/tidy-data/master/case-study/deaths.rdata; 
> I recieved a 403 Forbidden error. (R3)

> I think there should be a more thorough discussion explaining the goal of 
> analyzing this data set. I'm not familiar with the term "time course", and 
> the single sentence "The case study uses individual-level mortality data f
> rom Mexico, with the goal of finding causes of death that have notably 
> different time patterns within a day." really didn't clarify the goal 
> sufficiently for me. Perhaps an example of an unusual time course would help 
> to clarify things. (R3)

### 2014-01-07 reviews/response.md
Changes from reviewer 1

@@
+I really appreciate the thoughtful comments of all the reviewers. I've worked
+hard to make the paper easier to understand and the reviewer's comments were
+extremely useful.  A point-by-point summary of the comments and my rebuttal
+is below. Minor comments fixed without needing further discussion are listed
+at the end.
+
 # Reviewer 1
@@
 
-> Abstract: "take in and take out" "input and output"
-> p. 1 up 6: "subset" -> "component"
+> p. 1 last full sentence: "The reorganization makes...because it conforms to 
+> a standard that facilitates well an initial exploration and analysis of the 
+> data; you don't..."
 
-> p. 1 last full sentence: "The reorganization makes...because it conforms to 
-> a standard that facilitates well an initial exploration and analysis of the data; you don't..."
+I'm not quite sure what the reviewer meant here, but I've rewritten the 
+paragraph for clarity.
 
 > p. 2 line 2: Is "reformulating" an approximation for "munging"?
-￼
-> p. 2 line 9: -> "an extension..."
 
-> p. 2, 5 up from subtitle: "...techniques with real examples." -- could do 
-> nicely.
-
-> p. 2, 1 up from subtitle "...misses and what other approaches might be 
-> fruitful to pursue."
-
-> p. 3 2/3 down "...were were..."
+Yes, but I think mungling better conveys the inelegance of this operation :)
 
@@
 
-> p. 10, up 2. Sentence is garbled.
-
-> p. 11 bottom: (left) and (right) may not be the apt descriptors...
-
-> p. 12, line 9: comma after file name?
-
-> p. 12: This reviewer would prefer to avoid "hopefully" about 13 lines down. 
-> More substantively, the sentence is garbled.
-
-> p. 13 middle "...by the by preposition." Should second "by' be bolded? Or in 
-> quotes" Same question arises again.
+I've completely reorganised this section and it should be much more clear.
 
@@
 
-> p. 17 middle -> " diseases we work with have..."
-
-> p. 18 nice plots in an interesting context!
-
-> p. 19 last line "...seem like they should ..." -> "may"
-
-> p. 20 very effective use of exploratory data analysis
-
 > p. 21 down 13: might be clearer if "...an efficient equivalent to join." 
 > – bolding or perhaps quotes.
+
+I've removed those comments because I know no longer think that they a true.
 
@@
 > to clarify things. (R3)
+
+# Minor problems
+
+> Abstract: "take in and take out" "input and output"
+
+> p. 1 up 6: "subset" -> "component"
+
+> p. 2 line 9: -> "an extension..."
+
+> p. 2, 5 up from subtitle: "...techniques with real examples." -- could do 
+> nicely.
+
+> p. 2, 1 up from subtitle "...misses and what other approaches might be 
+> fruitful to pursue."
+
+> p. 3 2/3 down "...were were..."
+
+> p. 10, up 2. Sentence is garbled.
+
+> p. 11 bottom: (left) and (right) may not be the apt descriptors...
+
+> p. 12, line 9: comma after file name?
+
+> p. 12: This reviewer would prefer to avoid "hopefully" about 13 lines down. 
+> More substantively, the sentence is garbled.
+
+> p. 13 middle "...by the by preposition." Should second "by' be bolded? Or in 
+> quotes" Same question arises again.
+
+> p. 17 middle -> " diseases we work with have..."
+
+> p. 19 last line "...seem like they should ..." -> "may"

### 2014-01-07 tidy-data.tex
Changes from reviewer 1

@@
 \documentclass[article]{jss}
+\usepackage[utf8]{inputenc}
 
@@
 A huge amount of effort is spent cleaning data to get it ready for data analysis, but there has been little research on how to make data cleaning as easy and effective as possible.
-This paper tackles a small, but important, subset of data cleaning: data ``tidying''.
+This paper tackles a small, but important, component of data cleaning: data ``tidying''.
 Tidy datasets are easy to manipulate, model and visualise, and have a specific structure: variables are stored in columns, observations in rows, and a single type of experimental unit per dataset.
 This framework makes it easy to tidy messy datasets because only a small set of tools are needed to deal with a wide range of un-tidy datasets.
-This structure also makes it easier to develop tidy tools for data analysis, tools that both take in and put out tidy datasets.
+This structure also makes it easier to develop tidy tools for data analysis, tools that both input and output tidy datasets.
 The advantages of a consistent data structure and matching tools are demonstrated with a case study free from mundane data manipulation chores.
@@
 The principles of tidy data provide a standard way of organising the data values within a dataset.
-This makes initial data cleaning easier because you have a \emph{standard} that you can start with: you don't need to reinvent the wheel for each new dataset.
-It also makes it easier to develop data analysis tools that work together seamlessly.
+A standard makes initial data cleaning easier because you don't need to start from scratch each time and you can avoid reinventing the wheel.
+The tidy data standard has been designed to facilitate initial exploration and analysis of the data, and it makes it easier to develop data analysis tools that work together seamlessly.
 Current tools are often inconsistent, so you need to spend time munging the output from one tool into the input for another.
-Tidy datasets and tidy tools work hand in hand to make data analysis easier, allowing you to focus on the interesting domain problem, not the uninteresting details of data structure.
+Tidy datasets and tidy tools work hand in hand to make data analysis easier, allowing you to focus on the interesting domain problem, not the uninteresting logistics of your data.
 
@@
 Computer scientists have also contributed much to the study of data cleaning.
-For example, \citet{lakshmanan:1996} defines a extension to SQL allowing it to operate on messy datasets, \citet{raman:2001} provides a framework for cleaning datasets, and \citet{kandel:2011} develops an interactive tool that makes data cleaning easy, automatically creating the code to clean datasets with a friendly user interface.
-These tools are useful, but are presented in language foreign to most statisticians, fail to give much advice on how datasets should be structured, and lack connections to the tools of data analysis.
+For example, \citet{lakshmanan:1996} defines an extension to SQL allowing it to operate on messy datasets, \citet{raman:2001} provides a framework for cleaning datasets, and \citet{kandel:2011} develops an interactive tool that makes data cleaning easy, automatically creating the code to clean datasets with a friendly user interface.
+These tools are useful, but they are presented in a language foreign to most statisticians, they fail to give much advice on how datasets should be structured, and they lack connections to the tools of data analysis.
 
@@
 Section~\ref{sec:defining} begins by defining the three characteristics that make a dataset tidy.
-Most real world datasets are not tidy, so Section~\ref{sec:tidying} describes the operations needed to make messy datasets tidy, and illustrates the techniques with a range of real-life examples.
+Most real world datasets are not tidy, so Section~\ref{sec:tidying} describes the operations needed to make messy datasets tidy, and illustrates the techniques with a range of real examples.
 Section~\ref{sec:tidy-tools} defines tidy tools, tools that input and output tidy datasets, and discusses how together tidy data and tidy tools make it easier to do a data analysis.
 These principles are illustrated with a small case study in Section~\ref{sec:case-study}.
-Section~\ref{sec:discussion} concludes with a discussion of what this framework misses and other approaches that might be fruitful to pursue.
+Section~\ref{sec:discussion} concludes with a discussion of what this framework misses and what other approaches might be fruitful to pursue.
 
@@
 
-Like families, tidy datasets are all alike, but every messy dataset is messy in its own way. Here we will define the essence of tidy data, and provide important vocabulary for describing datasets.
+Like families, tidy datasets are all alike, but every messy dataset is messy in its own way.
+Tidy datasets provide a standard way to connect the structure of a dataset to its semantics.
+I'll provide some standard vocabulary for describing the structure of a dataset, it's semantics, and then use those definitions to define tidy data.
 
-Statistical datasets usually come in the form of a rectangular table, made up of rows and columns.
-Tables are usually labelled with column names, and sometimes with row names.
-The table is a collection of data values, typically either numeric (if quantitative) or character (if qualitative).
-Values are grouped into variables, measurements of a single quantifiable property.
-Multiple measurements made on the same observational unit form an observation.
+\subsection{Data structure}
 
-There is a little circularity in the definition, which is resolved by defining an observational unit as the smallest constituent atom of the experimental unit.
+Most statistical datasets are rectangular tables made up of rows and columns.
+The columns almost always labelled, and rows are often labelled.
+Table~\ref{tbl:preg-raw} provides some data about an imaginary experiment in a format commonly seen in the wild.
+The table has two columns and three rows, and both rows and columns are labelled.
+
+\begin{table}[htbp]
+  \centering
+  \input{data/preg-raw-1.tex}
+  \caption{Typical presentation dataset.}
+  \label{tbl:preg-raw-1}
+\end{table}
+
+There are many ways to structure the same underlying data.
+Figure~\ref{tbl:preg-raw-2} shows exactly the same data, but the rows and columns have been transposed.
+Describing the structure of a dataset is not enough, we also need some way to describe it's underlying semantics.
+
+\begin{table}[htbp]
+  \centering
+  \input{data/preg-raw-2.tex}
+  \caption{The same data as in Table~\ref{tbl:preg-raw-1} but structured differently.}
+  \label{tbl:preg-raw-2}
+\end{table}
+
+\subsection{Data semantics}
+
+A table is a collection of data \textbf{values}, typically either numbers (if quantitative) or strings (if qualitative).
+Values are grouped into \textbf{variables}, measurements of a single quantifiable property.
+Multiple measurements made on the same observational unit form an \textbf{observation}.
+An observational unit as the smallest constituent atom of the experimental unit.
 For example, in a longitudinal medical study, the experimental unit might be the patient, but the observational unit is the patient at each time point.
@@
 This definition of observational unit leads to variables which are physical properties, like blood pressure and weight, not conditional properties like blood pressure at time 1, or weight of pig 2.
-Structure in the variable names (`trtA', `trtB', `trtC'; `jan1', `feb1', `mar1') is difficult to exploit, and suggests that the names are values of other variables.
 
-% Experimental units are usually independent, and the statistical challenge in an analysis is capturing the dependence of observational units within an experimental unit.
-Failing to do so correctly leads to problems like pseudoreplication \citep{hurlbert:1984}.
-
-Table~\ref{tbl:preg-raw} illustrates a data format commonly seen in the wild.
-Here it is used to store information about an imaginary experiment with two treatments.
-There are two columns and three rows (plus column and row names), defining three variables (name, treatment, result) and either five or six observations, depending on how you think about the missing value.
+The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} has three variables (name, treatment, result), six observations (two for each person), and either five or six values, depending on how you think about the missing value.
+We can make this structure more obvious by putting the variables in the columns and the observations in the rows, as in Figure~\ref{tbl:preg-tidy}.
 
@@
   \centering
-  \begin{tabular}{lrr}
-  \toprule
-               & treatmentA & treatmentB \\
-  \midrule
-  John Smith   &      ---  &            5 \\
-  Jane Doe     &        1  &            4 \\
-  Mary Johnson &        2  &            3 \\
-  \bottomrule
-  \end{tabular}
-  \caption{Typical presentation dataset.}
-  \label{tbl:preg-raw}
+  \input{data/preg-tidy.tex}
+  \caption{Tidied data.}
+  \label{tbl:preg-tidy}
 \end{table}
 
-It's not always obvious what qualifies as a variable, and a variable in one analysis may be a value in another.
-For example, if the columns in the example were were height and weight we would have been happy to call them variables.
+For a given dataset, it's usually easy to figure out what are observations and what are variables, but precisely defining variables and observations in general is surprisingly difficult.
+For example, if the columns in the Table~\ref{fig:preg-raw-1} were height and weight we would have been happy to call them variables.
 If the columns were height and width, it would be less clear cut, as we might think of height and weight as values of a dimension variable.
@@
 A general rule of thumb is that it is easier to describe functional relationships between variables/columns (e.g., z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of rows/observations (e.g., average of group a vs.\ average of group b) than between columns.
-If you find yourself fighting these constraints, you may not have correctly identified the appropriate variables for your problem.
 
-In a given analysis, there may also be multiple types of observational unit.
+Note that in a given analysis, there may be multiple types of observational unit.
 For example, in a trial of new allergy medication we might have three observational types: demographic data collected on \emph{each person} (age, sex, race), on \emph{each person on each day} of hay fever season (number of sneezes, redness of eyes), and on \emph{each day} (temperature, pollen count).
 
-A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types. In tidy data:
+\subsection{Tidy data}
+
+Tidy data is a standard way of mapping the semantics of a dataset to its structure.
+A dataset is \emph{messy} or \emph{tidy} depending on how rows, columns and tables are matched up with observations, variables and types.
+In tidy data:
 
@@
 
-Table~\ref{tbl:preg-tidy} shows the tidy version of Table~\ref{tbl:preg-raw}.
-Now each row represents an observation, the result of one treatment on one person, and each column is a variable.
-We could also choose to drop the row representing the result of John Smith's treatment a, but we don't, because missing values should only be omitted if they are structural: impossible because of the design of the experiment.
-
-\begin{table}[htbp]
-  \centering
-  \begin{tabular}{lrr}
-    \toprule
-    name &    treatment    & result  \\
-    \midrule
-    Jane Doe      & a & 1 \\
-    Jane Doe      & b & 4 \\
-    John Smith    & a & --- \\
-    John Smith    & b & 5 \\
-    Mary Johnson  & a & 2 \\
-    Mary Johnson  & b & 3 \\
-    \bottomrule
-  \end{tabular}
-  \caption{Tidied data.}
-  \label{tbl:preg-tidy}
-\end{table}
+Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw} because each row represents an observation, the result of one treatment on one person, and each column is a variable.
+We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural: impossible because of the design of the experiment.
 
@@
 (b) Tidy weather dataset.
-Each row representation the meteorological measurements for a single day.
+Each row represents the meteorological measurements for a single day.
 There are two measured variables, minimum (\code{tmin}) and maximum (\code{tmax}) temperature; all other variables are fixed.}
@@
   \centering
-  \input{data/billboard-song.tex}\hspace{2em}%
+  \input{data/billboard-song.tex}\hspace{1em}%
   \input{data/billboard-rank.tex}
@@
 These tables are often split up by another variable, so each file represents a single year, person, or location.
-As long as the format for individual records is consistent, this is an easy problem to fix: read in all files, add a new column that records the original file name (because the filename is often the value of an important variable), then combine all tables into a single table.
+As long as the format for individual records is consistent, this is an easy problem to fix:
+
+\begin{enumerate}
+  \item read the files into a list of tables,
+
+\item for each table, add a new column that records the original file name (because the file name is often the value of an important variable), then
+
+  \item combine all tables into a single table. 
+\end{enumerate}
+
 Once you have a single table, you can perform additional tidying as needed.
@@
 
-A more complicated situation occurs when the dataset structure has changed over time: different variables, variable names, data storage formats, or different conventions for missing values.
-This may require each file to tidied individually (or hopefully in small groups) and then combined together.
+A more complicated situation occurs when the dataset structure has changed over time.
+For example, the datasets contain different variables, the same variables have different names, different file formats are used, or there are different conventions for missing values.
+This may require each file to tidied individually (or, if you're lucky, in small groups) and then combined once tidied.
 An example of this type of tidying is illustrated in \url{https://github.com/hadley/data-fuel-economy}, which shows the tidying of {\sc epa} fuel economy data for over 50,000 cars from 1978 to 2008.
@@
 
-The four verbs can be, and often are, modified by the by preposition.
+The four verbs can be, and often are, modified by the ``by'' preposition.
 We often need group-wise aggregates, transformations or subsets: pick the biggest in each group, average over replicates and so on.
@@
 We also record the sample size, the total number of deaths from that cause.
-To ensure that the disease we work with have a decent amount of underlying data we'll only
+To ensure that the diseases we work with have a decent amount of underlying data we'll only
 work with diseases with more than 50 deaths in total ($\sim$2/hour).
@@
 To date, my work has been driven by my experience doing data analysis, connections to relational database design, and introspection on the tools of data analysis.
-The human factors, user-centred design, and human-computer interaction communities seem like they should be able to add to this conversation, but the design of data and tools to work with it has not been an active research topic.
+The human factors, user-centred design, and human-computer interaction communities may be able to add to this conversation, but the design of data and tools to work with it has not been an active research topic.
 In the future, I hope to apply methodologies from these fields (user-testing, ethnography, talk-aloud protocols) to my work to build a better understanding of the cognitive side of data analysis, and how we can best design tools to support the process.
@@
 This approach is taken by the Pandas python data analysis library \citep{mckinney:2010}.
-The biggest challenge is to connect different datasets, requiring the development of an efficient equivalent to join, and to connect to existing modelling tools without an expensive re-expression in another format.
 Even more interestingly, we could consider tidy tools that can ignore the underlying data representation, automatically choosing between array-tidy and dataframe-tidy formats to optimise memory usage and performance.

### 2014-01-07 data/billboard.r

@@ -9,7 +9,7 @@ raw <- read.csv("billboard.csv")
 raw <- raw[, c("year", "artist.inverted", "track", "time", "date.entered", "x1st.week", "x2nd.week", "x3rd.week", "x4th.week", "x5th.week", "x6th.week", "x7th.week", "x8th.week", "x9th.week", "x10th.week", "x11th.week", "x12th.week", "x13th.week", "x14th.week", "x15th.week", "x16th.week", "x17th.week", "x18th.week", "x19th.week", "x20th.week", "x21st.week", "x22nd.week", "x23rd.week", "x24th.week", "x25th.week", "x26th.week", "x27th.week", "x28th.week", "x29th.week", "x30th.week", "x31st.week", "x32nd.week", "x33rd.week", "x34th.week", "x35th.week", "x36th.week", "x37th.week", "x38th.week", "x39th.week", "x40th.week", "x41st.week", "x42nd.week", "x43rd.week", "x44th.week", "x45th.week", "x46th.week", "x47th.week", "x48th.week", "x49th.week", "x50th.week", "x51st.week", "x52nd.week", "x53rd.week", "x54th.week", "x55th.week", "x56th.week", "x57th.week", "x58th.week", "x59th.week", "x60th.week", "x61st.week", "x62nd.week", "x63rd.week", "x64th.week", "x65th.week", "x66th.week", "x67th.week", "x68th.week", "x69th.week", "x70th.week", "x71st.week", "x72nd.week", "x73rd.week", "x74th.week", "x75th.week", "x76th.week")]
 names(raw)[2] <- "artist"
 
-raw$artist <- iconv(raw$artist, "MAC", "UTF-8")
+raw$artist <- iconv(raw$artist, "MAC", "ASCII//translit")
 raw$track <- str_replace(raw$track, " \\(.*?\\)", "")
 names(raw)[-(1:5)] <- str_c("wk", 1:76)
 raw <- arrange(raw, year, artist, track)
@@ -35,7 +35,11 @@ xtable(clean_out[1:15, ], "billboard-clean.tex")
 
 song <- unrowname(unique(clean[c("artist", "track", "time")]))
 song$id <- 1:nrow(song)
-xtable(song[1:15, c("id","artist", "track", "time")], "billboard-song.tex")
+
+narrow <- song[1:15, c("id","artist", "track", "time")]
+long_name <- nchar(narrow$track) > 20
[…]

### 2014-01-07 data/preg.R

@@ -1,0 +1,18 @@
+source("xtable.r")
+library(reshape2)
+set.seed(1014)
+
+preg <- matrix(c(NA, sample(20, 5)), ncol = 2, byrow = T)
+colnames(preg) <- paste0("treatment", c("a", "b"))
+rownames(preg) <- c("John Smith", "Jane Doe", "Mary Johnson")
+
+xtable(preg, "preg-raw-1.tex", rownames = TRUE, align = "lrr")
+xtable(t(preg), "preg-raw-2.tex", rownames = TRUE, align = "lrrr")
+
+# Make tidy version
+
+pregm <- melt(preg, id = "name")
+names(pregm) <- c("name", "trt", "result")
+pregm$trt <- gsub("treatment", "", pregm$trt)
+
+xtable(pregm, "preg-tidy.tex")

### 2014-01-07 data/xtable.r

@@ -2,12 +2,13 @@
 # Inspired by: http://cameron.bracken.bz/sweave-xtable-booktabs
 library(xtable)
 
-xtable <- function(x, file = "", ...){
+xtable <- function(x, file = "", ..., rownames = FALSE){
   table <- xtable::xtable(x, ...)
   print(table, floating = F, hline.after = NULL, 
     add.to.row = list(pos = list(-1,0, nrow(x)), 
     command = c('\\toprule\n ','\\midrule\n  ','\\bottomrule\n')),
-    include.rownames = FALSE, NA.string = "---",
-    file = file
+    include.rownames = rownames, NA.string = "---",
+    file = file, 
+    comment = FALSE, timestamp = FALSE
   )
 }

### 2014-01-08 tidy-data.tex
Feedback from Garrett

@@
 Like families, tidy datasets are all alike, but every messy dataset is messy in its own way.
-Tidy datasets provide a standard way to connect the structure of a dataset to its semantics.
-I'll provide some standard vocabulary for describing the structure of a dataset, it's semantics, and then use those definitions to define tidy data.
+Tidy datasets provide a standard way to connect the structure of a dataset (its physical layout) to its semantics (its meaning).
+I'll provide some standard vocabulary for describing the structure of a dataset, its semantics, and then use those definitions to define tidy data.
 
@@
 
-Most statistical datasets are rectangular tables made up of rows and columns.
-The columns almost always labelled, and rows are often labelled.
-Table~\ref{tbl:preg-raw} provides some data about an imaginary experiment in a format commonly seen in the wild.
+Most statistical datasets are rectangular tables made up of \textbf{columns} and \textbf{rows}.
+The columns are almost always labelled, and rows are often labelled.
+Table~\ref{tbl:preg-raw-1} provides some data about an imaginary experiment in a format commonly seen in the wild.
 The table has two columns and three rows, and both rows and columns are labelled.
@@
 There are many ways to structure the same underlying data.
-Figure~\ref{tbl:preg-raw-2} shows exactly the same data, but the rows and columns have been transposed.
-Describing the structure of a dataset is not enough, we also need some way to describe it's underlying semantics.
+Table~\ref{tbl:preg-raw-2} shows the same data as Table~\ref{tbl:preg-raw-1}, but the rows and columns have been transposed.
+The data is the same, but the layout is different.
+Our vocabulary of rows and columns is not rich enough to say why these two tables represent the same data.
+We also need a way to describe the meaning of the underlying data displayed in the table.
+It's not enough to describe the appearance of a dataset (how the values are physically laid out), we also need a way to describe the underlying semantics, the meaning, of the values in the dataset.
 
@@
 Multiple measurements made on the same observational unit form an \textbf{observation}.
-An observational unit as the smallest constituent atom of the experimental unit.
+An observational unit is the smallest constituent atom of the experimental unit.
 For example, in a longitudinal medical study, the experimental unit might be the patient, but the observational unit is the patient at each time point.
@@
 
-The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} has three variables (name, treatment, result), six observations (two for each person), and either five or six values, depending on how you think about the missing value.
-We can make this structure more obvious by putting the variables in the columns and the observations in the rows, as in Figure~\ref{tbl:preg-tidy}.
+The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} has three variables (name, treatment, result) and six observations (two for each person).
+It has two values of the treatment variable (a and b), three values of the treatment variable, and either five or six values of the result variable, depending on how you think about the missing value.
+We can make this structure more obvious by putting the variables in the columns and the observations in the rows, as in Table~\ref{tbl:preg-tidy}.
 
@@
 For a given dataset, it's usually easy to figure out what are observations and what are variables, but precisely defining variables and observations in general is surprisingly difficult.
-For example, if the columns in the Table~\ref{fig:preg-raw-1} were height and weight we would have been happy to call them variables.
-If the columns were height and width, it would be less clear cut, as we might think of height and weight as values of a dimension variable.
+For example, if the columns in the Table~\ref{tbl:preg-raw-1} were height and weight we would have been happy to call them variables.
+If the columns were height and width, it would be less clear cut, as we might think of height and width as values of a dimension variable.
 If the columns were home phone and work phone, we could treat these as two variables, but in a fraud detection environment we might want variables phone number and number type because the use of one phone number for multiple people might suggest fraud.
-A general rule of thumb is that it is easier to describe functional relationships between variables/columns (e.g., z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of rows/observations (e.g., average of group a vs.\ average of group b) than between columns.
+A general rule of thumb is that it is easier to describe functional relationships between variables (e.g., z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of observations (e.g., average of group a vs.\ average of group b) than between columns.
 
 Note that in a given analysis, there may be multiple types of observational unit.
-For example, in a trial of new allergy medication we might have three observational types: demographic data collected on \emph{each person} (age, sex, race), on \emph{each person on each day} of hay fever season (number of sneezes, redness of eyes), and on \emph{each day} (temperature, pollen count).
+For example, in a trial of new allergy medication we might have three observational types: demographic data collected from each person (age, sex, race), medical data collected from each person on each day (number of sneezes, redness of eyes), and meterological data collected on each day (temperature, pollen count).
 
@@
 
-Tidy data is a standard way of mapping the semantics of a dataset to its structure.
-A dataset is \emph{messy} or \emph{tidy} depending on how rows, columns and tables are matched up with observations, variables and types.
-In tidy data:
+Tidy data is a standard way of mapping the meaning of a dataset to its structure.
+A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types.
+In \textbf{tidy data}:
 
@@
   \item Each observation forms a row,
-  \item Each table (or file) stores information about one observational type.
+  \item Each table stores information on one type of observational unit.
 \end{enumerate}
@@
 \noindent This is Codd's 3rd normal form \citep{codd:1990}, but with the constraints framed in statistical language, and the focus on a single dataset, not the many connected datasets common in relational databases.
-I will call other arrangements of data messy.
+\textbf{Messy data} is any other other arrangement of the data.
 
-Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw} because each row represents an observation, the result of one treatment on one person, and each column is a variable.
+Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw-1} because each row represents an observation, the result of one treatment on one person, and each column is a variable.
 We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural: impossible because of the design of the experiment.
+
+Tidy data makes it easy for an analyst or a computer to extract needed variable because it provides a standard way of structuring a dataset.
+Contrast Table~\ref{tbl:preg-tidy} to Table~\ref{tbl:preg-raw-1}: in Table~\ref{tbl:preg-raw-1} you need to use different strategies for different variables.
+This slows analysis and invites errors.
+If you consider how many data analysis operations involve all of the values in a variable (every aggregating function), you can see how important it is to extract these values in a standard, simple way.
+Tidy data is particularly well suited for vectorised programming languages like \proglang{R}, because the layout ensures that values of different variables from the same observation are always paired.
 
@@
 Fixed variables describe the experimental design and are known in advance.
-These are often called dimensions by computer scientists, and typically denoted by subscripts on random variables.
+Computer scientists often call fixed variables dimensions, and statisticians usually denote them with subscripts on random variables.
 Measured variables are what we actually measure in the study.
 Fixed variables should come first, followed by measured variables, each ordered so that related variables are contiguous.
-Rows can then be ordered by the first variable, breaking ties with the second and subsequent variables.
+Rows can then be ordered by the first variable, breaking ties with the second and subsequent (fixed) variables.
 This is the convention adopted by all tabular displays in this paper.
-When using tables for communication, this so-called ``Alabama first'' \citep{wainer:2000} ordering should be abandoned, and replaced with an ordering based on a meaningful variable.
 

### 2014-01-08 reviews/response.md
Add stata example
Minor changes from R2

@@
 > at times. (R2)
-
 > - The author uses "colvars" in italics when first naming the terminology, and 
@@
 > italics. (R2)
-
 > - The author does not use formatting on the column names in the caption for 
@@
 
-> On page 13, middle of the page, it would be helpful to use quotes around 
-> "by" where applicable, as "modified by the by preposition" does not read well.
-> (R2)
+Thanks for pointing this out. Throughout the next I now use bold face 
+consistently for the first definition of a new term, and verbatim for
+variable names.
 
@@
 > appropriate, and in others, I would think semicolons might be more suitable.
-> Some tables also have unreadable characters in them, such as the degree sign 
-> in Table 12's top panel (row 6's artist). Same issue with Table 3's 5th 
-> religion. Table 12's caption also describes there being a left and right 
-> dataset, whereas the subtables are clearly placed above and below.
-> (R2)
-
-> The last sentence in the first paragraph of the discussion needs to be 
-> rephrased for clarity.
-> (R2)
-
-> - Abstract: "variables are stored in columns, observations in rows, and a 
-> single type of..." -- this list isn't logically homogeneous.
-
-> - p.5, 3.1 -- "The Pew Center is an American ... that collects data on 
-> attitudes to topics RANGING from religion to the internet"
-
-> - "Billboard" should always be capitalized as it is a proper noun.
-
-> - p.13, section 4.2 - "Tidy visualization tools ONLY NEED to be..."
-
-> - p.16, 4-th line from bottom: "Next, we work out THE overall..."
-
-> - p.16, 2nd line from the bottom: "Then finally, WE join.."
-
-> - p. 17, "To ensure that the diseaseS we work with..."
-
-> - p.19, "The causes of death fall INTO three main groups: ... " There should 
-> be a hyphen between "transportation" and "related".
-
-> - p. 19, 2nd paragraph of discussion: "This makes it easy"; remove "is".
-
-> - p. 21, last sentence of first paragraph: "and A BETTER KNOWLEDGE OF how 
-> we can best design tools..."
-
-> - p. 21, last paragraph before Acknowledgements "verifying experimental 
-> design, AND filling in..."
-
-> - Author information, last page "Adjunct ASSISTANT Professor"
 
@@
 > datasets doesn't really seem to be related to tidy data. (R3)
-
-> Section 4.3: I think readers of this paper might find examples from stata 
-> more relevant than those from spss. (R3)
 
@@
 > p. 19 last line "...seem like they should ..." -> "may"
+
+> Section 4.3: I think readers of this paper might find examples from stata 
+> more relevant than those from spss. (R3)
+
+> The last sentence in the first paragraph of the discussion needs to be 
+> rephrased for clarity.
+> (R2)
+
+> Some tables also have unreadable characters in them, such as the degree sign 
+> in Table 12's top panel (row 6's artist). Same issue with Table 3's 5th 
+> religion. Table 12's caption also describes there being a left and right 
+> dataset, whereas the subtables are clearly placed above and below.
+> (R2)
+
+> - Abstract: "variables are stored in columns, observations in rows, and a 
+> single type of..." -- this list isn't logically homogeneous.
+
+> - p.5, 3.1 -- "The Pew Center is an American ... that collects data on 
+> attitudes to topics RANGING from religion to the internet"
+
+> - "Billboard" should always be capitalized as it is a proper noun.
+
+> - p.13, section 4.2 - "Tidy visualization tools ONLY NEED to be..."
+
+> - p.16, 4-th line from bottom: "Next, we work out THE overall..."
+
+> - p.16, 2nd line from the bottom: "Then finally, WE join.."
+
+> - p.19, "The causes of death fall INTO three main groups: ... " There should 
+> be a hyphen between "transportation" and "related".
+
+> - p. 19, 2nd paragraph of discussion: "This makes it easy"; remove "is".
+
+> - p. 21, last sentence of first paragraph: "and A BETTER KNOWLEDGE OF how 
+> we can best design tools..."
+
+> - p. 21, last paragraph before Acknowledgements "verifying experimental 
+> design, AND filling in..."
+
+> - Author information, last page "Adjunct ASSISTANT Professor"

### 2014-01-08 tidy-data.tex
Add stata example
Minor changes from R2

@@
 This paper tackles a small, but important, component of data cleaning: data ``tidying''.
-Tidy datasets are easy to manipulate, model and visualise, and have a specific structure: variables are stored in columns, observations in rows, and a single type of experimental unit per dataset.
+Tidy datasets are easy to manipulate, model and visualise, and have a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table.
 This framework makes it easy to tidy messy datasets because only a small set of tools are needed to deal with a wide range of un-tidy datasets.
@@
   Chief Scientist, RStudio\\
-  Adjunct Assitant Professor, Rice University\\
+  Adjunct Professor, Rice University\\
   E-mail: \email{h.wickham@gmail.com}\\
@@
   \item Each observation forms a row,
-  \item Each table stores information on one type of observational unit.
+  \item Each type of observational unit forms a table.
 \end{enumerate}
@@
 This dataset explores the relationship between income and religion in the US, and comes from a report\footnote{\url{http://religions.pewforum.org/pdf/comparison-Income\%20Distribution\%20of\%20Religious\%20Traditions.pdf}} produced by the Pew Research Center.
-The Pew Center is an American think-tank that collects data on attitudes to topics from religion to the internet, and produces many reports that contain datasets in the format shown.
+The Pew Center is an American think-tank that collects data on attitudes to topics ranging from religion to the internet, and produces many reports that contain datasets in the format shown.
 
@@
 This makes ``wide'' datasets ``long'' or ``tall'', but I will avoid those terms because they are imprecise.
-Melting is parameterised by a list of columns that are already variables, or \emph{colvars} for short.
+Melting is parameterised by a list of columns that are already variables, or \textbf{colvar}s for short.
 The other columns are converted into two variables: a new ``column'' variable that contains repeated column headings and a ``value'' variable that contains a vector of data values concatenated from the previously separate columns.
@@
   \input{data/pew-clean.tex}
-  \caption{The first ten rows of the tidied Pew survey dataset on income and religion. The variable ``column'' has been renamed to income, and ``value'' to freq.}
+  \caption{The first ten rows of the tidied Pew survey dataset on income and religion. The variable \code{column} has been renamed to income, and \code{value} to freq.}
   \label{tbl:pew-clean}
@@
 Another common use of this data format is to record regularly spaced observations over time.
-For example, the billboard dataset shown in Table~\ref{tbl:billboard-raw} records the date a song first entered the billboard top 100.
+For example, the billboard dataset shown in Table~\ref{tbl:billboard-raw} records the date a song first entered the Billboard top 100.
 It has variables artist, track, date.entered, rank and week.
@@
   \input{data/billboard-raw.tex}
-  \caption{The first eight billboard top hits for 2000. Other columns not shown are \code{wk4}, \code{wk5}, ..., \code{wk75}.}
+  \caption{The first eight Billboard top hits for 2000. Other columns not shown are \code{wk4}, \code{wk5}, ..., \code{wk75}.}
   \label{tbl:billboard-raw}
@@
 Melting yields Table~\ref{tbl:billboard-clean}.
-I have also done a little cleaning as well as tidying: the column variable has been converted to a week variable by extracting the week number, and I have added a date variable that combines the date the song entered the billboard with the week number.
+I have also done a little cleaning as well as tidying: the column variable has been converted to a week variable by extracting the week number, and I have added a date variable that combines the date the song entered the Billboard top 100 with the week number.
 
@@
 
-Tidy visualisation tools need only to be input-tidy as their output is visual.
+Tidy visualisation tools only need to be input-tidy as their output is visual.
 Domain specific languages work particularly well for the visualisation of tidy datasets because they can describe a visualisation as a mapping between variables in the dataset and aesthetic properties of the graphic like position, size, shape and colour.
@@
 
-  \item \proglang{SPSS} (\code{glm}): \code{y BY a b c d / DESIGN a b c d c*d}
+  \item \proglang{SPSS} (\code{glm}): \code{y BY a b c d / DESIGN a b c d c * d}
+
+  \item \proglang{Stata} (\code{regress}): \code{y a b c \# d}
 
@@
 
-Next we work out overall proportion of people dying in each hour:
+Next, we work out the overall proportion of people dying in each hour:
 
@@
 
-Then finally join the overall dataset with the individual dataset to make it easier to compare the two:
+Then finally, we join the overall dataset with the individual dataset to make it easier to compare the two:
 
@@
 We break the diseases into two plots because of the differences in variability: the top plot shows diseases with over 350 deaths and the bottom with less than 350.
-The causes of death fall in to three main groups: murders, drowning, and transportation related.
+The causes of death fall into three main groups: murders, drowning, and transportation related.
 Murders are more common at night, drowning in the afternoon, and transportation related during commute times.
@@
 This paper carves out a small but important subset of data cleaning that I've called data tidying: getting the dataset in the right structure to make further manipulation, visualisation and modelling easy.
-There is still much work to be done: as well as incremental improvements as my understanding of tidy data and tidy tools improves, there are big improvements possible from exploring alternative formulations of tidy data and tackling other data cleaning problems.
+There is still much work to be done.
+Incremental improvements will happen as my understanding of tidy data and tidy tools improves, reducing the friction of getting data into a tidy form, and then manipulating, modelling and visualising it once it's tidy.
 
+Bigger improvements may be possible by exploring alternative formulations of tidiness.
 There is a chicken-and-egg problem with tidy data: if tidy data is only as useful as the tools that work with it, then tidy tools are inextricably linked to tidy data.
-This makes it is easy to get stuck in a local maxima where independently changing data structures or data tools does not result in an improved workflow.
+This makes it easy to get stuck in a local maxima where independently changing data structures or data tools does not result in an improved workflow.
 Breaking out of this local maxima is hard, and requires long-term concerted effort with many false starts.
@@
 The human factors, user-centred design, and human-computer interaction communities may be able to add to this conversation, but the design of data and tools to work with it has not been an active research topic.
-In the future, I hope to apply methodologies from these fields (user-testing, ethnography, talk-aloud protocols) to my work to build a better understanding of the cognitive side of data analysis, and how we can best design tools to support the process.
-
-% My definition of tidy data is not exhaustive, and it focusses on the type of data that statisticians most commonly encounter: rectangular data defined by rows and columns.
-It is not a good fit for data on networks, or entire texts.
-Additionally, because the focus of this definition is on explicit definition of the data, it may not be suitable for very large datasets, such as microarray or fMRI, which may require special infrastructure for performance.
+In the future, I hope to apply methodologies from these fields (user-testing, ethnography, talk-aloud protocols) to my work to build a better understanding of the cognitive side of data analysis, and a better knowledge of how we can design tools to support the process.
 
@@
 
-Apart from tidying, there are many other tasks involved in cleaning data: parsing dates and numbers, identifying missing values, correcting character encodings (for international data), matching similar but not identical values (created by typos), verifying experimental design, filling in structural missing values, not to mention model-based data cleaning that identifies suspicious values.
+Apart from tidying, there are many other tasks involved in cleaning data: parsing dates and numbers, identifying missing values, correcting character encodings (for international data), matching similar but not identical values (created by typos), verifying experimental design, and filling in structural missing values, not to mention model-based data cleaning that identifies suspicious values.
 Can we develop other frameworks to make these tasks easier?

### 2014-01-08 case-study/case-study.r
Add source & better compression for case study

@@ -6,15 +6,23 @@ library(stringr)
 library(MASS)
 source("../data/xtable.r")
 
-# Load data and do basic cleaning for this task
-if (!exists("deaths", inherit = FALSE)) {
-  load("deaths.rdata")
+if (!file.exists("deaths.rds")) {
+  src <- "https://github.com/hadley/mexico-mortality/raw/master/deaths/deaths08.csv.bz2"
+  file.download(src, "deaths.csv.bz2", quiet = TRUE)
+  
+  deaths <- read.csv("deaths08.csv.bz2")
+  unlink("deaths08.csv.bz2")
   deaths$hod[deaths$hod == 99] <- NA
   deaths$hod[deaths$hod == 24] <- 0
   deaths$hod[deaths$hod == 0] <- NA
   deaths$hod <- as.integer(deaths$hod)  
   deaths <- arrange(deaths, yod, mod, dod, hod, cod)
+  deaths <- deaths[c("yod", "mod", "dod", "hod", "cod")]
+  
+  saveRDS(deaths, "deaths.rds")
 }
+
+deaths <- readRDS("deaths.rds")
 
 ok <- subset(deaths, yod == 2008 & mod != 0 & dod != 0)
 xtable(ok[c(1, 1:14 * 2000), c("yod", "mod", "dod", "hod", "cod")], 

### 2014-01-08 reviews/response.md
Minor problems from reviewer 3

@@
 
+Removed.
+
 > Section 3.1: Since most readers of this paper will want to apply these ideas 
@@
 
+I think including the code in the paper would make it too long, but they
+are all available at the paper website at https://github.com/hadley/tidy-data
+
 > Section 3.5: I feel the examples in this section need to be explained in 
@@
 > section. (R3)
+
+I've added a brief snippet of R code to illustrate how you might tackle this.
 
@@
 
+Right, and I've now removed this sentence.
+
 > Section 5: I was not able to download the data for the case study from 
@@
 > I recieved a 403 Forbidden error. (R3)
+
+Smaller version that you can download now at 
+https://raw.github.com/hadley/tidy-data/master/case-study/deaths.rds
 

### 2014-01-08 tidy-data.tex
Minor problems from reviewer 3

@@
 
+The \pkg{plyr} package makes this straightforward in \proglang{R}.
+The following code generates a vector of file names in a directory (\code{data/}), matching a regular exrepssion (ends in \code{.csv}).
+Next we name each element of the vector with the name of the file.
+We do this because \pkg{plyr} will preserve the names in the next step, ensuring that each row in the final data frame is labelled with its source.
+Finally, \code{ldply()} loops over each path, reading in the csv file and combining the results into a single data frame.
+
+\begin{Verbatim}
+paths <- dir("data", pattern = "\\.csv$", full.names = TRUE)
+names(paths) <- basename(paths)
+ldply(paths, read.csv, stringsAsFactors = FALSE)
+\end{Verbatim}
+
 Once you have a single table, you can perform additional tidying as needed.

### 2014-01-08 tidy-data.tex
Consistently label defs and vars

@@
 \Abstract{
-A huge amount of effort is spent cleaning data to get it ready for data analysis, but there has been little research on how to make data cleaning as easy and effective as possible.
-This paper tackles a small, but important, component of data cleaning: data ``tidying''.
+A huge amount of effort is spent cleaning data to get it ready for analysis, but there has been little research on how to make data cleaning as easy and effective as possible.
+This paper tackles a small, but important, component of data cleaning: data tidying.
 Tidy datasets are easy to manipulate, model and visualise, and have a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table.
@@
 
-It's often said that 80\% of the effort in a data analysis is spent on data cleaning \citep{dasu:2003}, the process of getting the data ready to analyse.
+It's often said that 80\% of the effort in a data analysis is spent on data cleaning \citep{dasu:2003}, the process of getting the dataset ready to analyse.
 Data cleaning is not only a vital first step, but it is often repeated multiple times over the course of an analysis as new problems come to light.
@@
 Part of the challenge is the breadth of activities that cleaning encompasses, from outlier checking, to date parsing, to missing value imputation.
-To get a handle on the problem, this paper focusses on a small, but important, subset of data cleaning that I call data ``tidying'': structuring data values to make the process of data analysis as easy as possible.
+To get a handle on the problem, this paper focusses on a small, but important, subset of data cleaning that I call data \textbf{tidying}: structuring a dataset to make the process of data analysis as easy as possible.
 
@@
 
-The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} has three variables (name, treatment, result) and six observations (two for each person).
-It has two values of the treatment variable (a and b), three values of the treatment variable, and either five or six values of the result variable, depending on how you think about the missing value.
+The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} has three variables (\code{person}, \code{treatment}, \code{result}) and six observations (two for each person).
+It has two values of the \code{treatment} variable (a and b), three values of the \code{person} variable, and either five or six values of the \code{result} variable, depending on how you think about the missing value.
 We can make this structure more obvious by putting the variables in the columns and the observations in the rows, as in Table~\ref{tbl:preg-tidy}.
@@
   \input{data/preg-tidy.tex}
-  \caption{Tidied data.}
+  \caption{The same data as in Table~\ref{tbl:preg-raw-1} but with variables in columns and observations in rows.}
   \label{tbl:preg-tidy}
@@
 For a given dataset, it's usually easy to figure out what are observations and what are variables, but precisely defining variables and observations in general is surprisingly difficult.
-For example, if the columns in the Table~\ref{tbl:preg-raw-1} were height and weight we would have been happy to call them variables.
-If the columns were height and width, it would be less clear cut, as we might think of height and width as values of a dimension variable.
-If the columns were home phone and work phone, we could treat these as two variables, but in a fraud detection environment we might want variables phone number and number type because the use of one phone number for multiple people might suggest fraud.
-A general rule of thumb is that it is easier to describe functional relationships between variables (e.g., z is a linear combination of x and y, density is the ratio of weight to volume) than between rows, and it is easier to make comparisons between groups of observations (e.g., average of group a vs.\ average of group b) than between columns.
+For example, if the columns in the Table~\ref{tbl:preg-raw-1} were \code{height} and \code{weight} we would have been happy to call them variables.
+If the columns were
+\code{height} and \code{width}, it would be less clear cut, as we might think of height and width as values of a \code{dimension} variable.
+If the columns were \code{home phone} and \code{work phone}, we could treat these as two variables, but in a fraud detection environment we might want variables \code{phone number} and \code{number type} because the use of one phone number for multiple people might suggest fraud.
+A general rule of thumb is that it is easier to describe functional relationships between variables (e.g., \code{z} is a linear combination of \code{x} and \code{y}, \code{density} is the ratio of \code{weight} to \code{volume}) than between rows, and it is easier to make comparisons between groups of observations (e.g., average of group a vs.\ average of group b) than between columns.
 
-Note that in a given analysis, there may be multiple types of observational unit.
-For example, in a trial of new allergy medication we might have three observational types: demographic data collected from each person (age, sex, race), medical data collected from each person on each day (number of sneezes, redness of eyes), and meterological data collected on each day (temperature, pollen count).
+In a given analysis, there may be multiple types of observational unit.
+For example, in a trial of new allergy medication we might have three observational types: demographic data collected from each person (\code{age}, \code{sex}, \code{race}), medical data collected from each person on each day (\code{number of sneezes}, \code{redness of eyes}), and meterological data collected on each day (\code{temperature}, \code{pollen count}).
 
@@
 
-Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw-1} because each row represents an observation, the result of one treatment on one person, and each column is a variable.
+Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw-1} because each row represents an observation, the \code{result} of one \code{treatment} on one \code{person}, and each column is a variable.
 We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural: impossible because of the design of the experiment.
@@
 
-Surprisingly, most messy datasets, including types of messiness not explicitly described above, can be tidied with a small set of tools: ``melting'', string splitting, and ``casting''.
+Surprisingly, most messy datasets, including types of messiness not explicitly described above, can be tidied with a small set of tools: melting, string splitting, and casting.
 The following sections illustrate each problem with a real dataset that I have encountered in practice, and show how to tidy it.
@@
 
-This dataset has three variables, religion, income and frequency, and to tidy it we need to ``melt'', or stack it, turning columns into rows.
-This makes ``wide'' datasets ``long'' or ``tall'', but I will avoid those terms because they are imprecise.
+This dataset has three variables, \code{religion}, \code{income} and \code{frequency}, and to tidy it we need to \textbf{melt}, or stack it, turning columns into rows.
+This makes wide datasets long or tall, but I will avoid those terms because they are imprecise.
 Melting is parameterised by a list of columns that are already variables, or \textbf{colvar}s for short.
-The other columns are converted into two variables: a new ``column'' variable that contains repeated column headings and a ``value'' variable that contains a vector of data values concatenated from the previously separate columns.
+The other columns are converted into two variables: a new variable called \code{column} that contains repeated column headings and a new variable called \code{value} that contains a vector of data values concatenated from the previously separate columns.
 This is illustrated in Table~\ref{tbl:melt} with a toy dataset.
-The result of melting is a ``molten'' dataset.
+The result of melting is a \textbf{molten} dataset.
 
@@
 
-The Pew dataset has one colvar, religion, and melting yields Table~\ref{tbl:pew-clean}.
-The column variable has been renamed to \code{income}, and the value column to \code{freq}, to better reflect their roles in this dataset.
-This form is tidy: each column represents a variable and each row represents an observation, in this case a demographic unit corresponding to a combination of religion and income.
+The Pew dataset has one colvar, \code{religion}, and melting yields Table~\ref{tbl:pew-clean}.
+The \code{variable} column has been renamed to \code{income}, and the \code{value} column to \code{freq}, to better reflect their roles in this dataset.
+This form is tidy: each column represents a variable and each row represents an observation, in this case a demographic unit corresponding to a combination of \code{religion} and \code{income}.
 
@@
   \input{data/pew-clean.tex}
-  \caption{The first ten rows of the tidied Pew survey dataset on income and religion. The variable \code{column} has been renamed to income, and \code{value} to freq.}
+  \caption{The first ten rows of the tidied Pew survey dataset on income and religion. The  \code{column} has been renamed to \code{income}, and \code{value} to \code{freq}.}
   \label{tbl:pew-clean}
@@
 For example, the billboard dataset shown in Table~\ref{tbl:billboard-raw} records the date a song first entered the Billboard top 100.
-It has variables artist, track, date.entered, rank and week.
+It has variables \code{artist}, \code{track}, \code{date.entered}, \code{rank} and \code{week}.
 The rank in each week after it enters the top 100 is recorded in 75 columns, \code{wk1} to \code{wk75}.
@@
 
-This dataset has colvars year, artist, track, time, and date.entered.
+This dataset has colvars \code{year}, \code{artist}, \code{track}, \code{time}, and \code{date.entered}.
 Melting yields Table~\ref{tbl:billboard-clean}.
-I have also done a little cleaning as well as tidying: the column variable has been converted to a week variable by extracting the week number, and I have added a date variable that combines the date the song entered the Billboard top 100 with the week number.
+I have also done a little cleaning as well as tidying: the \code{column} variable has been converted to a \code{week} variable by extracting the number, and I have added a \code{date} variable computed from \code{date.entered} and \code{week}.
 
@@
 
-After melting, it often happens that the ``column'' variable is a combination of multiple underlying variables.
+After melting, it often happens that the \code{column} variable is a combination of multiple underlying variables.
 This is illustrated by the tuberculosis (TB) dataset, a sample of which is shown in Table~\ref{tbl:tb-raw}.
-This dataset comes from the World Health Organisation, and records the counts of confirmed tuberculosis cases by country, year, and demographic group.
-The demographic groups are broken down by sex (m, f) and age (0--14, 15--25, 25--34, 35--44, 45--54, 55--64, unknown).
+This dataset comes from the World Health Organisation, and records the counts of confirmed tuberculosis cases by \code{country}, \code{year}, and demographic group.
+The demographic groups are broken down by \code{sex} (m, f) and \code{age} (0--14, 15--25, 25--34, 35--44, 45--54, 55--64, unknown).
 
@@
 
-Table~\subref*{tbl:tb-molten} shows the results of melting the TB dataset, and Table~\subref*{tbl:tb-tidy} the results of splitting the column ``column'' into two real variables: age and sex.
+Table~\subref*{tbl:tb-molten} shows the results of melting the TB dataset, and Table~\subref*{tbl:tb-tidy} the results of splitting the column \code{column} into two real variables: \code{age} and \code{sex}.
 
@@
 
-  \caption{Tidying the TB dataset requires first melting, and then splitting the ``column'' column into two variables: sex and age.}
+  \caption{Tidying the TB dataset requires first melting, and then splitting the \code{column} column into two variables: \code{sex} and \code{age}.}
   \label{tbl:tb-clean}
@@
 Table~\ref{tbl:weather-raw} shows daily weather data from the Global Historical Climatology Network for one weather station (MX17004) in Mexico for five months in 2010.
-It has variables in individual columns (id, year, month), spread across columns (day, d1--d31) and across rows (tmin, tmax) (minimum and maximum temperature).
+It has variables in individual columns (\code{id}, \code{year}, \code{month}), spread across columns (\code{day}, d1--d31) and across rows (\code{tmin}, \code{tmax}) (minimum and maximum temperature).
 Months with less than 31 days have structural missing values for the last day(s) of the month.
-The element column is not a variable: it stores the names of variables.
+The \code{element} column is not a variable: it stores the names of variables.
 
-To tidy this dataset we first melt it with colvars id, year, month and the column that contains variable names, element, yielding Table~\subref*{tbl:weather-molten}.
+To tidy this dataset we first melt it with colvars \code{id}, \code{year}, \code{month} and the column that contains variable names, \code{element}, yielding Table~\subref*{tbl:weather-molten}.
 For presentation, we have dropped the missing values, making them implicit rather than explicit.
@@
 This dataset is mostly tidy, but we have two variables stored in rows: \code{tmin} and \code{tmax}, the type of observation.
-Not shown in this example are other meteorological variables prcp (precipitation) and snow (snowfall).
+Not shown in this example are other meteorological variables \code{prcp} (precipitation) and \code{snow} (snowfall).
 Fixing this requires the cast, or unstack, operation, which performs the inverse of melting, rotating the \code{element} variable back out into the columns to give Table~\subref*{tbl:weather-tidy}.
@@
 The billboard dataset described in Table~\ref{tbl:billboard-clean} actually contains observations on two types of observational unit: the song, and its rank in each week.
-This is revealed through the duplication of facts about the song: the artist and time are repeated for every song in each week.
-The billboard dataset needs to be broken down into two datasets: a song dataset which stores artist, song name and time; and a ranking dataset which gives the rank of the song in each week.
+This is revealed through the duplication of facts about the song: the \code{artist} and \code{time} are repeated for every song in each week.
+The billboard dataset needs to be broken down into two datasets: a song dataset which stores \code{artist}, \code{song name} and \code{time}; and a ranking dataset which gives the \code{rank} of the \code{song} in each \code{week}.
 Table~\ref{tbl:billboard-normal} shows these two datasets.
-You could also imagine a week dataset which would record background information about the week, maybe the total number of songs sold or similar ``demographic'' information.
+You could also imagine a week dataset which would record background information about the week, maybe the total number of songs sold or similar demographic information.
 
@@
 
-In \proglang{R}, filtering and transforming are performed by the base \proglang{R} functions \code{subset} and \code{transform}.
+In \proglang{R}, filtering and transforming are performed by the base \proglang{R} functions \code{subset()} and \code{transform()}.
 These are input and output-tidy.
-The \code{aggregate} function performs group-wise aggregation, is input-tidy, and is output-tidy providing a single aggregation method is used.
-The \pkg{plyr} package provides a tidy \code{summarise} and \code{arrange} function for aggregation and sorting.
+The \code{aggregate()} function performs group-wise aggregation, is input-tidy, and is output-tidy providing a single aggregation method is used.
+The \pkg{plyr} package provides tidy \code{summarise()} and \code{arrange()} functions for aggregation and sorting.
 
@@
 Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy.
-Base \proglang{R} possesses a \code{by} function, which is input-tidy, but not output-tidy, because it produces a list.
-The \code{ddply} function from the \pkg{plyr} package is a tidy alternative.
+Base \proglang{R} possesses a \code{by()} function, which is input-tidy, but not output-tidy, because it produces a list.
+The \code{ddply()} function from the \pkg{plyr} package is a tidy alternative.
 
@@
 An advantage of tidy data is the ease with which it can be combined with other tidy datasets: we just need a join operator which works by matching common variables and adding new columns.
-This is implemented in the \code{merge} function in base \proglang{R}, or the \code{join} function in \pkg{plyr}.
+This is implemented in the \code{merge()} function in base \proglang{R}, or the \code{join()} function in \pkg{plyr}.
 Compare this to the difficulty of combining datasets stored in arrays; these typically require painstaking alignment before matrix operations can be used, and errors can be very hard to detect.
@@
 
-Most graphical tools in \proglang{R} are input tidy, including the \pkg{base} \code{plot} function, the \pkg{lattice} family of plots \citep{sarkar:2008} and \pkg{ggplot2} \citep{me:ggplot2}.
+Most graphical tools in \proglang{R} are input tidy, including the \pkg{base} \code{plot()} function, the \pkg{lattice} family of plots \citep{sarkar:2008} and \pkg{ggplot2} \citep{me:ggplot2}.
 Some specialised tools exist for visualising messy datasets.
-Some base \proglang{R} functions like \code{barplot}, \code{matplot}, \code{dotchart}, and \code{mosaicplot}, work with messy datasets where variables are spread out over multiple columns.
+Some base \proglang{R} functions like \code{barplot()}, \code{matplot()}, \code{dotchart()}, and \code{mosaicplot()}, work with messy datasets where variables are spread out over multiple columns.
 Similarly, the parallel coordinates plot \citep{wegman:1990,inselberg:1985} can be used to create time series plots for messy datasets where each time point is a column.
@@
 
-  \item \proglang{R} (\code{lm}): \code{y ~ a + b + c * d}
+  \item \proglang{R} (\code{lm()}): \code{y ~ a + b + c * d}
 
@@
 If the data is stored as in Table~\subref*{tbl:paired}, then a paired t-test is just a t-test applied to the mean difference between x and y.
-If it is stored in the form of Table~\subref*{tbl:mixed}, then an equivalent model is to fit a mixed effects model, with a fixed dummy variable representing the ``variable'', and a random intercept for each id.
+If it is stored in the form of Table~\subref*{tbl:mixed}, then an equivalent model is to fit a mixed effects model, with a fixed dummy variable representing the \code{variable}, and a random intercept for each id.
 (In \proglang{R}'s lmer4 notation, this is \code{value ~ variable + (1 | id)}).
@@
 The full dataset has information on 539,530 deaths in Mexico in 2008 and 55 variables, including the the location and time of death, the cause of death, and demographics of the deceased.
-Table~\ref{fig:raw} shows a small sample of the dataset, focussing on variables related to time of death (year, month, day and hour), and cause (cod).
+Table~\ref{fig:raw} shows a small sample of the dataset, focussing on variables related to time of death (\code{year}, \code{month}, \code{day} and \code{hour}), and cause (\code{cod}).
 
@@
 
-Next we join the dataset to \code{codes} dataset that gives informative labels for the disease.
+Next we join the dataset to the \code{codes} dataset which gives informative labels for the disease.
 These two datasets are connected by the \code{cod} variable.
@@
 This means that it makes more sense to compare the proportion of deaths in each hour, rather than the total number.
-We compute this by breaking the dataset down by \code{cod}, and then \code{transform}ing to add a new \code{prop} column, the hourly frequency divided by the total number of deaths from that cause.
-This new columns is Table~\subref*{tbl:counts:3}.
+We compute this by breaking the dataset down by \code{cod}, and then \code{transform()}ing to add a new \code{prop} column, the hourly frequency divided by the total number of deaths from that cause.
+This new column is Table~\subref*{tbl:counts:3}.
 
-\code{ddply} breaks down its first argument (\code{hod2}) by its second argument (the \code{cod} variable), and applies its third argument (\code{transform}) to each piece.
+\code{ddply()} breaks down its first argument (\code{hod2}) by its second argument (the \code{cod} variable), and applies its third argument (\code{transform}) to each piece.
 The fourth argument (\code{prop = freq / sum(freq)}) is passed on to transform.
@@
 We also record the sample size, the total number of deaths from that cause.
-To ensure that the diseases we work with have a decent amount of underlying data we'll only
-work with diseases with more than 50 deaths in total ($\sim$2/hour).
+To ensure that the diseases we work with have a decent amount of underlying data we'll only work with diseases with more than 50 deaths in total ($\sim$2/hour).
 
@@
 
-We don't know the variance characteristics of this estimator, so we explore it visually by plotting n vs.\ deviation, Figure~\ref{fig:deviation-raw}.
+We don't know the variance characteristics of this estimator, so we explore it visually by plotting \code{n} vs.\ \code{deviation}, Figure~\ref{fig:deviation-raw}.
 The linear scale shows little, except the variability decreases with sample size, but on the log-log scale, Figure~\ref{fig:deviation-log} there is a clear pattern.

### 2014-01-08 tidy-data.tex
Final changes for reviewers

@@
 Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw-1} because each row represents an observation, the \code{result} of one \code{treatment} on one \code{person}, and each column is a variable.
-We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural: impossible because of the design of the experiment.
+We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural, impossible because of the design of the experiment.
 
@@
   \input{data/pew-raw.tex}
-  \caption{The first ten rows of data on income and religion from the Pew Forum. Two columns have been omitted to save space: `\$100-150k' and `\$150k'} 
+  \caption{The first ten rows of data on income and religion from the Pew Forum. Two columns, \code{\$100-150k} and \code{\$150k}, have been omitted} 
   \label{tbl:pew-raw}
@@
 The \code{variable} column has been renamed to \code{income}, and the \code{value} column to \code{freq}, to better reflect their roles in this dataset.
-This form is tidy: each column represents a variable and each row represents an observation, in this case a demographic unit corresponding to a combination of \code{religion} and \code{income}.
+This form is tidy because each column represents a variable and each row represents an observation, in this case a demographic unit corresponding to a combination of \code{religion} and \code{income}.
 
@@
 If a song is in the top 100 for less than 75 weeks the remaining columns are filled with missing values.
-This form of storage is not tidy, but is useful for data entry because it reduces duplication: otherwise each song in each week would need its own row, and song metadata like title and artist would need to be repeated.
+This form of storage is not tidy, but is useful for data entry because it reduces duplication; otherwise each song in each week would need its own row, and song metadata like title and artist would need to be repeated.
 This issue will be discussed in more depth in Section~\ref{sub:multiple-types}.
@@
 Melting yields Table~\ref{tbl:billboard-clean}.
-I have also done a little cleaning as well as tidying: the \code{column} variable has been converted to a \code{week} variable by extracting the number, and I have added a \code{date} variable computed from \code{date.entered} and \code{week}.
+I have also done a little cleaning as well as tidying: \code{column} has been converted to \code{week} by extracting the number, and \code{date} has been computed from \code{date.entered} and \code{week}.
 
@@
   \caption{Original TB dataset.
-Corresponding to each `m' column for males, there is also an `f' column for females: \code{f1524}, \code{f2534} and so on.
+Corresponding to each `m' column for males, there is also an `f' column for females, \code{f1524}, \code{f2534} and so on.
 These are not shown to conserve space.
-Note the mixture of 0s and missing values (---): this is due to the data collection process and the distinction is important for this dataset.}
+Note the mixture of 0s and missing values (---).
+This is due to the data collection process and the distinction is important for this dataset.}
   \label{tbl:tb-raw}
@@
 
-Storing the values in this form resolves another problem in the original data: we want to compare rates, not counts, but to compute rates we need to know the population.
+Storing the values in this form resolves another problem in the original data.
+We want to compare rates, not counts, but to compute rates we need to know the population.
 In the original format, there is no easy way to add a population variable, and it has to be stored in a separate table, which makes it hard to correctly match populations to counts.
-In tidy form, adding variables to store population and rate is easy: they just become additional columns.
+In tidy form, adding variables to store population and rate is easy; they just become additional columns.
 
@@
 Months with less than 31 days have structural missing values for the last day(s) of the month.
-The \code{element} column is not a variable: it stores the names of variables.
+The \code{element} column is not a variable; it stores the names of variables.
 
@@
 Fixing this requires the cast, or unstack, operation, which performs the inverse of melting, rotating the \code{element} variable back out into the columns to give Table~\subref*{tbl:weather-tidy}.
-This form is tidy: there is one variable in each column, and each row represents a day's observations.
+This form is tidy.
+There is one variable in each column, and each row represents a day's observations.
 The cast operation is described in depth in \citet{wickham:2007b}.
@@
 The billboard dataset described in Table~\ref{tbl:billboard-clean} actually contains observations on two types of observational unit: the song, and its rank in each week.
-This is revealed through the duplication of facts about the song: the \code{artist} and \code{time} are repeated for every song in each week.
+This is revealed through the duplication of facts about the song; \code{artist} and \code{time} are repeated for every song in each week.
 The billboard dataset needs to be broken down into two datasets: a song dataset which stores \code{artist}, \code{song name} and \code{time}; and a ranking dataset which gives the \code{rank} of the \code{song} in each \code{week}.
@@
 The four verbs can be, and often are, modified by the ``by'' preposition.
-We often need group-wise aggregates, transformations or subsets: pick the biggest in each group, average over replicates and so on.
+We often need group-wise aggregates, transformations and subsets, to pick the biggest in each group, oaverage over replicates and so on.
 Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time.
@@
 Other tools are needed when we have multiple datasets.
-An advantage of tidy data is the ease with which it can be combined with other tidy datasets: we just need a join operator which works by matching common variables and adding new columns.
+An advantage of tidy data is the ease with which it can be combined with other tidy datasets.
+All that is needed is a join operator that works by matching common variables and adding new columns.
 This is implemented in the \code{merge()} function in base \proglang{R}, or the \code{join()} function in \pkg{plyr}.
@@
 This makes it more difficult to combine results from multiple models.
-For example, in \proglang{R}, the default representation of model coefficients is not tidy because it does not have an explicit variable that records the variable name for each estimate: these are instead recorded as row names.
+For example, in \proglang{R}, the default representation of model coefficients is not tidy because it does not have an explicit variable that records the variable name for each estimate, they are instead recorded as row names.
 In \proglang{R}, row names must be unique, so combining coefficients from many models (e.g., from bootstrap resamples, or subgroups) requires workarounds to avoid losing important information.
 This knocks you out of the flow of analysis and makes it harder to combine the results from multiple models.
+I'm not currently aware of any packages that resolve this problem.
 
@@
 
-The following case study illustrates how tidy data and tidy tools make data analysis easier by easing the transitions between manipulation, visualisation and modelling: you will not see any code that exists solely to get the output of one function into the right format for input to another.
+The following case study illustrates how tidy data and tidy tools make data analysis easier by easing the transitions between manipulation, visualisation and modelling.
+You will not see any code that exists solely to get the output of one function into the right format for input to another.
 I'll show the \proglang{R} code that performs the analysis, but even if you're not familiar with \proglang{R} or the exact idioms I use, I've tried to make it easy to understand by tightly interleaving code, results and explanation.
@@
 
-We are interested in points that are high relative to other causes with the same number of deaths: these are diseases most different from the overall pattern.
+We are interested in points that are high relative to other causes with the same number of deaths, these are the diseases most different from the overall pattern.
 To find these unusual points, we fit a robust linear model and plot the residuals, as shown in Figure~\ref{fig:devi-resid}.

### 2014-01-10 tidy-data.tex
Introduction

@@
 
-It's often said that 80\% of the effort in a data analysis is spent on data cleaning \citep{dasu:2003}, the process of getting the dataset ready to analyse.
-Data cleaning is not only a vital first step, but it is often repeated multiple times over the course of an analysis as new problems come to light.
-Despite the amount of time it takes, there has been little research on how to clean data well.
-Part of the challenge is the breadth of activities that cleaning encompasses, from outlier checking, to date parsing, to missing value imputation.
-To get a handle on the problem, this paper focusses on a small, but important, subset of data cleaning that I call data \textbf{tidying}: structuring a dataset to make the process of data analysis as easy as possible.
+It is often said that 80\% of data analysis is spent on the process of cleaning and preparing the data \citep{dasu:2003}.
+However, since such data cleaning is often repeated over the course of analysis as new problems come to light, it is much more than just a first step.
+Despite the amount of time it takes, there has been surprisingly little research on how to clean data well.
+Part of the challenge is the breadth of activities it encompasses: from outlier checking, to date parsing, to missing value imputation.
+To get a handle on the problem, this paper focusses on a small, but important, aspect of data cleaning that I call data \textbf{tidying}: the structuring of datasets to facilitate analysis.
 
-The principles of tidy data provide a standard way of organising the data values within a dataset.
-A standard makes initial data cleaning easier because you don't need to start from scratch each time and you can avoid reinventing the wheel.
-The tidy data standard has been designed to facilitate initial exploration and analysis of the data, and it makes it easier to develop data analysis tools that work together seamlessly.
-Current tools are often inconsistent, so you need to spend time munging the output from one tool into the input for another.
-Tidy datasets and tidy tools work hand in hand to make data analysis easier, allowing you to focus on the interesting domain problem, not the uninteresting logistics of your data.
+The principles of tidy data provide a standard way of organising data values within a dataset.
+A standard makes the initial data cleaning easier because you don't need to start from scratch and reinvent the wheel each time.
+The tidy data standard has been designed to facilitate the initial exploration and analysis of the data, and to simplify the development of data analysis tools which can work together seamlessly.
+Current tools often require translation.
+You need to spend time munging the output from one tool so you can input it into another.
+Tidy datasets and tidy tools work hand in hand to make data analysis easier, allowing you to focus on the interesting domain problem, and not on the uninteresting logistics of your data.
 
-The principles of tidy data are closely tied to relational databases and Codd's relational algebra \citep{codd:1990}, but are framed in a language more familiar to statisticians.
+The principles of tidy data are closely tied to those of relational databases and Codd's relational algebra \citep{codd:1990}, but are framed in a language familiar to statisticians.
 Computer scientists have also contributed much to the study of data cleaning.
-For example, \citet{lakshmanan:1996} defines an extension to SQL allowing it to operate on messy datasets, \citet{raman:2001} provides a framework for cleaning datasets, and \citet{kandel:2011} develops an interactive tool that makes data cleaning easy, automatically creating the code to clean datasets with a friendly user interface.
-These tools are useful, but they are presented in a language foreign to most statisticians, they fail to give much advice on how datasets should be structured, and they lack connections to the tools of data analysis.
+For example, \citet{lakshmanan:1996} define an extension to SQL to allow it to operate on messy datasets, \citet{raman:2001} provide a framework for cleaning datasets, and \citet{kandel:2011} develop an interactive tool with a friendly user interface which automatically creates code to clean data.
+These tools are useful but they are presented in a language foreign to most statisticians, they fail to give much advice on how datasets should be structured, and they lack connections to the tools of data analysis.
 
-The development of tidy data has been driven by my struggles working with real-world datasets, which are often organised in bizarre ways.
-I have spent countless hours struggling to get these datasets organised in a way that makes data analysis possible, let alone easy.
-I have also struggled to impart these skills to my students, so they can tackle problems in the real-world, where there are few constraints on how datasets are organised.
-In the course of these struggles I developed the \pkg{reshape} and \pkg{reshape2} \citep{wickham:2007b} \proglang{R} packages, but while I could intuitively use the tools, I lacked the framework to make my understanding explicit.
-This paper provides that framework, and provides a consistent ``philosophy of data'' that underlies my work in the \pkg{plyr} \citep{me:plyr} and \pkg{ggplot2} \citep{me:ggplot2} packages.
+The development of tidy data has been driven by my experience working with real-world datasets.
+With few, if any, constraints on their organisation, such datasets are constructed in often bizarre ways.
+I have spent countless hours struggling to get such datasets organised in a way that makes data analysis possible, let alone easy.
+I have also struggled to impart these skills to my students so they could tackle real-world datasets on their own.
+In the course of these struggles I developed the \pkg{reshape} and \pkg{reshape2} \citep{wickham:2007b} \proglang{R} packages.
+While I could intuitively use the tools, I lacked the framework to make my intuition explicit.
+This paper provides that framework.
+It provides a comprehensive ``philosophy of data'': one that underlies my work in the \pkg{plyr} \citep{me:plyr} and \pkg{ggplot2} \citep{me:ggplot2} packages.
 
@@
 Section~\ref{sec:defining} begins by defining the three characteristics that make a dataset tidy.
-Most real world datasets are not tidy, so Section~\ref{sec:tidying} describes the operations needed to make messy datasets tidy, and illustrates the techniques with a range of real examples.
-Section~\ref{sec:tidy-tools} defines tidy tools, tools that input and output tidy datasets, and discusses how together tidy data and tidy tools make it easier to do a data analysis.
+Since most real world datasets are not tidy, Section~\ref{sec:tidying} describes the operations needed to make messy datasets tidy, and illustrates the techniques with a range of real examples.
+Section~\ref{sec:tidy-tools} defines tidy tools, tools that input and output tidy datasets, and discusses how tidy data and tidy tools together can make data analysis easier.
 These principles are illustrated with a small case study in Section~\ref{sec:case-study}.

### 2014-01-10 tidy-data.tex
Changes from Garrett

@@
 
-The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} has three variables (\code{person}, \code{treatment}, \code{result}) and six observations (two for each person).
-It has two values of the \code{treatment} variable (a and b), three values of the \code{person} variable, and either five or six values of the \code{result} variable, depending on how you think about the missing value.
-We can make this structure more obvious by putting the variables in the columns and the observations in the rows, as in Table~\ref{tbl:preg-tidy}.
+The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} contains three variables that comprise 10 values.
+The data contains a \code{person} variable with three values (John, Mary, and Jane), a \code{treatment} variable with two values (a and b), and a \code{result} variable with has five or six values depending on how you think of the missing value (-, 16, 3, 2, 11, 1).
+The layout of Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} makes the number of values clear, but obscures their meaning.
+
+The meaning of the dataset relies on how these values are related to each other.
+The \code{result} variable was collected over the course of six observations: each \code{person} was observed twice, once for each \code{treatment}.
+These observations give meaning to the specific values of result.
+For example, the value 16 is meaningful in the context of the observation: \emph{Jane} scored \emph{16} when she took treatment \code{a}.
+Table~\ref{tbl:preg-tidy} organizes the dataset around the six observations, making the meaning of the individual values more clear.
 
@@
 
-Tidy data makes it easy for an analyst or a computer to extract needed variable because it provides a standard way of structuring a dataset.
-Contrast Table~\ref{tbl:preg-tidy} to Table~\ref{tbl:preg-raw-1}: in Table~\ref{tbl:preg-raw-1} you need to use different strategies for different variables.
+Tidy data makes it easy for an analyst or a computer to extract needed variables because it provides a standard way of structuring a dataset.
+Contrast Table~\ref{tbl:preg-tidy} to Table~\ref{tbl:preg-raw-1}: in Table~\ref{tbl:preg-raw-1} you need to use different strategies to extract different variables.
 This slows analysis and invites errors.

### 2014-01-10 tidy-data.tex
2.0 and 2.1

@@
 
-Like families, tidy datasets are all alike, but every messy dataset is messy in its own way.
-Tidy datasets provide a standard way to connect the structure of a dataset (its physical layout) to its semantics (its meaning).
-I'll provide some standard vocabulary for describing the structure of a dataset, its semantics, and then use those definitions to define tidy data.
+Like families, tidy datasets are all alike but each is messy in its own way.
+Tidy datasets provide a standardized way to link the structure of a dataset (its physical layout) with its semantics (its meaning).
+In this section, I'll provide some standard vocabulary for describing the structure of a dataset, its semantics, and then use those definitions to define tidy data.
 
@@
 
-Most statistical datasets are rectangular tables made up of \textbf{columns} and \textbf{rows}.
-The columns are almost always labelled, and rows are often labelled.
+Most statistical datasets are rectangular tables made up of \textbf{rows} and \textbf{columns}.
+While columns are almost always labelled, rows are sometimes not.
 Table~\ref{tbl:preg-raw-1} provides some data about an imaginary experiment in a format commonly seen in the wild.
@@
 The data is the same, but the layout is different.
-Our vocabulary of rows and columns is not rich enough to say why these two tables represent the same data.
-We also need a way to describe the meaning of the underlying data displayed in the table.
-It's not enough to describe the appearance of a dataset (how the values are physically laid out), we also need a way to describe the underlying semantics, the meaning, of the values in the dataset.
+Our vocabulary of rows and columns is simply not rich enough to tell us why the two tables represent the same data.
+In addition to appearance, what we need is a way to describe the underlying semantics, or meaning, of the values displayed in the dataset.
 

### 2014-01-13 tidy-data.tex
Some minor tweaks

@@
 
-Like families, tidy datasets are all alike but each is messy in its own way.
+Like families, tidy datasets are all alike but every messy dataset is messy in its own way.
 Tidy datasets provide a standardized way to link the structure of a dataset (its physical layout) with its semantics (its meaning).
@@
 The data is the same, but the layout is different.
-Our vocabulary of rows and columns is simply not rich enough to tell us why the two tables represent the same data.
+Our vocabulary of rows and columns is simply not rich enough to describe why the two tables represent the same data.
 In addition to appearance, what we need is a way to describe the underlying semantics, or meaning, of the values displayed in the dataset.

### 2014-01-14 tidy-data.tex
2.2 Data Semantics Para 1

@@
 A table is a collection of data \textbf{values}, typically either numbers (if quantitative) or strings (if qualitative).
-Values are grouped into \textbf{variables}, measurements of a single quantifiable property.
-Multiple measurements made on the same observational unit form an \textbf{observation}.
+A \textbf{variable} is a set of measurements of a single quantifiable property for multiple observational units.
+An \textbf{observation} is a set of measurements of multiple quantifiable properties for a single observational unit.
 An observational unit is the smallest constituent atom of the experimental unit.
-For example, in a longitudinal medical study, the experimental unit might be the patient, but the observational unit is the patient at each time point.
-In an agricultural experiment, the experimental units might be the cages given different diets, but the observational units are the pigs whose individual weights are measured.
-This definition of observational unit leads to variables which are physical properties, like blood pressure and weight, not conditional properties like blood pressure at time 1, or weight of pig 2.
+For example, in a longitudinal medical study, the experimental unit might be the patient, but the observational unit is the patient at different points in time.
+In an agricultural experiment, the experimental unit might be the cages assigned different diets, but the observational units are the individual pigs in those cages.
+In short, while an observational unit represents a conditional property, like blood pressure at time 1 or the weight of pig 2, an experimental unit represents an unconditional property, like blood pressure and weight.
 

### 2014-01-14 tidy-data.tex
Tidy Data

--- 2014-01-13 tidy-data.tex
@@
 \begin{enumerate}
-  \item Each variable forms a column,
-  \item Each observation forms a row,
+  \item Each variable forms a column.
+  \item Each observation forms a row.
   \item Each type of observational unit forms a table.
@@
 
-\noindent This is Codd's 3rd normal form \citep{codd:1990}, but with the constraints framed in statistical language, and the focus on a single dataset, not the many connected datasets common in relational databases.
+\noindent This is Codd's 3rd normal form \citep{codd:1990}, but with the constraints framed in statistical language, and the focus put on a single dataset rather than the many connected datasets common in relational databases.
 \textbf{Messy data} is any other other arrangement of the data.
 
-Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw-1} because each row represents an observation, the \code{result} of one \code{treatment} on one \code{person}, and each column is a variable.
+Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw-1}.
+Each row represents an observation, the \code{result} of one \code{treatment} on one \code{person}, and each column is a variable.
 We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural, impossible because of the design of the experiment.
 
-Tidy data makes it easy for an analyst or a computer to extract needed variables because it provides a standard way of structuring a dataset.
+Because it provides a standard way of structuring a dataset, tidy data makes it easy for an analyst or a computer to extract needed variables.
 Contrast Table~\ref{tbl:preg-tidy} to Table~\ref{tbl:preg-raw-1}: in Table~\ref{tbl:preg-raw-1} you need to use different strategies to extract different variables.

### 2014-01-15 tidy-data.tex
3.x Tidying messy datasets

--- 2014-01-13 tidy-data.tex
@@
 
-Real datasets can violate the three precepts of tidy data in almost every way imaginable, and often do.
-While occasionally you do get a dataset that you can start analysing immediately, this is the exception, not the norm.
-This section describes the the five most common problems that make datasets messy, along with their remedies:
+Real datasets can, and often do, violate the three precepts of tidy data in almost every way imaginable.
+While occasionally you do get a dataset that you can start analysing immediately, this is the exception, not the rule.
+This section describes the five most common problems with messy datasets, along with their remedies:
 
@@
   \item variables are stored in both rows and columns
-  \item multiple types of observational unit are stored in the same table
+  \item multiple types of observational units are stored in the same table
   \item a single observational unit is stored in multiple tables
@@
 Surprisingly, most messy datasets, including types of messiness not explicitly described above, can be tidied with a small set of tools: melting, string splitting, and casting.
-The following sections illustrate each problem with a real dataset that I have encountered in practice, and show how to tidy it.
+The following sections illustrate each problem with a real dataset that I have encountered, and show how to tidy them.
 The complete datasets and the \proglang{R} code used to tidy them are available online at \url{https://github.com/hadley/tidy-data}, and in the online supplementary materials for this paper.
@@
 A common type of messy dataset is tabular data designed for presentation, where variables form both the rows and columns, and column headers are values, not variable names.
-While in this paper I call this arrangement messy, in some cases it can be extremely useful.
-It provides efficient storage for completely crossed designs, and it can provide extremely efficient computation if desired operations can be expressed as matrix operations.
-This issue is discussed in more depth in Section~\ref{sec:discussion}.
+While I would call this arrangement messy, in some cases it can be extremely useful.
+It provides efficient storage for completely crossed designs, and it can lead to extremely efficient computation if desired operations can be expressed as matrix operations.
+This issue is discussed in depth in Section~\ref{sec:discussion}.
 
 Table~\ref{tbl:pew-raw} shows a subset of a typical dataset of this form.
-This dataset explores the relationship between income and religion in the US, and comes from a report\footnote{\url{http://religions.pewforum.org/pdf/comparison-Income\%20Distribution\%20of\%20Religious\%20Traditions.pdf}} produced by the Pew Research Center.
-The Pew Center is an American think-tank that collects data on attitudes to topics ranging from religion to the internet, and produces many reports that contain datasets in the format shown.
+This dataset explores the relationship between income and religion in the US.
+It comes from a report\footnote{\url{http://religions.pewforum.org/pdf/comparison-Income\%20Distribution\%20of\%20Religious\%20Traditions.pdf}} produced by the Pew Research Center, an American think-tank that collects data on attitudes to topics ranging from religion to the internet, and produces many reports that contain datasets in this format.
 
@@
 
-This dataset has three variables, \code{religion}, \code{income} and \code{frequency}, and to tidy it we need to \textbf{melt}, or stack it, turning columns into rows.
-This makes wide datasets long or tall, but I will avoid those terms because they are imprecise.
+This dataset has three variables, \code{religion}, \code{income} and \code{frequency}.
+To tidy it, we need to \textbf{melt}, or stack it.
+In other words, we need to turn columns into rows.
+While this is often described as making wide datasets long or tall, I will avoid those terms because they are imprecise.
 Melting is parameterised by a list of columns that are already variables, or \textbf{colvar}s for short.
@@
 The Pew dataset has one colvar, \code{religion}, and melting yields Table~\ref{tbl:pew-clean}.
-The \code{variable} column has been renamed to \code{income}, and the \code{value} column to \code{freq}, to better reflect their roles in this dataset.
+To better reflect their roles in this dataset, the \code{variable} column has been renamed to \code{income}, and the \code{value} column to \code{freq}.
 This form is tidy because each column represents a variable and each row represents an observation, in this case a demographic unit corresponding to a combination of \code{religion} and \code{income}.
@@
 Another common use of this data format is to record regularly spaced observations over time.
-For example, the billboard dataset shown in Table~\ref{tbl:billboard-raw} records the date a song first entered the Billboard top 100.
-It has variables \code{artist}, \code{track}, \code{date.entered}, \code{rank} and \code{week}.
+For example, the Billboard dataset shown in Table~\ref{tbl:billboard-raw} records the date a song first entered the Billboard Top 100.
+It has variables for \code{artist}, \code{track}, \code{date.entered}, \code{rank} and \code{week}.
 The rank in each week after it enters the top 100 is recorded in 75 columns, \code{wk1} to \code{wk75}.
-If a song is in the top 100 for less than 75 weeks the remaining columns are filled with missing values.
-This form of storage is not tidy, but is useful for data entry because it reduces duplication; otherwise each song in each week would need its own row, and song metadata like title and artist would need to be repeated.
+If a song is in the Top 100 for less than 75 weeks the remaining columns are filled with missing values.
+This form of storage is not tidy, but it is useful for data entry.
+It reduces duplication since otherwise each song in each week would need its own row, and song metadata like title and artist would need to be repeated.
 This issue will be discussed in more depth in Section~\ref{sub:multiple-types}.
@@
 
-After melting, it often happens that the \code{column} variable is a combination of multiple underlying variables.
+After melting, the \code{column} variable often becomes a combination of multiple underlying variables.
 This is illustrated by the tuberculosis (TB) dataset, a sample of which is shown in Table~\ref{tbl:tb-raw}.
@@
 
-Column headers in this format are often separated by some character (\code{.}, \code{-}, \code{\_}, \code{:}) and the string can be broken into pieces using that character as a divider.
-In other cases, such as for this dataset, more careful string processing is required, or the variable names need to be matched to a lookup table that converts single compound values to values of multiple variables.
+Column headers in this format are often separated by some character (\code{.}, \code{-}, \code{\_}, \code{:}).
+While the string can be broken into pieces using that character as a divider, in other cases, such as for this dataset, more careful string processing is required.
+For example, the variable names can be matched to a lookup table that converts single compound value into multiple component values.
 
-Table~\subref*{tbl:tb-molten} shows the results of melting the TB dataset, and Table~\subref*{tbl:tb-tidy} the results of splitting the column \code{column} into two real variables: \code{age} and \code{sex}.
+Table~\subref*{tbl:tb-molten} shows the results of melting the TB dataset, and Table~\subref*{tbl:tb-tidy} shows the results of splitting the column \code{column} into two real variables: \code{age} and \code{sex}.
 
@@
 Storing the values in this form resolves another problem in the original data.
-We want to compare rates, not counts, but to compute rates we need to know the population.
-In the original format, there is no easy way to add a population variable, and it has to be stored in a separate table, which makes it hard to correctly match populations to counts.
-In tidy form, adding variables to store population and rate is easy; they just become additional columns.
+We want to compare rates, not counts.
+But to compute rates, we need to know the population.
+In the original format, there is no easy way to add a population variable.
+It has to be stored in a separate table, which makes it hard to correctly match populations to counts.
+In tidy form, adding variables for population and rate is easy: they just become additional columns.
 
@@
 
-The most complicated form of messy data is when variables have been stored in both rows and columns.
+The most complicated form of messy data occurs when variables are stored in both rows and columns.
 Table~\ref{tbl:weather-raw} shows daily weather data from the Global Historical Climatology Network for one weather station (MX17004) in Mexico for five months in 2010.
@@
 
-To tidy this dataset we first melt it with colvars \code{id}, \code{year}, \code{month} and the column that contains variable names, \code{element}, yielding Table~\subref*{tbl:weather-molten}.
+To tidy this dataset we first melt it with colvars \code{id}, \code{year}, \code{month} and the column that contains variable names, \code{element}.
+This yields Table~\subref*{tbl:weather-molten}.
 For presentation, we have dropped the missing values, making them implicit rather than explicit.
@@
 This dataset is mostly tidy, but we have two variables stored in rows: \code{tmin} and \code{tmax}, the type of observation.
-Not shown in this example are other meteorological variables \code{prcp} (precipitation) and \code{snow} (snowfall).
-Fixing this requires the cast, or unstack, operation, which performs the inverse of melting, rotating the \code{element} variable back out into the columns to give Table~\subref*{tbl:weather-tidy}.
+Not shown in this example are the other meteorological variables \code{prcp} (precipitation) and \code{snow} (snowfall).
+Fixing this requires the cast, or unstack, operation.
+This performs the inverse of melting by rotating the \code{element} variable back out into the columns (Table~\subref*{tbl:weather-tidy}).
 This form is tidy.
@@
   \caption{(a) Molten weather dataset.
-This is almost tidy, but the \code{element} column contains names of variables, not values.
+This is almost tidy, but instead of value, the \code{element} column contains names of variables.
 Missing values are dropped to conserve space.
@@
 
-Datasets often involve values collected at multiple levels, on different types of observational unit.
+Datasets often involve values collected at multiple levels, on different types of observational units.
 During tidying, each type of observational unit should be stored in its own table.
-This is closely related to the idea of database normalisation, where each fact is expressed in only one place; if not, it's possible for inconsistencies to occur.
+This is closely related to the idea of database normalisation, where each fact is expressed in only one place.
+If this is not done, it's possible for inconsistencies to occur.
 
@@
 
-The billboard dataset described in Table~\ref{tbl:billboard-clean} actually contains observations on two types of observational unit: the song, and its rank in each week.
-This is revealed through the duplication of facts about the song; \code{artist} and \code{time} are repeated for every song in each week.
-The billboard dataset needs to be broken down into two datasets: a song dataset which stores \code{artist}, \code{song name} and \code{time}; and a ranking dataset which gives the \code{rank} of the \code{song} in each \code{week}.
+The Billboard dataset described in Table~\ref{tbl:billboard-clean} actually contains observations on two types of observational units: the song and its rank in each week.
+This manifests itself through the duplication of facts about the song: \code{artist} and \code{time} are repeated for every song in each week.
+The billboard dataset needs to be broken down into two datasets: a song dataset which stores \code{artist}, \code{song name} and \code{time}, and a ranking dataset which gives the \code{rank} of the \code{song} in each \code{week}.
 Table~\ref{tbl:billboard-normal} shows these two datasets.
@@
 
-Normalisation is useful for tidying and eliminating inconsistencies, but there are few data analysis tools that work directly with relational data, so analysis usually requires denormalisation, merging the datasets back into one table.
+Normalisation is useful for tidying and eliminating inconsistencies.
+However, there are few data analysis tools that work directly with relational data, so analysis usually also requires denormalisation or the merging the datasets back into one table.
 
@@
 
-It's also common to find data values about a single type of observational unit spread out over multiple tables, or multiple files.
-These tables are often split up by another variable, so each file represents a single year, person, or location.
+It's also common to find data values about a single type of observational unit spread out over multiple tables or files.
+These tables are often split up by another variable, so that each file represents a single year, person, or location.
 As long as the format for individual records is consistent, this is an easy problem to fix:
@@
 \begin{enumerate}
-  \item read the files into a list of tables,
+  \item Read the files into a list of tables.
 
-\item for each table, add a new column that records the original file name (because the file name is often the value of an important variable), then
+\item For each table, add a new column that records the original file name (because the file name is often the value of an important variable).
 
-  \item combine all tables into a single table. 
+  \item Combine all tables into a single table. 
 \end{enumerate}
 
-The \pkg{plyr} package makes this straightforward in \proglang{R}.
-The following code generates a vector of file names in a directory (\code{data/}), matching a regular exrepssion (ends in \code{.csv}).
+The \pkg{plyr} package makes this a straightforward task in \proglang{R}.
+The following code generates a vector of file names in a directory (\code{data/}) which match a regular expression (ends in \code{.csv}).
 Next we name each element of the vector with the name of the file.
-We do this because \pkg{plyr} will preserve the names in the next step, ensuring that each row in the final data frame is labelled with its source.
+We do this because \pkg{plyr} will preserve the names in the following step, ensuring that each row in the final data frame is labelled with its source.
 Finally, \code{ldply()} loops over each path, reading in the csv file and combining the results into a single data frame.
@@
 
-A more complicated situation occurs when the dataset structure has changed over time.
-For example, the datasets contain different variables, the same variables have different names, different file formats are used, or there are different conventions for missing values.
-This may require each file to tidied individually (or, if you're lucky, in small groups) and then combined once tidied.
+A more complicated situation occurs when the dataset structure changes over time.
+For example, the datasets may contain different variables, the same variables with different names, different file formats, or different conventions for missing values.
+This may require you to tidy each file to individually (or, if you're lucky, in small groups) and then combine them once tidied.
 An example of this type of tidying is illustrated in \url{https://github.com/hadley/data-fuel-economy}, which shows the tidying of {\sc epa} fuel economy data for over 50,000 cars from 1978 to 2008.

### 2014-01-15 tidy-data.tex
typos

@@
 
-After melting, the \code{column} variable often becomes a combination of multiple underlying variables.
+After melting, the \code{column} variable names often becomes a combination of multiple underlying variable names.
 This is illustrated by the tuberculosis (TB) dataset, a sample of which is shown in Table~\ref{tbl:tb-raw}.
@@
 It has to be stored in a separate table, which makes it hard to correctly match populations to counts.
-In tidy form, adding variables for population and rate is easy: they just become additional columns.
+In tidy form, adding variables for population and rate is easy.
+They are just additional columns.
 
@@
   \caption{(a) Molten weather dataset.
-This is almost tidy, but instead of value, the \code{element} column contains names of variables.
+This is almost tidy, but instead of values, the \code{element} column contains names of variables.
 Missing values are dropped to conserve space.
@@
 It's also common to find data values about a single type of observational unit spread out over multiple tables or files.
-These tables are often split up by another variable, so that each file represents a single year, person, or location.
+These tables and files are often split up by another variable, so that each represents a single year, person, or location.
 As long as the format for individual records is consistent, this is an easy problem to fix:

### 2014-01-15 tidy-data.tex
Tidy Tools

--- 2014-01-14 tidy-data.tex
@@
 \begin{enumerate}
-  \item Each variable forms a column.
-  \item Each observation forms a row.
+  \item Each variable forms a column,
+  \item Each observation forms a row,
   \item Each type of observational unit forms a table.
@@
 
-\noindent This is Codd's 3rd normal form \citep{codd:1990}, but with the constraints framed in statistical language, and the focus put on a single dataset rather than the many connected datasets common in relational databases.
+\noindent This is Codd's 3rd normal form \citep{codd:1990}, but with the constraints framed in statistical language, and the focus on a single dataset, not the many connected datasets common in relational databases.
 \textbf{Messy data} is any other other arrangement of the data.
 
-Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw-1}.
-Each row represents an observation, the \code{result} of one \code{treatment} on one \code{person}, and each column is a variable.
+Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw-1} because each row represents an observation, the \code{result} of one \code{treatment} on one \code{person}, and each column is a variable.
 We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural, impossible because of the design of the experiment.
 
-Because it provides a standard way of structuring a dataset, tidy data makes it easy for an analyst or a computer to extract needed variables.
+Tidy data makes it easy for an analyst or a computer to extract needed variables because it provides a standard way of structuring a dataset.
 Contrast Table~\ref{tbl:preg-tidy} to Table~\ref{tbl:preg-raw-1}: in Table~\ref{tbl:preg-raw-1} you need to use different strategies to extract different variables.
@@
 This section discusses tidy tools, tools that take tidy datasets as input and return tidy datasets as output.
-Tidy tools are useful because the output of one tool can be used as the input to another, making it straightforward to compose multiple tools to solve a problem.
+Tidy tools are useful because the output of one tool can be used as the input to another.
+This allows you to simply and easily compose multiple tools to solve a problem.
 Tidy data also ensures that variables are stored in a consistent, explicit manner.
@@
 Data manipulation includes variable-by-variable transformation (e.g., \code{log} or \code{sqrt}), as well as aggregation, filtering and reordering.
-In my experience, there are four extremely common operations that are performed over and over again in the course of an analysis.
-These are the four fundamental verbs of data manipulation:
+In my experience, these are the four fundamental verbs of data manipulation:
 
@@
 
-  \item Filtering, or subsetting, where observations are removed based on some
+  \item Filter: subsetting or removing observations based on some
   condition.
 
-  \item Transforming, where new variables are added or existing variables
-  modified. These modifications can involve either a single variable (e.g., 
-  log-transforming a variable), or involve multiple variables (e.g., computing
-  density from weight and volume).
+  \item Transform: adding or modifying variables. These modifications can
+  involve either a single variable (e.g., log-transformation), or multiple 
+  variables (e.g., computing density from weight and volume).
 
-  \item Aggregating, where multiple values are collapsed into a single value,
-  e.g., by summing or taking means.
+  \item Aggregate: collapsing multiple values into a single value (e.g., by 
+  summing or taking means).
 
-  \item Sorting, where the order of observations is changed.
+  \item Sort: changing the order of observations.
 
@@
 
-All these operations are made easier when there is a consistent way to refer to variables. Tidy data provides this because each variable lives in its own column.
+All these operations are made easier when there is a consistent way to refer to variables. Tidy data provides this because each variable resides in its own column.
 
@@
 These are input and output-tidy.
-The \code{aggregate()} function performs group-wise aggregation, is input-tidy, and is output-tidy providing a single aggregation method is used.
+The \code{aggregate()} function performs group-wise aggregation.
+It is input-tidy.
+Provided that a single aggregation method is used, it is also output-tidy .
 The \pkg{plyr} package provides tidy \code{summarise()} and \code{arrange()} functions for aggregation and sorting.
@@
 The four verbs can be, and often are, modified by the ``by'' preposition.
-We often need group-wise aggregates, transformations and subsets, to pick the biggest in each group, oaverage over replicates and so on.
-Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time.
-Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy.
-Base \proglang{R} possesses a \code{by()} function, which is input-tidy, but not output-tidy, because it produces a list.
+We often need to perform group-wise aggregation, transformation and subsetting, to pick the biggest ??
+in each group, to average over replicates and so on.
+Combining a verb with a ``by'' operator is a concise way to apply that operation to subsets of a data frame.
+Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group.
+They are generally input-tidy.
+Base \proglang{R} possesses a \code{by()} function, which is input-tidy, but, because it produces a list, is not output-tidy.
 The \code{ddply()} function from the \pkg{plyr} package is a tidy alternative.
@@
 This is implemented in the \code{merge()} function in base \proglang{R}, or the \code{join()} function in \pkg{plyr}.
-Compare this to the difficulty of combining datasets stored in arrays; these typically require painstaking alignment before matrix operations can be used, and errors can be very hard to detect.
+Compare these operators with the difficulty of combining datasets stored in arrays.
+This task typically requires painstaking alignment before matrix operations can be used, which can can make errors very hard to detect.
 
@@
 Tidy visualisation tools only need to be input-tidy as their output is visual.
-Domain specific languages work particularly well for the visualisation of tidy datasets because they can describe a visualisation as a mapping between variables in the dataset and aesthetic properties of the graphic like position, size, shape and colour.
+Domain specific languages work particularly well for the visualisation of tidy datasets because they can describe a visualisation as a mapping between variables and aesthetic properties of the graph (e.g., position, size, shape and colour).
 This is the idea behind the grammar of graphics \citep{wilkinson:2006}, and the layered grammar of graphics \citep{wickham:2007d}, an extension tailored specifically for \proglang{R}.
 
-Most graphical tools in \proglang{R} are input tidy, including the \pkg{base} \code{plot()} function, the \pkg{lattice} family of plots \citep{sarkar:2008} and \pkg{ggplot2} \citep{me:ggplot2}.
+Most graphical tools in \proglang{R} are input-tidy, including the \pkg{base} \code{plot()} function, the \pkg{lattice} family of plots \citep{sarkar:2008} and \pkg{ggplot2} \citep{me:ggplot2}.
 Some specialised tools exist for visualising messy datasets.
@@
 
-Modelling is the inspiration that has driven much of this work, because most modelling tools work best with tidy datasets.
+Modelling is the driving inspiration of this work because most modelling tools work best with tidy datasets.
 Every statistical language has a way of describing a model as a connection between different variables, a domain specific language that connects responses to predictors:
@@
 Significant transformations take place to produce a numeric matrix that can easily be fed to standard linear algebra routines.
-Common transformations include adding an intercept column (column of all ones), turning categorical variables into multiple binary dummy variables, and for smooth terms like splines, projecting the data values on to the appropriate basis functions.
+Common transformations include adding an intercept column (a column of ones), turning categorical variables into multiple binary dummy variables, and projecting data onto the appropriate basis of a spline function.
 
@@
 For example, in \proglang{SAS}'s \code{proc glm}, multiple variables on the response side of the equation will be interpreted as repeated measures if the {\sc repeated} keyword is present.
-For the raw billboard data, we could construct a model of the form \code{wk1-wk76 = track} instead of \code{rank = week * track} on the tidy data (provided week is labelled as a categorical variable).
+For the raw Billboard data, we could construct a model of the form \code{wk1-wk76 = track} instead of \code{rank = week * track} on the tidy data (provided week is labelled as a categorical variable).
 
@@
 If the data is stored as in Table~\subref*{tbl:paired}, then a paired t-test is just a t-test applied to the mean difference between x and y.
-If it is stored in the form of Table~\subref*{tbl:mixed}, then an equivalent model is to fit a mixed effects model, with a fixed dummy variable representing the \code{variable}, and a random intercept for each id.
-(In \proglang{R}'s lmer4 notation, this is \code{value ~ variable + (1 | id)}).
+If it is stored in the form of Table~\subref*{tbl:mixed}, then an equivalent result can be produced by fitting a mixed effects model, with a fixed dummy variable representing the \code{variable}, and a random intercept for each id.
+(In \proglang{R}'s lmer4 notation, this is expressed as \code{value ~ variable + (1 | id)}).
 Either way of modelling the data yields the same result.
@@
 
-While model inputs usually require tidy inputs, model outputs, such as predictions and estimated coefficients, aren't always tidy.
+While model inputs usually require tidy inputs, such attention to detail doesn't carry over to model outputs.
+Outputs such as predictions and estimated coefficients aren't always tidy.
 This makes it more difficult to combine results from multiple models.

### 2014-01-15 tidy-data.tex
5-7 + typos

@@
 In a given analysis, there may be multiple types of observational unit.
-For example, in a trial of new allergy medication we might have three observational types: demographic data collected from each person (\code{age}, \code{sex}, \code{race}), medical data collected from each person on each day (\code{number of sneezes}, \code{redness of eyes}), and meterological data collected on each day (\code{temperature}, \code{pollen count}).
+For example, in a trial of new allergy medication we might have three observational types: demographic data collected from each person (\code{age}, \code{sex}, \code{race}), medical data collected from each person on each day (\code{number of sneezes}, \code{redness of eyes}), and meteorological data collected on each day (\code{temperature}, \code{pollen count}).
 
@@
 The \pkg{plyr} package makes this straightforward in \proglang{R}.
-The following code generates a vector of file names in a directory (\code{data/}), matching a regular exrepssion (ends in \code{.csv}).
+The following code generates a vector of file names in a directory (\code{data/}), matching a regular expression (ends in \code{.csv}).
 Next we name each element of the vector with the name of the file.
@@
 This section discusses tidy tools, tools that take tidy datasets as input and return tidy datasets as output.
-Tidy tools are useful because the output of one tool can be used as the input to another.
-This allows you to simply and easily compose multiple tools to solve a problem.
+Tidy tools are useful because the output of one tool can be used as the input to another, making it straightforward to compose multiple tools to solve a problem.
 Tidy data also ensures that variables are stored in a consistent, explicit manner.
@@
 Data manipulation includes variable-by-variable transformation (e.g., \code{log} or \code{sqrt}), as well as aggregation, filtering and reordering.
-In my experience, these are the four fundamental verbs of data manipulation:
+In my experience, there are four extremely common operations that are performed over and over again in the course of an analysis.
+These are the four fundamental verbs of data manipulation:
 
@@
 
-  \item Filter: subsetting or removing observations based on some
+  \item Filtering, or subsetting, where observations are removed based on some
   condition.
 
-  \item Transform: adding or modifying variables. These modifications can
-  involve either a single variable (e.g., log-transformation), or multiple 
-  variables (e.g., computing density from weight and volume).
+  \item Transforming, where new variables are added or existing variables
+  modified. These modifications can involve either a single variable (e.g., 
+  log-transforming a variable), or involve multiple variables (e.g., computing
+  density from weight and volume).
 
-  \item Aggregate: collapsing multiple values into a single value (e.g., by 
-  summing or taking means).
+  \item Aggregating, where multiple values are collapsed into a single value,
+  e.g., by summing or taking means.
 
-  \item Sort: changing the order of observations.
+  \item Sorting, where the order of observations is changed.
 
@@
 
-All these operations are made easier when there is a consistent way to refer to variables. Tidy data provides this because each variable resides in its own column.
+All these operations are made easier when there is a consistent way to refer to variables. Tidy data provides this because each variable lives in its own column.
 
@@
 These are input and output-tidy.
-The \code{aggregate()} function performs group-wise aggregation.
-It is input-tidy.
-Provided that a single aggregation method is used, it is also output-tidy .
+The \code{aggregate()} function performs group-wise aggregation, is input-tidy, and is output-tidy providing a single aggregation method is used.
 The \pkg{plyr} package provides tidy \code{summarise()} and \code{arrange()} functions for aggregation and sorting.
@@
 The four verbs can be, and often are, modified by the ``by'' preposition.
-We often need to perform group-wise aggregation, transformation and subsetting, to pick the biggest ??
-in each group, to average over replicates and so on.
-Combining a verb with a ``by'' operator is a concise way to apply that operation to subsets of a data frame.
-Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group.
-They are generally input-tidy.
-Base \proglang{R} possesses a \code{by()} function, which is input-tidy, but, because it produces a list, is not output-tidy.
+We often need group-wise aggregates, transformations and subsets, to pick the biggest in each group, to average over replicates and so on.
+Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time.
+Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy.
+Base \proglang{R} possesses a \code{by()} function, which is input-tidy, but not output-tidy, because it produces a list.
 The \code{ddply()} function from the \pkg{plyr} package is a tidy alternative.
@@
 This is implemented in the \code{merge()} function in base \proglang{R}, or the \code{join()} function in \pkg{plyr}.
-Compare these operators with the difficulty of combining datasets stored in arrays.
-This task typically requires painstaking alignment before matrix operations can be used, which can can make errors very hard to detect.
+Compare this to the difficulty of combining datasets stored in arrays; these typically require painstaking alignment before matrix operations can be used, and errors can be very hard to detect.
 
@@
 Tidy visualisation tools only need to be input-tidy as their output is visual.
-Domain specific languages work particularly well for the visualisation of tidy datasets because they can describe a visualisation as a mapping between variables and aesthetic properties of the graph (e.g., position, size, shape and colour).
+Domain specific languages work particularly well for the visualisation of tidy datasets because they can describe a visualisation as a mapping between variables in the dataset and aesthetic properties of the graphic like position, size, shape and colour.
 This is the idea behind the grammar of graphics \citep{wilkinson:2006}, and the layered grammar of graphics \citep{wickham:2007d}, an extension tailored specifically for \proglang{R}.
 
-Most graphical tools in \proglang{R} are input-tidy, including the \pkg{base} \code{plot()} function, the \pkg{lattice} family of plots \citep{sarkar:2008} and \pkg{ggplot2} \citep{me:ggplot2}.
+Most graphical tools in \proglang{R} are input tidy, including the \pkg{base} \code{plot()} function, the \pkg{lattice} family of plots \citep{sarkar:2008} and \pkg{ggplot2} \citep{me:ggplot2}.
 Some specialised tools exist for visualising messy datasets.
@@
 
-Modelling is the driving inspiration of this work because most modelling tools work best with tidy datasets.
+Modelling is the inspiration that has driven much of this work, because most modelling tools work best with tidy datasets.
 Every statistical language has a way of describing a model as a connection between different variables, a domain specific language that connects responses to predictors:
@@
 Significant transformations take place to produce a numeric matrix that can easily be fed to standard linear algebra routines.
-Common transformations include adding an intercept column (a column of ones), turning categorical variables into multiple binary dummy variables, and projecting data onto the appropriate basis of a spline function.
+Common transformations include adding an intercept column (column of all ones), turning categorical variables into multiple binary dummy variables, and for smooth terms like splines, projecting the data values on to the appropriate basis functions.
 
@@
 For example, in \proglang{SAS}'s \code{proc glm}, multiple variables on the response side of the equation will be interpreted as repeated measures if the {\sc repeated} keyword is present.
-For the raw Billboard data, we could construct a model of the form \code{wk1-wk76 = track} instead of \code{rank = week * track} on the tidy data (provided week is labelled as a categorical variable).
+For the raw billboard data, we could construct a model of the form \code{wk1-wk76 = track} instead of \code{rank = week * track} on the tidy data (provided week is labelled as a categorical variable).
 
@@
 If the data is stored as in Table~\subref*{tbl:paired}, then a paired t-test is just a t-test applied to the mean difference between x and y.
-If it is stored in the form of Table~\subref*{tbl:mixed}, then an equivalent result can be produced by fitting a mixed effects model, with a fixed dummy variable representing the \code{variable}, and a random intercept for each id.
-(In \proglang{R}'s lmer4 notation, this is expressed as \code{value ~ variable + (1 | id)}).
+If it is stored in the form of Table~\subref*{tbl:mixed}, then an equivalent model is to fit a mixed effects model, with a fixed dummy variable representing the \code{variable}, and a random intercept for each id.
+(In \proglang{R}'s lmer4 notation, this is \code{value ~ variable + (1 | id)}).
 Either way of modelling the data yields the same result.
@@
 
-While model inputs usually require tidy inputs, such attention to detail doesn't carry over to model outputs.
-Outputs such as predictions and estimated coefficients aren't always tidy.
+While model inputs usually require tidy inputs, model outputs, such as predictions and estimated coefficients, aren't always tidy.
 This makes it more difficult to combine results from multiple models.
@@
 The following case study illustrates how tidy data and tidy tools make data analysis easier by easing the transitions between manipulation, visualisation and modelling.
-You will not see any code that exists solely to get the output of one function into the right format for input to another.
+You will not see any code that exists solely to get the output of one function into the right format to input to another.
 I'll show the \proglang{R} code that performs the analysis, but even if you're not familiar with \proglang{R} or the exact idioms I use, I've tried to make it easy to understand by tightly interleaving code, results and explanation.
 
-The case study uses individual-level mortality data from Mexico, with the goal of finding causes of death that have notably different time patterns within a day.
+The case study uses individual-level mortality data from Mexico.
+The goal is to find causes of death with notably different temporal patterns.
 The full dataset has information on 539,530 deaths in Mexico in 2008 and 55 variables, including the the location and time of death, the cause of death, and demographics of the deceased.
@@
 
-To achieve our goal of finding unusual time courses, we'll start by counting the number of deaths in each hour (\code{hod}) for each cause (\code{cod}) with the tidy \code{count} function.
-Then we remove missing (and hence uninformative for our purpose) values with \code{subset}.
-This gives Table~\subref*{tbl:counts:1}.
-
-First, we count the the number of deaths in each hour of the day for each cause of death, and remove missing observations:
+To achieve our goal of finding unusual temporal patterns, we do the following.
+First, we count the number of deaths in each hour (\code{hod}) for each cause (\code{cod}) with the tidy \code{count} function.
 
@@
 hod2 <- count(deaths, c("hod", "cod"))
+\end{Verbatim}
+
+Then we remove missing (and hence uninformative for our purpose) values with \code{subset}.
+
+\begin{Verbatim}
 hod2 <- subset(hod2, !is.na(hod))
@@
 
-Next we join the dataset to the \code{codes} dataset which gives informative labels for the disease.
+This gives Table~\subref*{tbl:counts:1}.
+
+Next to provide informative labels for disease, we join the dataset to the \code{codes} dataset .
 These two datasets are connected by the \code{cod} variable.
@@
 
-The total deaths for each cause vary over several orders of magnitude: there were 46,794 deaths from heart attack, and 10 deaths from avalanche.
-This means that it makes more sense to compare the proportion of deaths in each hour, rather than the total number.
+The total deaths for each cause varies over several orders of magnitude: there are 46,794 deaths from heart attack but only 10 from avalanche.
+This means that rather than the total number, it makes more sense to compare the proportion of deaths in each hour.
 We compute this by breaking the dataset down by \code{cod}, and then \code{transform()}ing to add a new \code{prop} column, the hourly frequency divided by the total number of deaths from that cause.
@@
 
-\code{ddply()} breaks down its first argument (\code{hod2}) by its second argument (the \code{cod} variable), and applies its third argument (\code{transform}) to each piece.
-The fourth argument (\code{prop = freq / sum(freq)}) is passed on to transform.
+\code{ddply()} breaks down the first argument (\code{hod2}) by its second (the \code{cod} variable), and then applies the third argument (\code{transform}) to each resulting piece.
+The fourth argument (\code{prop = freq / sum(freq)}) is then passed on to transform.
 
@@
 We then compute the overall average death rate for each hour, and merge that back into the original dataset.
-This yields Table~\subref*{tbl:counts:4} and allows us to easily compare each disease with the overall pattern by comparing \code{prop} to \code{prop\_all}.
+This yields Table~\subref*{tbl:counts:4}.
+By comparing \code{prop} to \code{prop\_all}, we can easily compare each disease with the overall incidence pattern.
 
-First, we work out the number of people dying each hour by breaking down \code{hod2} by \code{hod}, summarising with a total for over each cause of death.
+First, we work out the number of people dying each hour.  We break down \code{hod2} by \code{hod}, and compute the total for each cause of death.
 
@@
 
-Next, we work out the overall proportion of people dying in each hour:
+Next, we work out the overall proportion of people dying each hour:
 
@@
 
-Then finally, we join the overall dataset with the individual dataset to make it easier to compare the two:
+Finally, we join the overall dataset with the individual dataset to make it easier to compare the two:
 
@@
 
-Next we compute a distance between the time course of each cause of death and the overall time course.
-Here we use a simple mean squared deviation.
+Next we compute a distance between the temporal course of each cause of death and the overall temporal course.
+Here we use the simple mean squared deviation.
 We also record the sample size, the total number of deaths from that cause.
-To ensure that the diseases we work with have a decent amount of underlying data we'll only work with diseases with more than 50 deaths in total ($\sim$2/hour).
+To ensure that the diseases we consider are sufficiently representative we'll only work with diseases with more than 50 total deaths ($\sim$2/hour).
 
@@
 
-We don't know the variance characteristics of this estimator, so we explore it visually by plotting \code{n} vs.\ \code{deviation}, Figure~\ref{fig:deviation-raw}.
-The linear scale shows little, except the variability decreases with sample size, but on the log-log scale, Figure~\ref{fig:deviation-log} there is a clear pattern.
-It's particularly easy to see the pattern when we add the line of best fit from a robust linear model.
+We don't know the variance characteristics of this estimator, but we can explore it visually by plotting \code{n} vs.\ \code{deviation}, Figure~\ref{fig:deviation-raw}.
+On a linear scale, the plot shows little, except that variability decreases with sample size.
+But on the log-log scale, Figure~\ref{fig:deviation-log}, there is a clear pattern.
+This is particularly easy to see the pattern when we add the line of best fit from a robust linear model.
 
@@
 
-  \caption{(Left) Plot of n vs deviation.
-Variability of deviation dominated by the sample size: small samples have large variability.
-(Right) Log-log plot makes it easy to see pattern of variation as well as unusually high values.
-Blue line is a robust line of best fit.}
+  \caption{(a) Plot of n vs deviation.
+Variability of deviation is dominated by sample size: small samples have large variability.
+(b) Log-log plot makes it easy to see the pattern of variation as well as unusually high values.
+The blue line is a robust line of best fit.}
   \label{fig:deviation}
@@
 
-We are interested in points that are high relative to other causes with the same number of deaths, these are the diseases most different from the overall pattern.
-To find these unusual points, we fit a robust linear model and plot the residuals, as shown in Figure~\ref{fig:devi-resid}.
-The plot shows an empty region around a residual of 1.5, so somewhat arbitrarily we select those diseases with a residual greater than 1.5.
-We do that in two steps: first selecting the appropriate rows from \code{devi} (one row per disease), and then finding the matching time course information in the original summary dataset (24 rows per disease).
+We are interested in points that are particularly higher than other points in the graph.
+Controlling for the number of deaths, these points represent the diseases which depart the most from the overall pattern.
+
+To find these unusual points, we fit a robust linear model and plot the residuals,  Figure~\ref{fig:devi-resid}.
+The plot shows an empty region around a residual of 1.5.
+So somewhat arbitrarily, we'll select those diseases with a residual greater than 1.5.
+We do this in two steps: first, we select the appropriate rows from \code{devi} (one row per disease), and then we find the matching temporal course information from the original summary dataset (24 rows per disease).
 
@@
 
-Finally, we plot the time courses of the unusual causes in Figure~\ref{fig:disease}.
-We break the diseases into two plots because of the differences in variability: the top plot shows diseases with over 350 deaths and the bottom with less than 350.
-The causes of death fall into three main groups: murders, drowning, and transportation related.
-Murders are more common at night, drowning in the afternoon, and transportation related during commute times.
-The pale gray line in the background shows the time course across all diseases.
+Finally, we plot the temporal course for each unusual cause, Figure~\ref{fig:disease}.
+We split the diseases into two plots because of differences in variability.
+The top plot shows diseases with over 350 deaths and the bottom with under 350.
+The causes of death fall into three main groups: murder, drowning, and transportation related.
+Murder is more common at night, drowning in the afternoon, and transportation related deaths during commute times.
+The pale gray line in the background shows the temporal course across all diseases.
 
@@
     \includegraphics[width=0.9\textwidth]{case-study/unusual-sml}
-  \caption{Causes of death with unusual time courses.
+  \caption{Causes of death with unusual temporal courses.
 Overall hourly death rate shown in grey.
@@
 (Bottom) Causes of death with less than 350 deaths.
-Note that different scales are used on the y axis.}
+Note that the y-axes are on different scales.}
   \label{fig:disease}
@@
 Data cleaning is an important problem, but it is an uncommon subject of study in statistics.
-This paper carves out a small but important subset of data cleaning that I've called data tidying: getting the dataset in the right structure to make further manipulation, visualisation and modelling easy.
+This paper carves out a small but important subset of data cleaning that I've called data tidying: structuring datasets to facilitate manipulation, visualisation and modelling.
 There is still much work to be done.
-Incremental improvements will happen as my understanding of tidy data and tidy tools improves, reducing the friction of getting data into a tidy form, and then manipulating, modelling and visualising it once it's tidy.
+Incremental improvements will happen as our understanding of tidy data and tidy tools improves, and as we improve our ability to reduce the friction of getting data into a tidy form.
 
 Bigger improvements may be possible by exploring alternative formulations of tidiness.
-There is a chicken-and-egg problem with tidy data: if tidy data is only as useful as the tools that work with it, then tidy tools are inextricably linked to tidy data.
-This makes it easy to get stuck in a local maxima where independently changing data structures or data tools does not result in an improved workflow.
-Breaking out of this local maxima is hard, and requires long-term concerted effort with many false starts.
-While I hope that this framework for tidy data is not a false start, equally, I don't see it as the final solution and I hope others will build on this framework to develop even better ways of storing data and better tools for working with it.
+There is a chicken-and-egg problem with tidy data: if tidy data is only as useful as the tools that work with it, then tidy tools will be inextricably linked to tidy data.
+This makes it easy to get stuck in a local maxima where independently changing data structures or data tools will not improve workflow.
+Breaking out of this local maxima is hard.
+It requires long-term concerted effort with the prospect of many false starts.
+While I hope that the tidy data framework is not one of those false starts, I also don't see it as the final solution.
+I hope others will build on this framework to develop even better data storage strategies and better tools.
 
-I have found few principles to guide the design of tidy data, which must acknowledge both statistical and cognitive factors.
-To date, my work has been driven by my experience doing data analysis, connections to relational database design, and introspection on the tools of data analysis.
-The human factors, user-centred design, and human-computer interaction communities may be able to add to this conversation, but the design of data and tools to work with it has not been an active research topic.
-In the future, I hope to apply methodologies from these fields (user-testing, ethnography, talk-aloud protocols) to my work to build a better understanding of the cognitive side of data analysis, and a better knowledge of how we can design tools to support the process.
+Surprisingly, I have found few principles to guide the design of tidy data, which acknowledge both statistical and cognitive factors.
+To date, my work has been driven by my experience doing data analysis, my knowledge of relational database design, and my own rumination on the tools of data analysis.
+The human factors, user-centered design, and human-computer interaction communities may be able to add to this conversation, but the design of data and tools to work with it has not been an active research topic in those fields.
+In the future, I hope to use methodologies from these fields (user-testing, ethnography, talk-aloud protocols) to improve our understanding of the cognitive side of data analysis, and to further improve our ability to design appropriate tools.
 
@@
 For example, it would be possible to construct a set of tools for dealing with values stored in multidimensional arrays.
-This is a common data storage format for large biomedical datasets such as microarray or fMRI data.
-It is also necessary for many multivariate methods whose underlying theory is based on matrix manipulation.
-This array-tidy format would be compact and efficient, because there are many efficient tools for working with high-dimensional arrays, even when sparse, and the format would connect more closely with the mathematical basis of statistics.
-This approach is taken by the Pandas python data analysis library \citep{mckinney:2010}.
-Even more interestingly, we could consider tidy tools that can ignore the underlying data representation, automatically choosing between array-tidy and dataframe-tidy formats to optimise memory usage and performance.
+This is a common storage format for large biomedical datasets generated by microarrays or fMRI's.
+It's also necessary for many multivariate methods based on matrix manipulation.
+Fortunately, because there are many efficient tools for working with high-dimensional arrays, even sparse ones, such an array-tidy format is not only likely to be quite compact and efficient, it should also be able to easily connect with the mathematical basis of statistics.
+This, in fact, is the approach taken by the Pandas python data analysis library \citep{mckinney:2010}.
+Even more interestingly, we could consider tidy tools that can ignore the underlying data representation and automatically choose between array-tidy and dataframe-tidy formats to optimise memory usage and performance.
 
@@
 
-This work wouldn't be possible without the many conversations I've had about data and how to deal with it statistically.
+This work wouldn't be possible without the many conversations I've had about data and how to deal with them statistically.
 I'd particularly like to thank Phil Dixon, Di Cook, and Heike Hofmann, who have put up with numerous questions over the years.
 I'd also like to thank the users of the \pkg{reshape} package who have provided many challenging problems, and my students who continue to challenge me to explain what I know in a way that they can understand.
-I'd also like to thank Bob Muenchen, Burt Gunter, Nick Horton and Garrett Grolemund who gave detailed comments on earlier drafts, and to particularly thank Ross Gayler who provided the nice example of the challenges of defining what a variable is and Ben Bolker who showed me the natural equivalent between a paired t-test and a mixed effects model.
+I'd also like to thank Bob Muenchen, Burt Gunter, Nick Horton and Garrett Grolemund who gave detailed comments on earlier drafts, and to particularly thank Ross Gayler who provided the nice example of the challenges of defining a variable and Ben Bolker who showed me the natural equivalence between a paired t-test and a mixed effects model.
 

### 2014-01-16 tidy-data.tex
biggest

@@
 The four verbs can be, and often are, modified by the ``by'' preposition.
-We often need to perform group-wise aggregation, transformation and subsetting, to pick the biggest ??
-in each group, to average over replicates and so on.
+We often need to perform group-wise aggregation, transformation and subsetting, to find the largest value in each group, to average over replicates and so on.
 Combining a verb with a ``by'' operator is a concise way to apply that operation to subsets of a data frame.

### 2014-01-17 tidy-data.tex
Another attempt at describing semantics

@@
 
-A table is a collection of data \textbf{values}, typically either numbers (if quantitative) or strings (if qualitative).
-A \textbf{variable} is a set of measurements of a single quantifiable property for multiple observational units.
-An \textbf{observation} is a set of measurements of multiple quantifiable properties for a single observational unit.
-An observational unit is the smallest constituent atom of the experimental unit.
-For example, in a longitudinal medical study, the experimental unit might be the patient, but the observational unit is the patient at different points in time.
-In an agricultural experiment, the experimental unit might be the cages assigned different diets, but the observational units are the individual pigs in those cages.
-In short, while an observational unit represents a conditional property, like blood pressure at time 1 or the weight of pig 2, an experimental unit represents an unconditional property, like blood pressure and weight.
+A table is a collection of data \textbf{values}, usually either numbers (if quantitative) or strings (if qualitative).
+Values are organised in two ways.
+Every value belongs to a \textbf{variable} and an \textbf{observation}.
+A variable collects all values that measure the same underlying attribute (like height, temperature, duration) across units.
+An observation collects all values measured on the same unit (like a person, or a day, or a race) across attributes.
 
-The dataset in Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} contains three variables that comprise 10 values.
-The data contains a \code{person} variable with three values (John, Mary, and Jane), a \code{treatment} variable with two values (a and b), and a \code{result} variable with has five or six values depending on how you think of the missing value (-, 16, 3, 2, 11, 1).
-The layout of Tables~\ref{tbl:preg-raw-1} and \ref{tbl:preg-raw-2} makes the number of values clear, but obscures their meaning.
+Table~\ref{tbl:preg-tidy} reorganises the data of Table~\ref{tbl:preg-raw-1} to make the values, variables and obserations more clear.
+We can see that the dataset contains 18 values representing three variables and six observations.
+The variables are:
 
-The meaning of the dataset relies on how these values are related to each other.
-The \code{result} variable was collected over the course of six observations: each \code{person} was observed twice, once for each \code{treatment}.
-These observations give meaning to the specific values of result.
-For example, the value 16 is meaningful in the context of the observation: \emph{Jane} scored \emph{16} when she took treatment \code{a}.
-Table~\ref{tbl:preg-tidy} organizes the dataset around the six observations, making the meaning of the individual values more clear.
+\begin{enumeriate}
+
+\item \code{person}, with three possible values (John, Mary, and Jane),
+
+\item \code{treatment}, with two possible values (a and b), and
+
+\item \code{result}, with five or six values depending on how you think of the missing value (-, 16, 3, 2, 11, 1).
+
+\end{enumerate}
+
+The experimental design tells us more about the structure of the observations. In this experiment, every combination of of \code{person} and \code{treatment} was measured, a completely crossed design.
 
@@
 For example, if the columns in the Table~\ref{tbl:preg-raw-1} were \code{height} and \code{weight} we would have been happy to call them variables.
-If the columns were
-\code{height} and \code{width}, it would be less clear cut, as we might think of height and width as values of a \code{dimension} variable.
+If the columns were \code{height} and \code{width}, it would be less clear cut, as we might think of height and width as values of a \code{dimension} variable.
 If the columns were \code{home phone} and \code{work phone}, we could treat these as two variables, but in a fraud detection environment we might want variables \code{phone number} and \code{number type} because the use of one phone number for multiple people might suggest fraud.

### 2014-01-17 tidy-data.tex
Manual merge. Closes #7

@@
 
-\begin{enumeriate}
+\begin{enumerate}
 
@@
 The four verbs can be, and often are, modified by the ``by'' preposition.
-We often need to perform group-wise aggregation, transformation and subsetting, to find the largest value in each group, to average over replicates and so on.
-Combining a verb with a ``by'' operator is a concise way to apply that operation to subsets of a data frame.
-Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group.
-They are generally input-tidy.
-Base \proglang{R} possesses a \code{by()} function, which is input-tidy, but, because it produces a list, is not output-tidy.
+We often need group-wise aggregates, transformations and subsets, to pick the biggest in each group, to average over replicates and so on.
+Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time.
+Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy.
+Base \proglang{R} possesses a \code{by()} function, which is input-tidy, but not output-tidy, because it produces a list.
 The \code{ddply()} function from the \pkg{plyr} package is a tidy alternative.
@@
 
-To achieve our goal of finding unusual time courses, we'll start by counting the number of deaths in each hour (\code{hod}) for each cause (\code{cod}) with the tidy \code{count} function.
-Then we remove missing (and hence uninformative for our purpose) values with \code{subset}.
-This gives Table~\subref*{tbl:counts:1}.
+To achieve our goal of finding unusual temporal patterns, we do the following.
+First, we count the number of deaths in each hour (\code{hod}) for each cause (\code{cod}) with the tidy \code{count} function.
 
@@
 
-Next we join the dataset to the \code{codes} dataset which gives informative labels for the disease.
+This gives Table~\subref*{tbl:counts:1}.
+
+To provide informative labels for disease, we next join the dataset to the \code{codes} dataset .
 These two datasets are connected by the \code{cod} variable.
-This adds a new variable, \code{disease}, shown in Table~\subref*{tbl:counts:2}).
+This adds a new variable, \code{disease}, shown in Table~\subref*{tbl
+:counts:2})
 
@@
 
-We are interested in points that are high relative to other causes with the same number of deaths, these are the diseases most different from the overall pattern.
-To find these unusual points, we fit a robust linear model and plot the residuals, as shown in Figure~\ref{fig:devi-resid}.
-The plot shows an empty region around a residual of 1.5, so somewhat arbitrarily we select those diseases with a residual greater than 1.5.
-We do that in two steps: first selecting the appropriate rows from \code{devi} (one row per disease), and then finding the matching time course information in the original summary dataset (24 rows per disease).
+We are interested in points that have high $y$-values, relative to their $x$-neighbours.
+Controlling for the number of deaths, these points represent the diseases which depart the most from the overall pattern.
+
+To find these unusual points, we fit a robust linear model and plot the residuals,  Figure~\ref{fig:devi-resid}.
+The plot shows an empty region around a residual of 1.5.
+So somewhat arbitrarily, we'll select those diseases with a residual greater than 1.5.
+We do this in two steps: first, we select the appropriate rows from \code{devi} (one row per disease), and then we find the matching temporal course information from the original summary dataset (24 rows per disease).
 
@@
     \includegraphics[width=0.9\textwidth]{case-study/unusual-sml}
-  \caption{Causes of death with unusual time courses.
+  \caption{Causes of death with unusual temporal courses.
 Overall hourly death rate shown in grey.
 (Top) Causes of death with more than 350 deaths over a year.
-(Bottom) Causes of death with less than 350 deaths.
-Note that different scales are used on the y axis.}
+(Bottom) Causes of death with fewer than 350 deaths.
+Note that the y-axes are on different scales.}
   \label{fig:disease}

### 2014-01-17 tidy-data.tex
More tweaking. Add overall time course plot

@@
 It is often said that 80\% of data analysis is spent on the process of cleaning and preparing the data \citep{dasu:2003}.
-However, since such data cleaning is often repeated over the course of analysis as new problems come to light, it is much more than just a first step.
+Data preparation is not just a first step, but must be repeated many over the course of analysis as new problems come to light or new data is collected.
 Despite the amount of time it takes, there has been surprisingly little research on how to clean data well.
 Part of the challenge is the breadth of activities it encompasses: from outlier checking, to date parsing, to missing value imputation.
-To get a handle on the problem, this paper focusses on a small, but important, aspect of data cleaning that I call data \textbf{tidying}: the structuring of datasets to facilitate analysis.
+To get a handle on the problem, this paper focusses on a small, but important, aspect of data cleaning that I call data \textbf{tidying}: structuring datasets to facilitate analysis.
 
-The principles of tidy data provide a standard way of organising data values within a dataset.
-A standard makes the initial data cleaning easier because you don't need to start from scratch and reinvent the wheel each time.
-The tidy data standard has been designed to facilitate the initial exploration and analysis of the data, and to simplify the development of data analysis tools which can work together seamlessly.
+The principles of tidy data provide a standard way to organise data values within a dataset.
+A standard makes initial data cleaning easier because you don't need to start from scratch and reinvent the wheel every time.
+The tidy data standard has been designed to facilitate initial exploration and analysis of the data, and to simplify the development of data analysis tools that work well together.
 Current tools often require translation.
-You need to spend time munging the output from one tool so you can input it into another.
-Tidy datasets and tidy tools work hand in hand to make data analysis easier, allowing you to focus on the interesting domain problem, and not on the uninteresting logistics of your data.
+You have to spend time munging the output from one tool so you can input it into another.
+Tidy datasets and tidy tools work hand in hand to make data analysis easier, allowing you to focus on the interesting domain problem, not on the uninteresting logistics of data.
 
@@
 The development of tidy data has been driven by my experience working with real-world datasets.
-With few, if any, constraints on their organisation, such datasets are constructed in often bizarre ways.
+With few, if any, constraints on their organisation, such datasets are often constructed in bizarre ways.
 I have spent countless hours struggling to get such datasets organised in a way that makes data analysis possible, let alone easy.
 I have also struggled to impart these skills to my students so they could tackle real-world datasets on their own.
-In the course of these struggles I developed the \pkg{reshape} and \pkg{reshape2} \citep{wickham:2007b} \proglang{R} packages.
-While I could intuitively use the tools, I lacked the framework to make my intuition explicit.
+In the course of these struggles I developed the \pkg{reshape} and \pkg{reshape2} \citep{wickham:2007b} packages.
+While I could intuitively use the tools and teach them through examples, I lacked the framework to make my intuition explicit.
 This paper provides that framework.
@@
 Tidy datasets provide a standardized way to link the structure of a dataset (its physical layout) with its semantics (its meaning).
-In this section, I'll provide some standard vocabulary for describing the structure of a dataset, its semantics, and then use those definitions to define tidy data.
+In this section, I'll provide some standard vocabulary for describing the structure and semantics of a dataset, and then use those definitions to define tidy data.
 
@@
 Most statistical datasets are rectangular tables made up of \textbf{rows} and \textbf{columns}.
-While columns are almost always labelled, rows are sometimes not.
+The columns are almost always labelled and the rows are sometimes labelled.
 Table~\ref{tbl:preg-raw-1} provides some data about an imaginary experiment in a format commonly seen in the wild.
@@
 Our vocabulary of rows and columns is simply not rich enough to describe why the two tables represent the same data.
-In addition to appearance, what we need is a way to describe the underlying semantics, or meaning, of the values displayed in the dataset.
+In addition to appearance, we need a way to describe the underlying semantics, or meaning, of the values displayed in table.
 
@@
 
-A table is a collection of data \textbf{values}, usually either numbers (if quantitative) or strings (if qualitative).
+A dataset is a collection of \textbf{values}, usually either numbers (if quantitative) or strings (if qualitative).
 Values are organised in two ways.
 Every value belongs to a \textbf{variable} and an \textbf{observation}.
-A variable collects all values that measure the same underlying attribute (like height, temperature, duration) across units.
-An observation collects all values measured on the same unit (like a person, or a day, or a race) across attributes.
+A variable contains all values that measure the same underlying attribute (like height, temperature, duration) across units.
+An observation contains all values measured on the same unit (like a person, or a day, or a race) across attributes.
 
-Table~\ref{tbl:preg-tidy} reorganises the data of Table~\ref{tbl:preg-raw-1} to make the values, variables and obserations more clear.
-We can see that the dataset contains 18 values representing three variables and six observations.
+Table~\ref{tbl:preg-tidy} reorganises Table~\ref{tbl:preg-raw-1} to make the values, variables and obserations more clear.
+The dataset contains 18 values representing three variables and six observations.
 The variables are:
@@
 
-The experimental design tells us more about the structure of the observations. In this experiment, every combination of of \code{person} and \code{treatment} was measured, a completely crossed design.
+The experimental design tells us more about the structure of the observations.
+In this experiment, every combination of of \code{person} and \code{treatment} was measured, a completely crossed design.
+The experimental design also determines whether or not missing values can be safely dropped.
+In this experiment, the missing value represents an observation that should have been made, but wasn't, so it's important to keep it.
+Structural missing values, which represent measurements that can not be made (e.g.
+the count of pregnant males) can be safely removed.
 
@@
 
-For a given dataset, it's usually easy to figure out what are observations and what are variables, but precisely defining variables and observations in general is surprisingly difficult.
+For a given dataset, it's usually easy to figure out what are observations and what are variables, but it is surprisingly difficult to precisely define variables and observations in general.
 For example, if the columns in the Table~\ref{tbl:preg-raw-1} were \code{height} and \code{weight} we would have been happy to call them variables.
@@
 If the columns were \code{home phone} and \code{work phone}, we could treat these as two variables, but in a fraud detection environment we might want variables \code{phone number} and \code{number type} because the use of one phone number for multiple people might suggest fraud.
-A general rule of thumb is that it is easier to describe functional relationships between variables (e.g., \code{z} is a linear combination of \code{x} and \code{y}, \code{density} is the ratio of \code{weight} to \code{volume}) than between rows, and it is easier to make comparisons between groups of observations (e.g., average of group a vs.\ average of group b) than between columns.
+A general rule of thumb is that it is easier to describe functional relationships between variables (e.g., \code{z} is a linear combination of \code{x} and \code{y}, \code{density} is the ratio of \code{weight} to \code{volume}) than between rows, and it is easier to make comparisons between groups of observations (e.g., average of group a vs.\ average of group b) than between groups of columns.
 
-In a given analysis, there may be multiple types of observational unit.
+In a given analysis, there may be multiple levels of observation.
 For example, in a trial of new allergy medication we might have three observational types: demographic data collected from each person (\code{age}, \code{sex}, \code{race}), medical data collected from each person on each day (\code{number of sneezes}, \code{redness of eyes}), and meterological data collected on each day (\code{temperature}, \code{pollen count}).
@@
 Each row represents an observation, the \code{result} of one \code{treatment} on one \code{person}, and each column is a variable.
-We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural, impossible because of the design of the experiment.
 
@@
 This slows analysis and invites errors.
-If you consider how many data analysis operations involve all of the values in a variable (every aggregating function), you can see how important it is to extract these values in a standard, simple way.
+If you consider how many data analysis operations involve all of the values in a variable (every aggregating function), you can see how important it is to extract these values in a simple, standard way.
 Tidy data is particularly well suited for vectorised programming languages like \proglang{R}, because the layout ensures that values of different variables from the same observation are always paired.
@@
 The case study uses individual-level mortality data from Mexico.
-The goal is to find causes of death with notably different temporal patterns.
+The goal is to find causes of death with unusual temporal patterns within a day.
+Figure~\ref{fig:overall} shows the temporal pattern, the number of deaths per hour, for all causes of death.
+My goal is to find the diseases that differ most from this pattern.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.65\linewidth]{case-study/overall}
+  \caption{Temporal pattern of all causes of death.}
+  \label{fig:overall}
+\end{figure}
+
 The full dataset has information on 539,530 deaths in Mexico in 2008 and 55 variables, including the the location and time of death, the cause of death, and demographics of the deceased.
@@
 
-First, we count the the number of deaths in each hour of the day for each cause of death, and remove missing observations:
-
 \begin{Verbatim}
@@
 This gives Table~\subref*{tbl:counts:1}.
-
-To provide informative labels for disease, we next join the dataset to the \code{codes} dataset .
-These two datasets are connected by the \code{cod} variable.
-This adds a new variable, \code{disease}, shown in Table~\subref*{tbl
-:counts:2})
+To provide informative labels for disease, we next join the dataset to the \code{codes} dataset, connected by the \code{cod} variable.
+This adds a new variable, \code{disease}, shown in Table~\subref*{tbl:counts:2}.
 
@@
 
-We don't know the variance characteristics of this estimator, but we can explore it visually by plotting \code{n} vs.\ \code{deviation}, Figure~\ref{fig:deviation-raw}.
+We don't know the variance characteristics of this estimator, but we can explore it visually by plotting \code{n} vs.\ \code{deviation}, Figure~\subref*{fig:deviation-raw}.
 On a linear scale, the plot shows little, except that variability decreases with sample size.
-But on the log-log scale, Figure~\ref{fig:deviation-log}, there is a clear pattern.
+But on the log-log scale, Figure~\subref*{fig:deviation-log}, there is a clear pattern.
 This is particularly easy to see the pattern when we add the line of best fit from a robust linear model.

### 2014-01-17 case-study/case-study.r

@@ -34,6 +34,13 @@ codes$disease <- sapply(codes$disease, function(x)
 names(codes)[1] <- "cod"
 codes <- codes[!duplicated(codes$cod), ]
 
+# Display overall hourly deaths
+hod_all <- subset(count(deaths, "hod"), !is.na(hod))
+qplot(hod, freq, data = hod_all, geom = "line") + 
+  scale_y_continuous("Number of deaths", labels = function(x) format(x, big.mark = ",")) + 
+  xlab("Hour of day")
+ggsave("overall.pdf", width = 10, height = 6)
+
 # Count deaths per hour, per disease
 hod2 <- count(deaths, c("cod", "hod"))
 hod2 <- subset(hod2, !is.na(hod))

### 2014-01-17 tidy-data.tex
structural missingness

--- 2014-01-17 tidy-data.tex
@@
 Each row represents an observation, the \code{result} of one \code{treatment} on one \code{person}, and each column is a variable.
-We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural, impossible because of the design of the experiment.
+We might be tempted to drop the row containing the missing value.
+However, since missing values should only be omitted if they are structural (e.g., the number of pregnant men), doing so in this case would not be justified.
+It's an observation that should have been made.
 

### 2014-01-17 tidy-data.tex
Final tweaks

--- 2014-01-17 tidy-data.tex
@@
 In this experiment, the missing value represents an observation that should have been made, but wasn't, so it's important to keep it.
-Structural missing values, which represent measurements that can not be made (e.g.
+Structural missing values, which represent measurements that can't be made (e.g.
 the count of pregnant males) can be safely removed.
@@
 
-Because it provides a standard way of structuring a dataset, tidy data makes it easy for an analyst or a computer to extract needed variables.
-Contrast Table~\ref{tbl:preg-tidy} to Table~\ref{tbl:preg-raw-1}: in Table~\ref{tbl:preg-raw-1} you need to use different strategies to extract different variables.
+Tidy data makes it easy for an analyst or a computer to extract needed variables because it provides a standard way of structuring a dataset.
+Compare Table~\ref{tbl:preg-tidy} to Table~\ref{tbl:preg-raw-1}: in Table~\ref{tbl:preg-raw-1} you need to use different strategies to extract different variables.
 This slows analysis and invites errors.
-If you consider how many data analysis operations involve all of the values in a variable (every aggregating function), you can see how important it is to extract these values in a simple, standard way.
+If you consider how many data analysis operations involve all of the values in a variable (every aggregation function), you can see how important it is to extract these values in a simple, standard way.
 Tidy data is particularly well suited for vectorised programming languages like \proglang{R}, because the layout ensures that values of different variables from the same observation are always paired.

### 2014-01-22 data/melt.r
Improve melting example

@@ -1,7 +1,7 @@
 library(reshape2)
 source("xtable.r")
 
-df <- data.frame(row = letters[1:3], a = 1:3, b = 4:6, c = 7:9)
+df <- data.frame(row = LETTERS[1:3], a = 1:3, b = 4:6, c = 7:9)
 xtable(df, "melt-raw.tex")
 
 dfm <- melt(df, id = "row")

### 2014-01-22 tidy-data.tex
Tweaks from Bob

@@
 
-While order of variables and observations does not affect analysis, a good ordering makes it easier to scan the raw values.
+While the order of variables and observations does not affect analysis, a good ordering makes it easier to scan the raw values.
 One way of organising variables is by their role in the analysis: are values fixed by the design of the data collection, or are they measured during the course of the experiment?
@@
   \input{data/pew-raw.tex}
-  \caption{The first ten rows of data on income and religion from the Pew Forum. Two columns, \code{\$100-150k} and \code{\$150k}, have been omitted} 
+  \caption{The first ten rows of data on income and religion from the Pew Forum. Two columns, \code{\$100-150k} and \code{>150k}, have been omitted} 
   \label{tbl:pew-raw}
@@
 Melting is parameterised by a list of columns that are already variables, or \textbf{colvar}s for short.
-The other columns are converted into two variables: a new variable called \code{column} that contains repeated column headings and a new variable called \code{value} that contains a vector of data values concatenated from the previously separate columns.
+The other columns are converted into two variables: a new variable called \code{column} that contains repeated column headings and a new variable called \code{value} that contains the concatenated data values from the previously separate columns.
 This is illustrated in Table~\ref{tbl:melt} with a toy dataset.
@@
 
-Table~\subref*{tbl:tb-molten} shows the results of melting the TB dataset, and Table~\subref*{tbl:tb-tidy} shows the results of splitting the column \code{column} into two real variables: \code{age} and \code{sex}.
+Table~\subref*{tbl:tb-molten} shows the results of melting the TB dataset, and Table~\subref*{tbl:tb-tidy} shows the results of splitting the single column \code{column} into two real variables: \code{age} and \code{sex}.
 
@@
 Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time.
-Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy.
+Most \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy.
 Base \proglang{R} possesses a \code{by()} function, which is input-tidy, but not output-tidy, because it produces a list.
@@
 Modelling is the driving inspiration of this work because most modelling tools work best with tidy datasets.
-Every statistical language has a way of describing a model as a connection between different variables, a domain specific language that connects responses to predictors:
+Every statistical language has a way of describing a model as a connection among different variables, a domain specific language that connects responses to predictors:
 
@@
 
-  \item \proglang{SAS} (\code{PROC GLM}): \code{y = a + b + c * d}
+  \item \proglang{SAS} (\code{PROC GLM}): \code{y = a + b + c + d + c * d}
 
@@
 
-  \item \proglang{Stata} (\code{regress}): \code{y a b c \# d}
+  \item \proglang{Stata} (\code{regress}): \code{y a b c\#d}
 
@@
 For example, in \proglang{SAS}'s \code{proc glm}, multiple variables on the response side of the equation will be interpreted as repeated measures if the {\sc repeated} keyword is present.
-For the raw Billboard data, we could construct a model of the form \code{wk1-wk76 = track} instead of \code{rank = week * track} on the tidy data (provided week is labelled as a categorical variable).
+For the raw Billboard data, we could construct a model of the form \code{wk1-wk76 = track} instead of \code{rank = week * track} on the tidy data.
 
@@
 The full dataset has information on 539,530 deaths in Mexico in 2008 and 55 variables, including the the location and time of death, the cause of death, and demographics of the deceased.
-Table~\ref{fig:raw} shows a small sample of the dataset, focussing on variables related to time of death (\code{year}, \code{month}, \code{day} and \code{hour}), and cause (\code{cod}).
+Table~\ref{fig:raw} shows a small sample of the dataset, focussing on variables related to time of death (\code{year}, \code{month}, \code{day} and \code{hour}), and cause of death (\code{cod}).
 
@@
 \code{ddply()} breaks down the first argument (\code{hod2}) by its second (the \code{cod} variable), and then applies the third argument (\code{transform}) to each resulting piece.
-The fourth argument (\code{prop = freq / sum(freq)}) is then passed on to transform.
+The fourth argument (\code{prop = freq / sum(freq)}) is then passed on to \code{transform()}.
 
@@
 Next we compute a distance between the temporal course of each cause of death and the overall temporal course.
-Here we use the simple mean squared deviation.
+Here we use a simple mean squared deviation between the proportion of people who died from a particular cause and all causes, for each hour.
 We also record the sample size, the total number of deaths from that cause.

### 2014-01-22 tidy-data.tex
More minor tweaks

@@
 
-Next we compute a distance between the temporal course of each cause of death and the overall temporal course.
-Here we use a simple mean squared deviation between the proportion of people who died from a particular cause and all causes, for each hour.
+Next we compute a distance between the temporal pattern of each cause of death and the overall temporal pattern.
+There are many ways to measure this distance, but I found a simple mean squared deviation to be revealing.
 We also record the sample size, the total number of deaths from that cause.

### 2014-05-09 tidy-data.tex
First round of fixes

@@
 
-\item \code{person}, with three possible values (John, Mary, and Jane),
+\item \code{person}, with three possible values (John, Mary, and Jane).
 
-\item \code{treatment}, with two possible values (a and b), and
+\item \code{treatment}, with two possible values (a and b).
 
@@
 In this experiment, the missing value represents an observation that should have been made, but wasn't, so it's important to keep it.
-Structural missing values, which represent measurements that can't be made (e.g.
-the count of pregnant males) can be safely removed.
+Structural missing values, which represent measurements that can't be made (e.g., the count of pregnant males) can be safely removed.
 
@@
 \begin{itemize}
-  \item column headers are values, not variable names
-  \item multiple variables are stored in one column
-  \item variables are stored in both rows and columns
-  \item multiple types of observational units are stored in the same table
-  \item a single observational unit is stored in multiple tables
+  \item Column headers are values, not variable names.
+  \item Multiple variables are stored in one column.
+  \item Variables are stored in both rows and columns.
+  \item Multiple types of observational units are stored in the same table.
+  \item A single observational unit is stored in multiple tables.
 \end{itemize}
@@
   \input{data/pew-raw.tex}
-  \caption{The first ten rows of data on income and religion from the Pew Forum. Two columns, \code{\$100-150k} and \code{>150k}, have been omitted} 
+  \caption{The first ten rows of data on income and religion from the Pew Forum. Three columns, \code{\$75-100k}, \code{\$100-150k} and \code{>150k}, have been omitted} 
   \label{tbl:pew-raw}
@@
   \input{data/weather-raw.tex}
-  \caption{Original weather dataset.  There is a column for each possible day in the month.  Columns \code{d11} to \code{d31} have been omitted to conserve space.}
+  \caption{Original weather dataset.  There is a column for each possible day in the month.  Columns \code{d9} to \code{d31} have been omitted to conserve space.}
   \label{tbl:weather-raw}
@@
 
-  \item \proglang{R} (\code{lm()}): \code{y ~ a + b + c * d}
+  \item \proglang{R} (\code{lm()}): \code{y ~ a + b + c * d}.
 
-  \item \proglang{SAS} (\code{PROC GLM}): \code{y = a + b + c + d + c * d}
+  \item \proglang{SAS} (\code{PROC GLM}): \code{y = a + b + c + d + c * d}.
 
-  \item \proglang{SPSS} (\code{glm}): \code{y BY a b c d / DESIGN a b c d c * d}
+  \item \proglang{SPSS} (\code{glm}): \code{y BY a b c d / DESIGN a b c d c * d}.
 
-  \item \proglang{Stata} (\code{regress}): \code{y a b c\#d}
+  \item \proglang{Stata} (\code{regress}): \code{y a b c\#d}.
 

### 2014-05-09 data/billboard.r

@@ -14,6 +14,10 @@ raw$track <- str_replace(raw$track, " \\(.*?\\)", "")
 names(raw)[-(1:5)] <- str_c("wk", 1:76)
 raw <- arrange(raw, year, artist, track)
 
+long_name <- nchar(raw$track) > 20
+raw$track[long_name] <- paste0(substr(raw$track[long_name], 0, 20), "...")
+
+
 xtable(raw[c(1:3, 6:10), 1:8], "billboard-raw.tex")
 
 clean <- melt(raw, id = 1:5, na.rm = T)
@@ -37,8 +41,6 @@ song <- unrowname(unique(clean[c("artist", "track", "time")]))
 song$id <- 1:nrow(song)
 
 narrow <- song[1:15, c("id","artist", "track", "time")]
-long_name <- nchar(narrow$track) > 20
-narrow$track[long_name] <- paste0(substr(narrow$track[long_name], 0, 20), "...")
 xtable(narrow, "billboard-song.tex")
 
 rank <- join(clean, song, match = "first")

### 2014-05-09 data/pew.r

@@ -43,4 +43,4 @@ xtable(counts[1:10, ], file = "pew-clean.tex")
 # Convert into the form in which I originally saw it -------------------------
 
 raw <- dcast(counts, religion ~ income)
-xtable(raw[1:10, 1:8], file = "pew-raw.tex")
+xtable(raw[1:10, 1:7], file = "pew-raw.tex")

### 2014-05-09 data/weather.r

@@ -31,7 +31,7 @@ raw[-(1:4)] <- raw[-(1:4)] / 10
 rownames(raw) <- NULL
 raw$element <- tolower(raw$element)
 
-xtable(raw[1:10, 1:14], file = "weather-raw.tex", digits = 1)
+xtable(raw[1:10, 1:12], file = "weather-raw.tex", digits = 1)
 
 # Melt and tidy
 
